---
title: "DifferentialExpressionAnalyis"
output: html_document
date: "2025-05-26"
---


```{r, include=FALSE}

rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)

# load ggplot
pacman::p_load(ggplot2)

# set standard plot theme
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 14,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)

# choose project
studyNR = "EUT046"

```




# Set up analysis {.tabset}

## Load packages

### CRAN
```{r}
library(tidyverse)
library(data.table)
library(ggfortify)
library(here)
```


### BioManager
```{r}
# if (!require("BiocManager", quietly = TRUE)){
#     install.packages("BiocManager")
# }
# 
# if (!require(DESeq2, quietly = TRUE)){
#   BiocManager::install("DESeq2")
# }

library(DESeq2)


```



## Load functions
Load manual functions we need in the script.
```{r}
source(file.path(here(), "functions_new.R"), local = knitr::knit_global())
```





## Set color scheme
```{r}

# color list for plotting
color25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown", "forestgreen"
)   


```


## Set directories
```{r}
# data directory 
if(!dir.exists(paste0("data/"))){
  dir.create(path = file.path(here(), "data"))
}
dataDir <- file.path(here(), "data", studyNR)

# input directory
if(!dir.exists(paste0("input/", studyNR, "/"))){
  dir.create(path = file.path(here(), "input", studyNR))
}
inputDir <- file.path(here(), "input", studyNR)

# output directory
if(!dir.exists(paste0("output/", studyNR, "updatedVersion"))){
  dir.create(path = file.path(here(), "output", studyNR, "updatedVersion"))
}
outputDir <- file.path(here(), "output", studyNR, "updatedVersion")

# plot directory
if(!dir.exists(paste0("figures/", studyNR, "updatedVersion/"))){
  dir.create(path = file.path(here(), "figures", studyNR, "updatedVersion"))
}
plotDir <- file.path(here(), "figures", studyNR, "updatedVersion")


```

## Load data
```{r}
load(file.path(outputDir, paste0(studyNR, "_dataAfterSampleQC_3IQR_repfilter.RData")))
```


# Differential gene expression {.tabset}

## Generate contrast table
```{r}

contrast_control <- metadata_libSize_repFilter %>%
    filter(COMPOUND_ABBR %in% c("MEDIUM", "DMSO", "STARVMEDIUM", "DMEM", "NaCl")) %>%
    distinct(MEAN_ID, TIME, COMPOUND_ABBR, TREATMENT)


contrast_treatment <- metadata_libSize_repFilter %>%
    filter(!COMPOUND_ABBR %in% c("MEDIUM", "DMSO", "STARVMEDIUM", "DMEM", "NaCl")) %>%
    distinct(MEAN_ID, TIME, COMPOUND_ABBR, REPLICATE, TREATMENT)



contrast_table <- left_join(x = contrast_treatment %>% 
                              select(MEAN_ID_EXPERIMENT = MEAN_ID, TIME, EXPERIMENT = COMPOUND_ABBR, REPLICATE), 
                            y = contrast_control %>%
                              select(MEAN_ID_CONTROL = MEAN_ID, TIME, CONTROL=COMPOUND_ABBR), 
                            by = c("TIME")) %>% 
  distinct()


contrast_table %>%
    select(MEAN_ID_EXPERIMENT, MEAN_ID_CONTROL) %>%
    as_tibble() %>%
    print(n = 1000) 


```



# Run DESEQ


## Run DESeq2 function
#' Run DESeq2 analysis for specified contrasts
#'
#' @param contrast A data frame/tibble with columns 'caseCondition', 'controlCondition', and 'main_diagnosis_splitgroups'
#' @param countdata A data frame with 'gene_symbol' and count columns for each sample.
#' @param metadata A data frame with 'sample_id', 'tmPathoCondition', and covariates (e.g., PT, EC, FIB).
#' @param workers Number of parallel workers for DESeq2 computations.
#'
#' @return A list containing a tibble of DESeq2 results for each specified contrast.

### Choose settings
```{r}
# settings
# custom_design <- ~ PT + EC + FIB + bxSourcePatho
custom_design <- ~ MEAN_ID
shrinkage = FALSE
shrinkage_type = "normal"

if(shrinkage){
  contrast_method = paste0(shrinkage_type, "Shrinkage")
}else{
  contrast_method = "contrast"
}

print(contrast_method)

```


### Generate dds object
```{r, warning=FALSE}

start.time = Sys.time()

dds_list <- generate_dds(countdata = rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter %>% 
                           dplyr::select(gene_symbol_entrez_id, metadata_libSize_repFilter$SAMPLE_ID),
                    metadata = metadata_libSize_repFilter,
                    design = custom_design,
                    contrast = contrast_table)

end.time = Sys.time()
cat("Generating dds object took", round(end.time - start.time, 2), "hours")

# save dds file for later use
write_rds(dds_list, file.path(outputDir,"DESeq_output", contrast_method, "dds_list.rds"))

```


### Run contrast
```{r}

# dds_list <- read_rds(file.path(outputDir, "dds_list_groupPatho.rds"))

start.time = Sys.time()

deseq_output <- run_contrast(dds_list = dds_list, contrast = contrast_table, shrinkage = shrinkage, shrinkage_type = shrinkage_type)

end.time = Sys.time()
cat("Generating contrast took", round(end.time - start.time, 2), "hours")


```


# Generate TXG-MAPr upload files
```{r}
# expression_long
expression_long = deseq_output %>% 
  dplyr::select(sample_id = MEAN_ID_experiment,
                gene_symbol_entrez_id,
                log2fc = log2FoldChange,
                pvalue,
                padj)
  


# expression_wide
expression_wide = expression_long %>%
  pivot_wider(names_from = sample_id, names_glue = "{sample_id}__{.value}",
              values_from = c(log2fc, pvalue, padj))


TXG_upload_file = expression_long %>% left_join(metadata_libSize_repFilter %>% dplyr::select("experiment" = "COMPOUND", "time" = "TIME", "conc" = "DOSE", "sample_id" = "MEAN_ID", "REPLICATE"),
                                                by="sample_id") %>%
  separate(gene_symbol_entrez_id, into=c("gene_symbol", "entrez_id"), sep="_") %>%
  dplyr::rename("gene_id" = entrez_id) %>%
  mutate(experiment = paste0(experiment, REPLICATE)) %>%
  select(everything(), -"sample_id", -"REPLICATE", -"gene_symbol")

```



# Save output
```{r}
write_rds(deseq_output, file.path(outputDir, "updatedVersion", "DESeq_output", contrast_method, "deseq_output.rds"))

save(contrast_table,
     deseq_output,
     dds_list,
     
     file = file.path(outputDir, "updatedVersion", "DESeq_output", contrast_method, "DESeq2_output_files.RData")) 


data.table::fwrite(TXG_upload_file, file.path(outputDir, "DESeq_output", paste0(studyNR, "_TXG_upload_file_replicates_separate.txt")))

```


# Check DEGs
```{r}
log2fc_threshold = log2(1.5)
padj_threshold = 0.01


# add timepoint & concentration
deseq_output %>%
  separate(
    MEAN_ID_experiment,
    into = c("cellType", "experiment", "timepoint", "concentration", "replicate"),
    sep = "_",
    remove = FALSE
  ) %>%
  mutate(timepoint = factor(timepoint, levels = c("4hr", "8hr", "16hr", "24hr", "48hr", "72hr")),
         concentration = factor(concentration, levels = c("0.1uM", "0.5uM", "1uM", "2.5uM", "5uM", "10uM", "20uM", "30uM", "50uM"))) -> deseq_output_tpconc


#  plot
deseq_output_tpconc %>%
  filter(padj < padj_threshold & abs(log2FoldChange) >= log2fc_threshold) %>%
  mutate(direction = if_else(log2FoldChange >= 0, true = "UP", false = "DOWN")) %>%
  group_by(timepoint, concentration, direction) %>%
  summarize(n()) %>%
  ggplot(aes(x=concentration, y=`n()`, fill = direction)) +
  geom_bar(stat = "identity") +
  facet_wrap(~timepoint) +
  labs(y = "number of DEGs") +
  theme(axis.text.x = element_text(angle = 90),
        legend.title = element_blank()) +
  scale_fill_manual(values = c("firebrick", "darkblue"),
                     breaks = c("UP", "DOWN"))

ggsave(file.path(plotDir, "DEGs_per_timepoint_and_concentration.pdf"), width = 10, height = 8)

```


