---
title: "Quality Control"
subtitle: "EUT046 - Version 2"
date: "2025-05-23"
author: 
  Imke Bruns (0000-0001-9681-6693)
output:
  html_document:
    code_download: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 4
---

```{r, include=FALSE}

knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)

# load ggplot
pacman::p_load(ggplot2)

# set standard plot theme
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 14,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)

# choose project
studyNR = "EUT046"

```




# Set up analysis {.tabset}

## Load packages

### CRAN
Install and load CRAN packages.
```{r warning=FALSE, message=FALSE}
# if(!require(pacman, quietly = TRUE)){
#   install.packages("pacman")
# }

pacman::p_load(tidyverse, data.table, ggfortify, here)
```


### BioManager
Install and load Bioconductor packages.
```{r warning=FALSE, message=FALSE}
# if (!require("BiocManager", quietly = TRUE)){
#     install.packages("BiocManager")
# }
# 
# if (!require(DESeq2, quietly = TRUE)){
#   BiocManager::install("DESeq2")
# }

library(DESeq2)

```


## Load functions
Load manual functions we need in the script.
```{r}
source(file.path(here(), "functions_new.R"), local = knitr::knit_global())
```



## Set color scheme
Set standard color scheme for plots.
```{r}

# color list for plotting
color25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown", "forestgreen"
)   


```


## Set directories
Set directories.
```{r}
# data directory 
if(!dir.exists(paste0("data/"))){
  dir.create(path = file.path(here(), "data"))
}
dataDir <- file.path(here(), "data", studyNR)

# input directory
if(!dir.exists(paste0("input/", studyNR, "/"))){
  dir.create(path = file.path(here(), "input", studyNR))
}
inputDir <- file.path(here(), "input", studyNR)

# output directory
if(!dir.exists(paste0("output/", studyNR, "/"))){
  dir.create(path = file.path(here(), "output", studyNR))
}
outputDir <- file.path(here(), "output", studyNR)

# plot directory
if(!dir.exists(paste0("figures/", studyNR, "/"))){
  dir.create(path = file.path(here(), "figures", studyNR))
}
plotDir <- file.path(here(), "figures", studyNR)


```


## Thresholds for library size
Here, we set the threshold for the library size filter. We set this to be 20% of the expected read depth which is in this case 3 000 000.
```{r}
expectedReadDepth = 3000000
libSizeFilter = 0.2*expectedReadDepth

```


## Import data
Import needed data files.
```{r}
# raw counts
rawcounts = read.csv(file.path(inputDir, paste0(studyNR, "_counts_per_gene_per_sample.csv"))) %>%
    dplyr::rename("probe_id" = Column1)


# metadata
metadata = read.csv(file.path(inputDir, paste0(studyNR, "_metadata.csv"))) %>%
  mutate(
    TREATMENT_ID = gsub(" ", "", TREATMENT_ID),
    MEAN_ID = if_else(COMPOUND %in% c("DMEM", "DMSO", "NaCl"), true = TREATMENT_ID, false = paste0(TREATMENT_ID, "_R", REPLICATE))) %>%
  mutate(
    CONCENTRATION = factor(CONCENTRATION, levels = c("0uM", "0.1uM", "0.5uM", "1uM", "2.5uM", "5uM", "10uM", "20uM", "30uM", "50uM")),
    TIMEPOINT = factor(TIMEPOINT, levels = c("0hr", "4hr", "8hr", "16hr", "24hr", "48hr", "72hr")),
    DOSE = as.numeric(gsub("uM", "", CONCENTRATION))
  ) 

# manifest
manifest = read.delim(file.path(dirname(dataDir), "Manifests", "Temposeq_manifest_Human_WT_1.2_realignment_96percent_2023-05-15.txt"))

```


# Quality control 

## Probe QC {.tabset}
We start with the quality control for the probes.

### STEP 1: Probe alignment
Here, we align the probes with the manifest.
```{r}
rawcounts_entrezid = rawcounts %>% left_join(manifest %>% select(Probe_ID, gene_symbol_entrez_id),
                                             by = c("probe_id" = "Probe_ID")) %>% select(probe_id, gene_symbol_entrez_id, everything())

```


### STEP 2: Relevance filter
Apply relevance filter adapted from the R-ODAF pipeline (see: https://doi.org/10.1016/j.yrtph.2022.105143)
```{r warnings = FALSE}
relevance_summary = relevance_filter(
  countdata = apply(
    rawcounts_entrezid %>% column_to_rownames("probe_id") %>% select(-gene_symbol_entrez_id),
    2,
    cpm_normalization
  ) %>% as.data.frame(),
  metadata = metadata
  )


rawcounts_entrezid_relevanceFilter = rawcounts_entrezid %>%
  filter(probe_id %in% (relevance_summary %>%
                          filter(keep == TRUE) %>%
                          pull(probe_id))) 
# %>%
#   select(-probe_id)


cat("The relevance filter removed", nrow(rawcounts_entrezid) - nrow(rawcounts_entrezid_relevanceFilter), "probes.", nrow(rawcounts_entrezid_relevanceFilter), "are left for further analysis.")

```

### STEP 3: Spurious spikes
Group replicate groups + probe_id, calculate the ratio between the maximum of probeX and the sum of probeX for each replicate group. If this ratio exceeds 1.4*number of replicates^-0.66, flag. The flagged genes are in the `spike_summary` table.

```{r}

log2cpmcounts_entrezid_relevanceFilter_long = log2(apply(
  rawcounts_entrezid_relevanceFilter %>% column_to_rownames("probe_id") %>% select(-gene_symbol_entrez_id),
  2,
  cpm_normalization
) + 1) %>%
  as.data.frame() %>%
  rownames_to_column("probe_id") %>%
  mutate(gene_symbol_entrez_id = rawcounts_entrezid_relevanceFilter$gene_symbol_entrez_id) %>%
  pivot_longer(-c(gene_symbol_entrez_id, probe_id),
               names_to = "SAMPLE_ID",
               values_to = "log2cpm") %>% 
  left_join(metadata %>% select(SAMPLE_ID, TREATMENT_ID), by = "SAMPLE_ID")



# calculate spike ratios and thresholds per gene group
spike_check = log2cpmcounts_entrezid_relevanceFilter_long %>%
  group_by(probe_id, TREATMENT_ID) %>%
  summarize(
    max_count = max(log2cpm),
    sum_count = sum(log2cpm),
    n_samples = n(),
    ratio = if_else(sum_count == 0, true = 0, false = max_count / sum_count),
    threshold = 1.4*(n_samples^-0.66),
    spike = ratio >= threshold,
    .groups = "drop"
  )


# count how many groups have spikes per gene
spike_summary = spike_check %>%
  group_by(probe_id) %>%
  summarize(
    groups_with_spikes = sum(spike),
    pass_spike_filter = groups_with_spikes <= 1,
    .groups = "drop"
  )

head(spike_summary)

table(spike_summary$pass_spike_filter)

spike_flag = spike_summary %>% filter(pass_spike_filter == FALSE) %>% 
  left_join(log2cpmcounts_entrezid_relevanceFilter_long %>% select(probe_id, gene_symbol_entrez_id), by = "probe_id") %>%
  pull(gene_symbol_entrez_id) %>% unique()

cat("The spurious spike filter flagged", length(spike_flag), "genes.")


fwrite(spike_summary, file.path(outputDir, "updatedVersion", "spikeSummary.txt"))

```


### STEP 4: Sum probes
We sum the probes targeting the same gene.
```{r}
# count isomeric probes
cat(
  "Before sum counts isomeric probes, there are",
  length(
    rawcounts_entrezid_relevanceFilter$gene_symbol_entrez_id %>% duplicated() %>% which()
  ),
  "duplicated gene symbols \n"
)

rawcounts_entrezid_relevanceFilter_sumProbes = aggregate(
  rawcounts_entrezid_relevanceFilter %>% select(where(is.numeric)),
  by = list(gene_symbol_entrez_id = rawcounts_entrezid_relevanceFilter$gene_symbol_entrez_id),
  FUN = sum
)

cat("After all probe QC steps, there are", nrow(rawcounts_entrezid_relevanceFilter_sumProbes), "unique genes left")
```


### Probe QC output
Generate output files after the probe QC.
```{r}
# cpm counts
cpmcounts_entrezid_relevancefilter_sumProbes = apply(rawcounts_entrezid_relevanceFilter_sumProbes %>% column_to_rownames("gene_symbol_entrez_id"), 2, cpm_normalization) %>% as.data.frame() %>% rownames_to_column("gene_symbol_entrez_id")

# log2cpm counts
log2cpmcounts_entrezid_relevanceFilter_sumProbes = log2(cpmcounts_entrezid_relevancefilter_sumProbes %>% column_to_rownames("gene_symbol_entrez_id") + 1) %>% rownames_to_column("gene_symbol_entrez_id")

```


### Visualize in PCA 

Show low QC genes in PCA scale
```{r}
pcaData_beforeProbeFilter = log2(apply(
  rawcounts %>% column_to_rownames("probe_id"),
  2,
  cpm_normalization
) + 1) %>%
  as.data.frame() %>%
  rownames_to_column("probe_id") %>%
  mutate(
    removed_probeQC = if_else(
      probe_id %in% (relevance_summary %>%
                       filter(keep == TRUE) %>%
                       pull(probe_id)),
      true = "passed",
      false = "notPassed"
    ),
    flagged_probeQC = if_else(
      probe_id %in% (
        spike_summary %>%
          filter(pass_spike_filter == TRUE) %>%
          pull(probe_id)
      ),
      true = "passed",
      false = "spike_flagged"
    ),
    flagged_probeQC_10per = if_else(
      probe_id %in% (
        spike_summary %>%
          filter(groups_with_spikes <= 54*0.1) %>%
          pull(probe_id)
      ),
      true = "passed",
      false = "spike_flagged"
    )
  )
  

pcaOutput_beforeProbeFilter = prcomp(pcaData_beforeProbeFilter %>% select(where(is.numeric)))

autoplot(
  pcaOutput_beforeProbeFilter,
  data = pcaData_beforeProbeFilter,
  color = "removed_probeQC",
  size = 1
) +
  labs(title = "PCA gene expression (log2CPM)",
       subtitle = "Probes") +
  scale_color_manual(values = c("firebrick", "forestgreen"),
                     breaks = c("notPassed", "passed"))

```


Show flagged genes in PCA scale
```{r}
autoplot(
  pcaOutput_beforeProbeFilter,
  data = pcaData_beforeProbeFilter,
  color = "flagged_probeQC",
  size = 1
) +
  labs(title = "PCA gene expression (log2CPM)",
       subtitle = "Probes") +
  scale_color_manual(values = c("orange", "black"),
                     breaks = c("spike_flagged", "passed"))

```

Show flagged genes in PCA scale (now flagged probes are probes with a spike in >= 20% of the groups)
```{r}
autoplot(
  pcaOutput_beforeProbeFilter,
  data = pcaData_beforeProbeFilter,
  color = "flagged_probeQC_10per",
  size = 1
) +
  labs(title = "PCA gene expression (log2CPM)",
       subtitle = "Probes") +
  scale_color_manual(values = c("orange", "black"),
                     breaks = c("spike_flagged", "passed"))

```


### Save output
```{r}
save(
  # metadata
  metadata,
  # raw counts
  rawcounts_entrezid_relevanceFilter_sumProbes,
  # cpm counts
  cpmcounts_entrezid_relevancefilter_sumProbes,
  # log2cpm counts
  log2cpmcounts_entrezid_relevanceFilter_sumProbes,
  
  file = file.path(outputDir, "updatedVersion", paste0(studyNR, "_dataAfterProbeQC.RData"))
)
```



## Sample QC {.tabset}
Now we perform the quality control for the samples.

### STEP 5: Library size filter (NMR)
Remove samples if library size is < 20% of the intended read depth. The filter has been set on the top of the script.

#### Add library size to metadata
```{r}
metadata %>%
  mutate(
    LIBRARY_SIZE = colSums(rawcounts_entrezid_relevanceFilter_sumProbes %>% select(where(is.numeric))),
    LIBRARY_SIZE_PURE = colSums(rawcounts %>% select(where(is.numeric)))) -> metadata

```


#### Plot library size
Also compare pure library size obtained form raw counts with library size after probe QC
```{r}
# compare pure library size (from raw counts) with library size after probe QC
p1 = ggplot(metadata, aes(x=LIBRARY_SIZE_PURE, y=LIBRARY_SIZE)) +
  geom_point() +
  geom_abline()

p1

# show differences in table
metadata %>% 
  select(LIBRARY_SIZE, LIBRARY_SIZE_PURE) %>%
  head(10)


# plot library size per concentration/timepoint
p2 = ggplot(metadata, aes(x=CONCENTRATION, y=LIBRARY_SIZE, color = LIBRARY_SIZE > libSizeFilter, size = LIBRARY_SIZE > libSizeFilter)) + 
  geom_point() +
  facet_wrap(~TIMEPOINT) + 
  geom_hline(yintercept=libSizeFilter, lty=2, color="firebrick", linewidth=0.8) +
  scale_color_manual(values = c("firebrick", "black"),
                     breaks = c(FALSE, TRUE)) +
  scale_size_manual(values = c(1, 2.5),
                    breaks = c(TRUE, FALSE)) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Library size per concentration/timepoint")

p2

```

#### Remove samples

From metadata
```{r}
cat("We filter all samples with a library size of <", libSizeFilter, "\n")

metadata_libSize = metadata %>%
  filter(LIBRARY_SIZE >= libSizeFilter)

cat("The following samples have been removed due to low library size:\n",
    paste(metadata %>% filter(LIBRARY_SIZE < libSizeFilter) %>% pull(SAMPLE_ID), collapse = ", "))

```

From counts
```{r}
# raw counts
rawcounts_entrezid_relevanceFilter_sumProbes_libSize = rawcounts_entrezid_relevanceFilter_sumProbes %>% select(gene_symbol_entrez_id, all_of(metadata_libSize$SAMPLE_ID))

# cpm counts
cpmcounts_entrezid_relevancefilter_sumProbes_libSize = cpmcounts_entrezid_relevancefilter_sumProbes %>% select(gene_symbol_entrez_id, all_of(metadata_libSize$SAMPLE_ID))

# log2cpm counts
log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize = log2cpmcounts_entrezid_relevanceFilter_sumProbes %>% select(gene_symbol_entrez_id, all_of(metadata_libSize$SAMPLE_ID))

```



### STEP 6: Replicate correlation (rCOR)
We perform replicate correlation on the log2cpm counts, to minimize the effect of outliers.


#### Pairwise
Calculate replicate correlation for each pair combination

```{r}
replicateFilterOutput <- data.frame()

for(i in unique(metadata_libSize$TREATMENT_ID)){
  metaDataReps <- metadata_libSize %>% 
    filter(TREATMENT_ID == i)
  
  if(nrow(metaDataReps) > 1) {
    log2cpmDataReps <- log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize[, metaDataReps$SAMPLE_ID] 
    
    colnames(log2cpmDataReps)
    
    for(j in 1:ncol(log2cpmDataReps)){
      for(k in 1:ncol(log2cpmDataReps)){
        sample_A <- colnames(log2cpmDataReps)[j]
        sample_B <- colnames(log2cpmDataReps)[k]
        
        if(sample_A != sample_B){
        
        paste0("Looking at ", sample_A, " vs. ", sample_B)
        
        r2 <- cor(log2cpmDataReps[,j], log2cpmDataReps[,k])
        
        replicateFilterOutput <- rbind(replicateFilterOutput, 
                        data.frame(TREATMENT_ID = i, 
                                   sample_A = sample_A,
                                   sample_B = sample_B,
                                   comp = paste0(sample_A, " v. ", sample_B),
                                   r2 = r2))
        }
      }
    }
  }
}

replicateFilterOutput <- replicateFilterOutput %>% arrange(TREATMENT_ID)

head(replicateFilterOutput)

```

Calculate the highest replicate correlation for each sample
```{r}

replicateFilterOutput <- replicateFilterOutput %>% 
  separate(TREATMENT_ID, into = c("cell_line", "compound", "time", "concentration"), sep = "_", remove = FALSE) %>%
  group_by(sample_A) %>% 
  mutate(max_r2 = max(r2, na.rm=T),
         factor_conc = as.numeric(gsub("uM", "", concentration)),
             concentration = factor(concentration, levels = c("0uM","0.1uM", "0.5uM", "1uM", "2.5uM", "5uM", "10uM", "20uM", "30uM", "50uM")),
            time = factor(time, levels = c("0hr", "4hr", "8hr", "16hr", "24hr", "48hr", "72hr"))) %>% 
  ungroup()

loop <- unique(replicateFilterOutput$time)
loop

```
Look at each timepoint
```{r}

for(i in loop){
  print(
    replicateFilterOutput %>% 
      filter(time %in% i) %>% 
      ggplot(aes(x=reorder(sample_A, factor_conc), y=r2)) +
      geom_boxplot(color='grey90') +
      geom_point(color='grey60') +
      geom_point(aes(y=max_r2, fill=concentration), color='grey60', shape=21, size=3) +
      ylab('') + xlab('Sample ID') + ggtitle(i) +
      theme_bw() +
      theme(axis.text.x = element_blank(),
            legend.position = 'none',
            plot.title = element_text(hjust = 0.5, face = "bold", size = 18)) +
      facet_wrap(~time+concentration, scales='free_x', nrow = 1)
  )
}


```

Look at the distribution of maximum pairwise correlations
```{r}
replicateFilterMax <- replicateFilterOutput %>% 
  select(time, max_r2) %>%
  distinct()

replicateFilterMax %>% 
  group_by(time) %>% 
  ungroup() %>% 
  ggplot(aes(y=time, x=max_r2)) +
  geom_boxplot() + ylab('')

```

Some compounds have less variation in max r2 than others, but we want to capture acceptable variability. 

Make an IQR for the test system (RPTEC-TERT1) and then remove max r2 that are more than 3 IQR from the best. Flag samples as an outlier if their highest obtained replicate correlation is further beneath the experimental medi than the best performing replicate correlation. 

```{r}

replicateFilterDetect <- replicateFilterOutput %>% 
  select(TREATMENT_ID, r2, max_r2) %>% 
  distinct() %>% 
  mutate(testsystem_IQR = 3*IQR(r2)) %>% 
  group_by(TREATMENT_ID) %>% 
  mutate(experimental_max = max(max_r2),
         experimental_threshold = experimental_max - testsystem_IQR) %>% 
  ungroup() %>% 
  select(-r2, -max_r2) %>% 
  distinct()

replicateFilterNew <- left_join(replicateFilterOutput, replicateFilterDetect, by="TREATMENT_ID") %>% distinct()

replicateFilterNew <- replicateFilterNew %>% 
  mutate(keep_replicate = ifelse(max_r2 < experimental_threshold, "No", "Yes"))

for(i in loop){
print(replicateFilterNew %>% 
  filter(time == i) %>% 
  ggplot(aes(x=reorder(sample_A, -factor_conc))) +
  geom_boxplot(aes(y=r2), color='grey80') +
    geom_hline(aes(yintercept = experimental_max), color='grey70') +
    geom_hline(aes(yintercept = experimental_threshold), color='grey70', linetype='dashed') +
  geom_point(aes(y=r2), color='grey80') +
  geom_point(aes(y=max_r2, color=keep_replicate)) +
      scale_color_manual(values = c("Yes" = "#6DACBC", "No" = "#CF7976")) +
    ggtitle(i) + xlab('Sample') + ylab('') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)) +
  facet_wrap(~time+concentration, scales="free_x", nrow=1))
}

cat("The replicate filter removes", replicateFilterNew %>% filter(keep_replicate == "No") %>% pull(sample_A) %>% unique() %>% length(), "samples.")


```
#### Remove samples

From metadata
```{r}
keepSamples = replicateFilterNew %>%
           filter(keep_replicate == "Yes") %>% 
           pull(sample_A) %>% unique()

metadata_libSize_repFilter = metadata_libSize %>%
  filter(SAMPLE_ID %in% keepSamples)

cat("The following samples have been removed due to low replicate correlation:\n",
    paste(metadata_libSize %>% filter(!SAMPLE_ID %in% keepSamples) %>% pull(SAMPLE_ID), collapse = ", "))

```

From counts
```{r}
# raw counts
rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter = rawcounts_entrezid_relevanceFilter_sumProbes_libSize %>% select(gene_symbol_entrez_id, metadata_libSize_repFilter$SAMPLE_ID)

# cpm counts
cpmcounts_entrezid_relevancefilter_sumProbes_libSize_repFilter = cpmcounts_entrezid_relevancefilter_sumProbes_libSize %>% select(gene_symbol_entrez_id, metadata_libSize_repFilter$SAMPLE_ID)

# log2cpm counts
log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter = log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize %>% select(gene_symbol_entrez_id, metadata_libSize_repFilter$SAMPLE_ID)

```


### STEP 7: Viable Cells (FrVC)
Flag samples with < 70% viable cells. For this dataset, this information is not available.


### STEP 8: Fraction uniquely mapped reads (FMR)
Reject samples if percentage mapped is < 70%. For this dataset, this information is not available.


### STEP 9: Number of genes with >= 5 uniquely mapped reads (Ncov 5)
Flag samples if less than 5000 genes have >= 5 uniquely mapped reads.

```{r}

# create long df
rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter_long = rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter %>%
  pivot_longer(-gene_symbol_entrez_id, names_to = "SAMPLE_ID", values_to = "counts") 

Ncov5_summary = rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter_long %>%
  group_by(SAMPLE_ID) %>%
  summarize(
    genesAbove5 = sum(counts > 5),
    lowQuality = genesAbove5 < 5000,
    .groups = "drop"
  )

head(Ncov5_summary)

table(Ncov5_summary$lowQuality)

Ncov5_flag = Ncov5_summary %>%
  filter(lowQuality == TRUE) %>%
  pull(SAMPLE_ID)

```



### STEP 10: Number of probes capture top 80% of signal in sample (Nsig 80)
Flag samples if < 1000 genes capture the top 80% of the signal in a sample.

```{r}

Nsig80_summary = rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter_long %>%
  group_by(SAMPLE_ID) %>%
  arrange(SAMPLE_ID, desc(counts)) %>%
  mutate(
    cum_sum = cumsum(counts),
    total = sum(counts),
    cum_frac = cum_sum / total,
    gene_rank = row_number()
  ) %>%
  filter(cum_frac >= 0.8) %>%
  summarize(
    Nsig80 = min(gene_rank),
    lowComplexity = Nsig80 < 1000,
    .groups = "drop"
  )


head(Nsig80_summary)

table(Nsig80_summary$lowComplexity)

Nsig80_flag = Nsig80_summary %>%
  filter(lowComplexity == TRUE) %>% 
  pull(SAMPLE_ID)

```


### Add sample flags to metadata
```{r}
metadata_libSize_repFilter %>%
  mutate(flagged_sampleQC = case_when(
    SAMPLE_ID %in% Ncov5_flag ~ "Ncov5_flag",
    SAMPLE_ID %in% Nsig80_flag ~ "Nsig80_flag",
    .default = "passed"
  )) -> metadata_libSize_repFilter

```


### Check filters in PCA 
Check the removed samples in PCA, do we see additional outliers, are these removed if we increase the libsize filter/adjust the replicate correlation? 

Before the low quality samples have been removed
```{r}
pcaData_beforeSampFilter = log2cpmcounts_entrezid_relevanceFilter_sumProbes %>% column_to_rownames("gene_symbol_entrez_id") %>% t() %>% as.data.frame() %>% rownames_to_column("SAMPLE_ID") %>% left_join(metadata, by = "SAMPLE_ID") %>%
  mutate(removed_sampleQC = if_else(SAMPLE_ID %in% metadata_libSize_repFilter$SAMPLE_ID, true = "passed", false = "notPassed"),
         Ncov5_flag = if_else(SAMPLE_ID %in% Ncov5_flag, true = "flag", false = "passed"),
         Nsig80_flag = if_else(SAMPLE_ID %in% Nsig80_flag, true = "flag", false = "passed"))

pcaOutput_beforeSampFilter = prcomp(pcaData_beforeSampFilter %>% column_to_rownames("SAMPLE_ID") %>% select(log2cpmcounts_entrezid_relevanceFilter_sumProbes$gene_symbol_entrez_id))

autoplot(
  pcaOutput_beforeSampFilter,
  data = pcaData_beforeSampFilter,
  color = "CONCENTRATION",
  shape = "TIMEPOINT",
  size = 2
) + # change to 3
  labs(title = "PCA gene expression (log2CPM)") +
  scale_shape_manual(values = c(15, 16, 17, 18, 7, 8, 13)) +
  scale_color_manual(values = color25)

```

Color removed samples
```{r}
autoplot(
  pcaOutput_beforeSampFilter,
  data = pcaData_beforeSampFilter,
  color = "removed_sampleQC",
  shape = "TIMEPOINT",
  size = 2
) + # change to 3
  labs(title = "PCA gene expression (log2CPM)") +
  scale_shape_manual(values = c(15, 16, 17, 18, 7, 8, 13)) +
  scale_color_manual(values = c("firebrick", "forestgreen"),
                     breaks = c("notPassed", "passed"))

```

Check flagged samples
```{r}

autoplot(
  pcaOutput_beforeSampFilter,
  data = pcaData_beforeSampFilter,
  color = "Nsig80_flag",
  shape = "TIMEPOINT",
  size = 2
) + # change to 3
  labs(title = "PCA gene expression (log2CPM)") +
  scale_shape_manual(values = c(15, 16, 17, 18, 7, 8, 13)) +
  scale_color_manual(values = c("firebrick", "forestgreen"),
                     breaks = c("flag", "passed"))


```



After the low quality samples have been removed
```{r}
pcaData_afterSampleFilter = log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter %>% column_to_rownames("gene_symbol_entrez_id") %>% t() %>% as.data.frame() %>% rownames_to_column("SAMPLE_ID") %>% left_join(metadata_libSize_repFilter, by = "SAMPLE_ID")

pcaOutput_afterSampleFilter = prcomp(pcaData_afterSampleFilter %>% column_to_rownames("SAMPLE_ID") %>% select(log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter$gene_symbol_entrez_id))

autoplot(
  pcaOutput_afterSampleFilter,
  data = pcaData_afterSampleFilter,
  color = "CONCENTRATION",
  shape = "TIMEPOINT",
  size = 2
) + # change to 3
  labs(title = "PCA gene expression (log2CPM)") +
  scale_shape_manual(values = c(15, 16, 17, 18, 7, 8, 13)) +
  scale_color_manual(values = color25)

```


Color flagged samples after filter
```{r}
autoplot(
  pcaOutput_afterSampleFilter,
  data = pcaData_afterSampleFilter,
  color = "flagged_sampleQC",
  shape = "TIMEPOINT",
  size = 2
) + # change to 3
  labs(title = "PCA gene expression (log2CPM)") +
  scale_shape_manual(values = c(15, 16, 17, 18, 7, 8, 13)) +
  scale_color_manual(values = c("black", "orange", "purple"),
                     breaks = c("passed", "Nsig80_flag", "Ncov5_flag"))

```
The flagged samples are nicely clustered with their replicate, so for now I decided to keep them in.


### Save data
```{r}
save(
  # metadata
  metadata_libSize_repFilter,
  
  # raw counts
  rawcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter,
  
  # cpm counts
  cpmcounts_entrezid_relevancefilter_sumProbes_libSize_repFilter,
  
  # log2cpm counts
  log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter,
  
  file = file.path(outputDir, "updatedVersion", paste0(studyNR, "_dataAfterSampleQC_3IQR_repfilter.RData"))
  
)


fwrite(metadata_libSize_repFilter, file.path(outputDir, "updatedVersion", "sample_metadata.txt"))
fwrite(log2cpmcounts_entrezid_relevanceFilter_sumProbes_libSize_repFilter, file.path(outputDir, "updatedVersion", "normalized_counts.txt"))

```

