---
title: "BMD Reporting Template (Generalized)"
author: "Your Lab / Team"
date: "`r format(Sys.Date())`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true

params:
  study_id: "EUT046"
  input_dir: "input/EUT046/newTemplate"
  output_dir: "output/EUT046"
  figures_dir: "figures/EUT046"

  run_bmd_log2fc: true
  run_bmd_pathway_scores: true
  have_median_pathway_scores: true
  have_williams_trend_test: true

  files:
    value:
      sample_metadata: "sample_metadata.txt"
      normalized_counts: "normalized_counts.txt"
      log2fc_table: "log2fc_table.txt"
      pathway_overview: "pathway_overview.txt"
      bmd_gene_log2cpm: "BMD_output_normalized_counts.txt"
      bmd_gene_log2fc: "BMD_output_log2fc.txt"
      bmd_pathway_scores: "BMD_output_pathway_scores.txt"
      median_bmd_normalized_counts: "medianBMD_pathway_normalized_counts.txt"
      median_bmd_log2fc: "medianBMD_pathways_log2fc.txt"
      wtt_log2cpm: "WTT_normalized_counts.txt"
      wtt_log2fc: "WTT_log2fc.txt"
      wtt_pathway_scores: "WTT_pathway_scores.txt"

  columns:
    value:
      metadata:
        sample_id: ["sample_id", "sampleid", "id"]
        dose: ["dose", "conc", "concentration", "dose_um", "dose_uM"]
        timepoint: ["timepoint", "time", "tp", "hrs", "hours"]
        compound: ["compound", "treatment", "chemical", "agent"]
        replicate: ["replicate", "rep", "bio_rep", "tech_rep"]
      gene_table:
        gene_id: ["gene_symbol_entrez_id", "gene", "symbol", "gene_id"]
      pathway_overview:
        gene_id: ["gene_symbol", "gene", "symbol", "gene_id"] # adjust later
        pathway: ["module", "pathway", "network", "set", "term"]
      bmd_gene:
        gene_id: ["gene_symbol_entrez_id", "gene", "symbol", "gene_id"]
        timepoint: ["timepoint", "tp", "hours", "hrs"]
        bmd: ["Best.BMD", "bmd"]
        bmdl: ["Best.BMDL", "bmdl"]
        bmdu: ["Best.BMDU", "bmdu"]
      bmd_pathway:
        pathway_id: ["module", "pathway"]
        timepoint: ["timepoint", "tp", "hours", "hrs"]
        bmd: ["Best.BMD", "bmd"]
        bmdl: ["Best.BMDL", "bmdl"]
        bmdu: ["Best.BMDU", "bmdu"]
      bmd_median_pathway:
        pathway: ["pathway", "module", "network", "set", "term", "GO.Pathway.Gene.Set.Gene.Name"]
        timepoint: ["timepoint", "tp", "hours", "hrs"]
        bmd: ["BMD.Median", "bmd"]
        bmdl: ["BMDL.Median", "bmdl"]
        bmdu: ["BMDU.Median", "bmdu"]       
      wtt:
        gene_id: ["gene_symbol_entrez_id", "Probe.ID"]
        timepoint: ["timepoint", "tp", "hours", "hrs"]
        adj_p: ["Adjusted.P.Value", "padj", "fdr", "qvalue"]
        p: ["P.Value", "p", "pval", "Unadjusted.P.Value"]
        max_fold_change: ["Max.Fold.Change", "max_fc"]
      wtt_pathway:
        pathway_id: ["pathway", "module"]
        timepoint: ["timepoint", "tp", "hours", "hrs"]
        adj_p: ["Adjusted.P.Value", "padj", "fdr", "qvalue"]
        p: ["P.Value", "p", "pval", "Unadjusted.P.Value"]
        max_fold_change: ["Max.Fold.Change", "max_fc"]        
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
```

## Load packages
```{r}
pacman::p_load(
  tidyverse, readr, readxl, janitor, glue, here, fs,
  ggfortify, gridExtra, pheatmap, plotly, DT, knitr, cowplot, kableExtra, scales, data.table
)
```

## Directories
```{r}
# Resolve & create directories if needed
input_dir   <- fs::path(params$input_dir)
output_dir  <- fs::path(params$output_dir)
fig_dir     <- fs::path(params$figures_dir)

fs::dir_create(input_dir)
fs::dir_create(output_dir)
fs::dir_create(fig_dir)
```

## Helpers
Reusable utilities for reading, standardizing, and checking data.

```{r}
source("R/read_utils.R")
```


## Import data
```{r}
files <- params$files
cols  <- params$columns

# Sample metadata
sample_metadata <- read_any(files$sample_metadata) %>%
  rename_using_mapping(cols$metadata) %>%
  require_cols(c("sample_id","dose","timepoint"), "sample_metadata") %>%
  mutate(
    dose = as.numeric(dose),
    timepoint = as.character(timepoint),
    timepoint = order_tp(timepoint)
  )


# Normalized counts (genes x samples); first column = gene id
normalized_counts <- read_any(files$normalized_counts)

# Map synonyms to standard names
normalized_counts <- rename_using_mapping(normalized_counts, cols$gene_table)

# Ensure we have a 'gene_id' column; if not, promote the first matching synonym
if (!"gene_id" %in% names(normalized_counts)) {
  gene_syn <- janitor::make_clean_names(unlist(cols$gene_table$gene_id, use.names = FALSE))
  hit <- intersect(gene_syn, names(normalized_counts))
  if (length(hit) >= 1) {
    names(normalized_counts)[match(hit[1], names(normalized_counts))] <- "gene_id"
  }
}


# Optional log2FC table
log2fc_table <- NULL
if (params$run_bmd_log2fc && fs::file_exists(fs::path(input_dir, files$log2fc_table))) {
  log2fc_table <- read_any(files$log2fc_table)
  colnames(log2fc_table)[1] <- "gene_id"
}

# Pathway overview (gene -> pathway)
pathway_overview <- read_any(files$pathway_overview) %>%
  rename_using_mapping(cols$pathway_overview) %>%
  require_cols(c("gene_id", "pathway"), "pathway_overview")



# --- BMD outputs (gene-level) ---
BMD_gene_log2cpm <- read_any(files$bmd_gene_log2cpm) %>%
  rename_using_mapping(cols$bmd_gene) %>%
  require_cols(c("gene_id","timepoint","bmd","bmdl","bmdu"), "BMD gene log2cpm")

BMD_gene_log2fc <- NULL
if (params$run_bmd_log2fc && fs::file_exists(fs::path(input_dir, files$bmd_gene_log2fc))) {
  BMD_gene_log2fc <- read_any(files$bmd_gene_log2fc) %>%
    rename_using_mapping(cols$bmd_gene) %>%
    require_cols(c("gene_id","timepoint","bmd","bmdl","bmdu"), "BMD gene log2fc")
}

# --- BMD outputs (pathway-level) ---
BMD_pathway_scores <- NULL
if (params$run_bmd_pathway_scores && fs::file_exists(fs::path(input_dir, files$bmd_pathway_scores))) {
  BMD_pathway_scores <- read_any(files$bmd_pathway_scores) %>%
    rename_using_mapping(cols$bmd_pathway) %>%
    require_cols(c("pathway_id","timepoint","bmd","bmdl","bmdu"), "BMD pathway scores")
}

medianBMD_pathways_from_genes <- NULL
if (params$have_median_pathway_scores && fs::file_exists(fs::path(input_dir, files$median_bmd_normalized_counts))) {
  medianBMD_pathways_from_genes <- read_any(files$median_bmd_normalized_counts) %>%
    rename_using_mapping(cols$bmd_median_pathway) %>%
    require_cols(c("pathway","bmd","bmdl","bmdu", "timepoint"), "median pathway BMD from genes")
}

# --- Williams Trend Test (optional) ---
WTT_log2cpm <- NULL
if (params$have_williams_trend_test && fs::file_exists(fs::path(input_dir, files$wtt_log2cpm))) {
  WTT_log2cpm <- read_any(files$wtt_log2cpm) %>%
    rename_using_mapping(cols$wtt) %>%
    require_cols(c("gene_id","timepoint","adj_p", "p", "max_fold_change"), "WTT log2cpm")
}

WTT_log2fc <- NULL
if (params$have_williams_trend_test && params$run_bmd_log2fc && fs::file_exists(fs::path(input_dir, files$wtt_log2fc))) {
  WTT_log2fc <- read_any(files$wtt_log2fc) %>%
    rename_using_mapping(cols$wtt) %>%
    require_cols(c("gene_id","timepoint","adj_p", "p", "max_fold_change"), "WTT log2fc")
}

WTT_pathway_scores <- NULL
if (params$have_williams_trend_test && params$run_bmd_pathway_scores && fs::file_exists(fs::path(input_dir, files$wtt_pathway_scores))) {
  WTT_pathways <- read_any(files$wtt_pathway_scores) %>%
    rename_using_mapping(cols$wtt_pathway) %>%
    require_cols(c("pathway_id","timepoint","adj_p", "p", "max_fold_change"), "WTT pathway scores")
}

```


## Distribution based 

### Dose Responsive Genes (WTT) {.tabset}

```{r}
source("R/wtt_results.R")
```

#### Gene level {.tabset}

##### log2cpm
```{r}

doseresponsive_plot(WTT = WTT_log2cpm, level = "genes", p_col = "p", p_thresh = 0.05)

dose_responsive_table(WTT = WTT_log2cpm)

```

##### log2fc
```{r}

doseresponsive_plot(WTT = WTT_log2fc, level = "genes", p_col = "p", p_thresh = 0.05)

dose_responsive_table(WTT = WTT_log2fc)

```

#### Network level
```{r}

doseresponsive_plot(WTT = WTT_pathways, level = "modules", p_col = "p", p_thresh = 0.05)

dose_responsive_table(WTT = WTT_pathways)

```



### BMD accumulation plots {.tabset}
Accumulative distribution of BMDs with standard QC filters applied.

#### Gene level {.tabset}
```{r}
source("R/plots_accumulation.R")
```


##### Normalized counts
```{r}
accumulation_plot(BMD_gene_log2cpm, title_prefix = "Gene")
```

##### log2fc
```{r}
accumulation_plot(BMD_gene_log2fc, title_prefix = "Gene")

```

#### Network level
```{r}
accumulation_plot(BMD_pathway_scores, title_prefix = "Module")

```


### Distribution based tPODs {.tabset}

#### Gene level {.tabset}
```{r}
source("R/calculate_distributionBased_tPODs.R")
```


##### Normalized counts
```{r}
bmd_summary_table(BMD_gene_log2cpm, caption = "Gene-level BMD metrics (log2CPM)")

# Get the underlying data frame:
tbl <- bmd_summary_table(BMD_gene_log2cpm, return_data = TRUE)
data.table::fwrite(tbl, file.path(output_dir, "distribution_based_tPODs_normalized_counts.txt"), sep = "\t")

```

##### Log2fc
```{r}
bmd_summary_table(BMD_gene_log2fc, caption = "Gene-level BMD metrics (log2FC)")

# Get the underlying data frame:
tbl <- bmd_summary_table(BMD_gene_log2fc, return_data = TRUE)
data.table::fwrite(tbl, file.path(output_dir, "distribution_based_tPODs_log2fc.txt"), sep = "\t")

```



#### Network level
```{r}
bmd_summary_table(BMD_pathway_scores, caption = "Pathway-level BMD metrics (modules)")

# Get the underlying data frame:
tbl <- bmd_summary_table(BMD_pathway_scores, return_data = TRUE)
data.table::fwrite(tbl, file.path(output_dir, "distribution_based_tPODs_pathway_scores.txt"), sep = "\t")
```



## Gene set based analysis
Select top 5 networks with the lowest median BMD per timepoint; indicating the most sensitive pathways to compound exposure

```{r}
source("R/table_lowest_median_pathways.R")

top_pathways_per_timepoint(df = medianBMD_pathways_from_genes)

```



