---
title: "RODAF BMD reporting"
author: "IBBruns"
date: "2024-07-04"
output: 
html_document:
  code_folding: hide
  toc: true
  toc_depth: 4
  toc_float: true
  number_sections: true
---


```{r setup, include=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)


# set standard plot theme
library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 14,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)


```


# Settings {.tabset}

## Manually set parameters
Here you assign which analysis you have performed
```{r}
# choose project
studyNR = "EUT046"

# did you run the BMD also for log2fc values? If yes: TRUE, else: FALSE
runBMDonLog2fc = TRUE

# did you perform a williams trend test and do you have the output? If yes: TRUE, else: FALSE
williamsTrendTest = TRUE

# did you run the BMD on pathway enrichment scores? If yes: TRUE, else: FALSE (NOTE: if you determine pathway BMDs based on the median gene BMD assign FALSE)
runBMDonNetworkScores = TRUE

# did you upload enrichment scores (e.g. eigengene scores/normalized enrichment scores)
upload_enrichmentScores = TRUE
```


## Load packages
```{r, class.source='fold-show'}
pacman::p_load(tidyverse, ggfortify, gridExtra, pheatmap, plotly, DT, knitr, cowplot, kableExtra)
```


## Set directories
```{r}
# data directory
if(!dir.exists(paste0("data/"))){
  dir.create(path = file.path(getwd(), "data"))
}
dataDir <- file.path(getwd(), "data")

# input directory
if(!dir.exists(paste0("input/", studyNR, "/RODAF/"))){
  dir.create(path = file.path(getwd(), "input", studyNR, "RODAF"))
}
inputDir <- file.path(getwd(), "input", studyNR, "RODAF")

```


## Import data
Load data, we need:  
REQUIRED:    
1. `sample_metadata` = metadata provided with the raw data    
2. `normalized_counts` = normalized counts (e.g. CPM or log2CPM). sample_id should be as column names, Probe_ID in the first column, so genes as rows.  
3. `pathway_overview` = overview of which gene belongs to which pathway. Data should contain the following columns: Probe_ID, pathway.   
4. `BMD_output_normalized_counts` = output van BMD modeling. Data should at least contain the following columns: Probe_ID, Best.BMD (this can also be the average BMD of all your models), Best.BMDL (BMDL from your best model (or average)), Best.BMDU, Best.R.Squared. NOTE: these are needed to apply the filters I did, if you want to apply different filters, you need different columns, can be adjusted in the code.   
5. `medianBMD_pathways_normalized_counts` = median BMD for genes in your assessed pathways. Data should contain at least the following columns: Genes.That.Passed.All.Filters (which indicates the number of genes per pathway after all selection criteria), Pathway.Name containing the names of the pathways.   

OPTIONAL:   
6. `WTT_normalized_counts` = Williams Trend Test output (only needed when WilliamsTrendTest = TRUE). Here at least the following columns are necessary: Max.Fold.Change (indicates the direction of the response (i.e. up- or downregulation)), Unadjusted.P.Value, Adjusted.P.Value (depending on which filter you want to apply).   
7. `enrichment_scores` = enrichment scores (e.g. normalized enrichment scores after performing a GSEA or eigengene scores). Data should contain the following columns: sample_id, pathway, enrichment_score.   

NOTE:   
if runBMDonLog2fc = TRUE, we also need the following input:   
8. `log2fc` = log2fc values. The following columns are needed: sample_id, Probe_ID, log2fc.   
9. `BMD_output_log2fc` = BMD output when ran on log2fc values.   
10. `WTT_log2fc` = WTT output when ran on log2fc values (optional).   

if runBMDonNetworkScores = TRUE, we also need the following input:   
11. `BMD_output_pathway` = BMD output when ran on pathway enrichment scores   
12. `WTT_pathway` = WTT output when ran on pathway enrichment scores (optional).   


  
```{r}

load(file.path(inputDir, "R-ODAF_DESeq2_output_files.RData"))

# here load the pathway gene association; one column all Probe_ID, other column pathway association.
pathway_overview = read.delim(file.path(inputDir, paste0(studyNR, "_gene_module_overview.txt"))) %>%
  mutate(
    gene_symbol = gsub("id_", "", gene_symbol),
    entrez_id = gsub("id_", "", entrez_id)
  ) %>%
  distinct(gene_symbol, module) %>%
  dplyr::rename("pathway" = module)


if(upload_enrichmentScores){
  # here load eigengene scores/NES scores: column names are sample_id, pathway and enrichment_score
enrichment_scores =  read.delim(file.path(inputDir, paste0(studyNR, "_R-ODAF_TXGMAPr_module_egs.txt"))) %>%
  dplyr::rename("enrichment_score" = eg_score,
                "pathway" = module) %>%
  select(sample_id, pathway, enrichment_score) 
}


if(runBMDonLog2fc){
  log2fc = expression_long # load log2fc values in long format (sample_id, gene_symbol_entrez_id, log2fc, pvalue, padj)
  
  if(williamsTrendTest){
    WTT_log2fc = read.delim(file.path(inputDir, paste0(studyNR, "_RODAF_log2fc_WTT_output.txt"))) %>%
    mutate(timepoint = str_extract(Analysis, "\\d+hr"))
  }
  
  BMD_output_log2fc = read.delim(file.path(inputDir, paste0(studyNR, "_RODAF_log2fc_BMD_output.txt"))) %>%
  mutate(timepoint = str_extract(Analysis, "\\d+hr")) %>%
    separate("Probe.ID", into=c("gene_symbol", "probe_NR"), remove = FALSE, sep="_") %>%
  left_join(pathway_overview, by = "gene_symbol") %>% # add the pathway assignment to each gene
  select(-"gene_symbol", -"probe_NR")
    
  # medianBMD_pathways_log2fc = read.delim(file.path(inputDir, paste0(studyNR, "_log2fc_functionalClas_output.txt"))) %>%
  # mutate(timepoint = str_extract(Analysis, "\\d+hr")) %>% dplyr::rename("Pathway.Name" = GO.Pathway.Gene.Set.Gene.Name)
}



if(runBMDonNetworkScores) {
  BMD_output_pathway = read.delim(file.path(inputDir, paste0(studyNR, "_RODAF_module_egs_BMD_output.txt"))) %>%
    mutate(timepoint = str_extract(Analysis, "\\d+hr")) 
  
  if (williamsTrendTest) {
    WTT_pathway = read.delim(file.path(inputDir, paste0(
      studyNR, "_RODAF_module_egs_WTT_output.txt"
    ))) %>%
      mutate(timepoint = str_extract(Analysis, "\\d+hr"))
  }
}


# txgmapr_enrichment = readRDS(file.path(dataDir, "module_enrichment.rds"))
# 
# txgmapr_tf_enrichment = readRDS(file.path(dataDir, "tf_enrichment.rds"))

```



# Williams Trend Test (WTT) {.tabset}

Plot the output of the Williams Trend Test (WTT) to get an idea of the number of responsive genes and networks  
We used a threshold of padj < 0.05 to identify dose-responsive genes and no fold change filter was applied

```{r}

doseresponsives_plot = function(WTToutput, level){
  
  # add timepoint and regulation to data
  WTToutput = WTToutput %>%
    mutate(regulation = if_else(WTToutput$Max.Fold.Change > 0, true = "upregulated", false = "downregulated")) %>%
    filter(Unadjusted.P.Value < 0.05)
  
  
  # make sure that all timepoints are in final figure (also timepoints without significant dose/responses)
   all_combinations <- expand.grid(
     timepoint = c("4hr", "8hr", "16hr", "24hr", "48hr", "72hr"),
     regulation = WTToutput %>% pull(regulation) %>% unique()
   )
   
   counts = WTToutput %>%
     group_by(timepoint, regulation) %>%
     summarize(count = n()) %>%
     right_join(all_combinations, by = c("timepoint", "regulation")) %>%
     mutate(count = coalesce(count, 0))
   

   # change timepoint to hours
  counts$timepoint = gsub("hr", "", counts$timepoint)
  counts$timepoint = paste0(counts$timepoint, "_hours")
  counts$timepoint = factor(counts$timepoint, levels = c("4_hours", "8_hours", "16_hours", "24_hours", "48_hours", "72_hours"))
  
  # create a barplot
  ggplot(counts, aes(x = timepoint, y = count, fill = regulation)) +
    geom_bar(stat = "identity", position = "stack") + 
    labs(title = paste0("Number of dose responsive ", level, " per timepoint"),
         x = "", 
         y = paste0("Number of dose responsive ", level)) + 
    scale_fill_manual(values = c("upregulated" = "firebrick", "downregulated" = "darkblue"), name = "Regulation")
}



```

## Gene level {.tabset}

### log2fc
```{r}
if(williamsTrendTest & runBMDonLog2fc){
 doseresponsives_plot(WTToutput = WTT_log2fc, level = "genes") 
}
```

## Network level
```{r}
if(williamsTrendTest & runBMDonNetworkScores){
  doseresponsives_plot(WTToutput = WTT_pathway, level = "modules")
}

```


# Benchmark Dose Analysis


## BMD accumulation plot {.tabset}
Interactive plot to visualize the accumulative benchmark doses. Here, we also filter the BMD output for: 
Best.BMD.BMDL < 20   
Best.BMDU.BMD < 20
Best.RSquared >= 0.8
Best.BMD >= 0.1*lowest concentration <= highest concentration

### Filter BMD data
#### Set function
```{r}
filterBMDoutput = function(BMDoutput){
  
  lowest_conc = 0.1
  highest_conc = 50
    
  BMDoutput %>%
    filter(Best.BMD.BMDL < 20 & Best.BMDU.BMD < 20 & Best.RSquared >= 0.8 & Best.BMD >= 0.1*lowest_conc & Best.BMD <= highest_conc) -> BMDoutputFilter
  
  return(BMDoutputFilter)
}
```

#### Filter data
```{r}

if(runBMDonLog2fc){
  BMD_output_log2fc_filter = filterBMDoutput(BMDoutput = BMD_output_log2fc)
}

if(runBMDonNetworkScores){
  BMD_output_pathway_filter = filterBMDoutput(BMDoutput = BMD_output_pathway)
}

```


```{r}

BMDaccumulationPlot = function(BMDoutput, timepoint_oi, level){
  
  if(level == "gene"){
    hover = "gene: "
  }
  if(level == "pathway"){
    hover = "pathway: "
  }
  
  BMDoutput %>%
    separate(Probe.ID, into=c("gene_symbol", "probe_NR"), sep = "_") %>%
    filter(timepoint == timepoint_oi) %>%
    arrange(Best.BMD) %>%
    ggplot(aes(Best.BMD, y=cumsum(Best.BMD), text = paste0(hover, gene_symbol))) +
    geom_point(stat="identity") +
    scale_x_log10() +
    labs(title = paste0("BMD accumulation plot after ", timepoint_oi),
         y = "Cumulative BMD", 
         x = "BMD")
}



```

### Gene level {.tabset}


#### log2fc
```{r}
if(runBMDonLog2fc){
  BMDaccumulationPlot(BMDoutput = BMD_output_log2fc_filter, timepoint_oi = "4hr", level = "gene") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_log2fc_filter, timepoint_oi = "8hr", level = "gene") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_log2fc_filter, timepoint_oi = "16hr", level = "gene") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_log2fc_filter, timepoint_oi = "24hr", level = "gene") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_log2fc_filter, timepoint_oi = "48hr", level = "gene") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_log2fc_filter, timepoint_oi = "72hr", level = "gene") %>% ggplotly() %>% print()
    
}

```


### Network level
```{r}
if(runBMDonNetworkScores){
  BMDaccumulationPlot(BMDoutput = BMD_output_pathway_filter, timepoint_oi = "4hr", level = "pathway") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_pathway_filter, timepoint_oi = "8hr", level = "pathway") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_pathway_filter, timepoint_oi = "16hr", level = "pathway") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_pathway_filter, timepoint_oi = "24hr", level = "pathway") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_pathway_filter, timepoint_oi = "48hr", level = "pathway") %>% ggplotly() %>% print()
  BMDaccumulationPlot(BMDoutput = BMD_output_pathway_filter, timepoint_oi = "72hr", level = "pathway") %>% ggplotly() %>% print()
  
}

```

## Analyse BMD of most sensitive genes {.tabset}

### Define functions for different methods {.tabset}
#### 5th percentile
Gene closest to the 5th percentile of the BMD distribution; conservative metric value for the tPOD that targets the most responsive genes that may contribute to the toxicological response
```{r}

timepoints = c("4hr", "8hr", "16hr", "24hr", "48hr", "72hr")
  

get_5thpercentile = function(BMDoutput) {
 output_df = list()
   
  for(time in timepoints){
    BMDoutput_timeoi = BMDoutput %>% filter(timepoint == time)
    
  target_BMD = quantile(BMDoutput_timeoi$Best.BMD, 0.05)
  closest_df = BMDoutput_timeoi %>%
    mutate(difference = abs(Best.BMD - target_BMD)) %>%
    arrange(difference) %>%
    head(1)
  closest_gene = closest_df %>% pull(Probe.ID)
  closest_BMD = closest_df %>% pull(Best.BMD)
  output_df[[time]] = data.frame(closest_gene, closest_BMD)
    }  
return(output_df)
  }



if(runBMDonLog2fc){
  percentile_log2fc = get_5thpercentile(BMDoutput = BMD_output_log2fc_filter)
}


if(runBMDonNetworkScores){
  percentile_pathway = get_5thpercentile(BMDoutput = BMD_output_pathway_filter)
}



```


#### 25th lowest ranked gene
The 25th rank ordered gene's BMD isset as theshold, which indicates where a defined change in the transcriptome has occurred
```{r}


get_25thranked = function(BMDoutput) {
  output_df = list()
  
  for (time in timepoints) {
    BMDoutput_timeoi = BMDoutput %>% filter(timepoint == time)
    
    ranked_df = BMDoutput_timeoi %>% arrange(Best.BMD) %>% dplyr::slice(25)
    ranked_gene = ifelse((nrow(ranked_df) == 0), NA, ranked_df %>% pull(Probe.ID))
    ranked_BMD = ifelse((nrow(ranked_df) == 0), NA, ranked_df %>% pull(Best.BMD))
    output_df[[time]] = data.frame(ranked_gene, ranked_BMD)
    
    
  }
  return(output_df)
  
}



if(runBMDonLog2fc){
  ranked_log2fc = get_25thranked(BMDoutput = BMD_output_log2fc_filter)
}


if(runBMDonNetworkScores){
  ranked_pathway = get_25thranked(BMDoutput = BMD_output_pathway_filter)
}



```

#### Lowest Consistent Reponse Dose (LCRD)
The LCRD is identified as the lowest BMC in a rank order of gene BMCs where all subsequent ratio values from adjacently ranked BMCs are within 1.66 (thereby eliminating unrealistically low gene BMCs that may be biological noise). 
```{r}

get_LCRD = function(BMDoutput) {
  output_df = list()
  
  for(time in timepoints){
    BMDoutput_timeoi = BMDoutput %>% filter(timepoint == time)
    
    if(nrow(BMDoutput_timeoi) >= 2){
      # step 1: sort BMD by Best.BMD
      BMD_output_sort <- 
        BMDoutput_timeoi %>% arrange(Best.BMD)
      
      # Inititalize variables
      current_group <- BMD_output_sort$Best.BMD
      largest_crgb <- current_group
      
      # loop through sorted BMDs
      for(i in 2:length(BMD_output_sort$Best.BMD)){
        # calculate ratio
        BMD_ratio = BMD_output_sort$Best.BMD[i]/BMD_output_sort$Best.BMD[i-1]
        
        if(BMD_ratio <= 1.66){
          current_group = c(current_group, BMD_output_sort$Best.BMD[i])
        } else {
          # check if current group is larger than largets found so far
          if(length(current_group) > length(largest_crgb)) {
            largest_crgb <- current_group
          }
          # start new group
          current_group = BMD_output_sort$Best.BMD[i]
        }
        
        # final check to ensure the last group is considered
        if(length(current_group) > length(largest_crgb)) {
          largest_crgb <- current_group
        }
        
        # determine LCRD
        LCRD_BMD <- min(largest_crgb)
        
        # extract genes in largest CRGB
        genes_in_largest_crgb <- 
          BMD_output_sort[BMD_output_sort$Best.BMD %in% largest_crgb, ] %>% pull(Probe.ID)
      }
    } else {
      LCRD_BMD = NA
      genes_in_largest_crgb = NA
    }
    output_df[[time]] = data.frame(CRGB_lowest_gene = genes_in_largest_crgb[1], LCRD_BMD)
  } 
  return(output_df)
}






if(runBMDonLog2fc){
  LCRD_log2fc = get_LCRD(BMDoutput = BMD_output_log2fc_filter)
}


if(runBMDonNetworkScores){
  LCRD_pathway = get_LCRD(BMDoutput = BMD_output_pathway_filter)
}


```



#### First mode
The first mode is defined as the first point at which the first derivative changes from positive to negative with a negative second derivative (second derivative test).

```{r}


get_firstMode = function(BMDoutput) {
  output_df = list()
  
  for (time in timepoints) {
    BMDoutput_timeoi = BMDoutput %>% filter(timepoint == time)
    
    if (nrow(BMDoutput_timeoi) >= 2) {
      density_estimate = density(BMDoutput_timeoi$Best.BMD)
      
      # compute derivatives
      first_derivative = diff(density_estimate$y) / diff(density_estimate$x)
      second_derivative = diff(first_derivative) / diff(density_estimate$x[-1])
      
      # identify first mode
      first_mode_index <- NULL
      
      for (i in 2:(length(first_derivative) - 1)) {
        if (first_derivative[i - 1] > 0 &&
            first_derivative[i] < 0 && second_derivative[i] < 0) {
          first_mode_index <- i
        }
      }
      
      
      if (!is.null(first_mode_index)) {
        first_mode <- density_estimate$x[first_mode_index]
        
        # match BMD and gene (find closest BMD value to first mode)
        closest_BMD_index = which.min(abs(BMDoutput_timeoi$Best.BMD - first_mode))
        closest_bmd = BMDoutput_timeoi$Best.BMD[closest_BMD_index]
        gene_name = BMDoutput_timeoi$Probe.ID[closest_BMD_index]
        
      }
      
    } else {
      closest_bmd = NA
      gene_name = NA
    }
    
    # output BMD and gene name
    output_df[[time]] = data.frame(firstMode_BMD = closest_bmd, firstMode_gene = gene_name)
  }
  return(output_df)
}



if(runBMDonLog2fc){
  firstMode_log2fc = get_firstMode(BMDoutput = BMD_output_log2fc_filter)
}


if(runBMDonNetworkScores){
  firstMode_pathway = get_firstMode(BMDoutput = BMD_output_pathway_filter)
}


```




### Create table {.tabset}
#### Design function
First design function to generate table
```{r}

createBMDtable = function(percentile_list, ranked_list, LCRD_list, firstMode_list) {
  data_matrix <- matrix(NA,
                        nrow = 6,
                        ncol = 8,
                        dimnames = list(
                          paste0(timepoints),
                          c(
                            "5th_percentile_gene",
                            "5th_percentile_BMD",
                            "25th_ranked_gene",
                            "25th_ranked_BMD",
                            "LCRD_gene",
                            "LCRD_BMD",
                            "first_mode_gene",
                            "first_mode_BMD"
                          )
                        ))
  
  # Assign values to the matrix
  # genes 5th percentile
  data_matrix[timepoints[1], "5th_percentile_gene"] <-
    percentile_list[[1]]$closest_gene
  data_matrix[timepoints[2], "5th_percentile_gene"] <-
    percentile_list[[2]]$closest_gene
  data_matrix[timepoints[3], "5th_percentile_gene"] <-
    percentile_list[[3]]$closest_gene
  data_matrix[timepoints[4], "5th_percentile_gene"] <-
    percentile_list[[4]]$closest_gene
  data_matrix[timepoints[5], "5th_percentile_gene"] <-
    percentile_list[[5]]$closest_gene
  data_matrix[timepoints[6], "5th_percentile_gene"] <-
    percentile_list[[6]]$closest_gene
  
  
  # BMD 5th percentile
  data_matrix[timepoints[1], "5th_percentile_BMD"] <-
    percentile_list[[1]]$closest_BMD
  data_matrix[timepoints[2], "5th_percentile_BMD"] <-
    percentile_list[[2]]$closest_BMD
  data_matrix[timepoints[3], "5th_percentile_BMD"] <-
    percentile_list[[3]]$closest_BMD
  data_matrix[timepoints[4], "5th_percentile_BMD"] <-
    percentile_list[[4]]$closest_BMD
  data_matrix[timepoints[5], "5th_percentile_BMD"] <-
    percentile_list[[5]]$closest_BMD
  data_matrix[timepoints[6], "5th_percentile_BMD"] <-
    percentile_list[[6]]$closest_BMD
  
  
  # genes 25th ranked
  data_matrix[timepoints[1], "25th_ranked_gene"] <-
    ranked_list[[1]]$ranked_gene
  data_matrix[timepoints[2], "25th_ranked_gene"] <-
    ranked_list[[2]]$ranked_gene
  data_matrix[timepoints[3], "25th_ranked_gene"] <-
    ranked_list[[3]]$ranked_gene
  data_matrix[timepoints[4], "25th_ranked_gene"] <-
    ranked_list[[4]]$ranked_gene
  data_matrix[timepoints[5], "25th_ranked_gene"] <-
    ranked_list[[5]]$ranked_gene
  data_matrix[timepoints[6], "25th_ranked_gene"] <-
    ranked_list[[6]]$ranked_gene
  
  
  # BMD 25th ranked
  data_matrix[timepoints[1], "25th_ranked_BMD"] <-
    ranked_list[[1]]$ranked_BMD
  data_matrix[timepoints[2], "25th_ranked_BMD"] <-
    ranked_list[[2]]$ranked_BMD
  data_matrix[timepoints[3], "25th_ranked_BMD"] <-
    ranked_list[[3]]$ranked_BMD
  data_matrix[timepoints[4], "25th_ranked_BMD"] <-
    ranked_list[[4]]$ranked_BMD
  data_matrix[timepoints[5], "25th_ranked_BMD"] <-
    ranked_list[[5]]$ranked_BMD
  data_matrix[timepoints[6], "25th_ranked_BMD"] <-
    ranked_list[[6]]$ranked_BMD
  
  
  # genes LCRD
  data_matrix[timepoints[1], "LCRD_gene"] <-
    LCRD_list[[1]]$CRGB_lowest_gene
  data_matrix[timepoints[2], "LCRD_gene"] <-
    LCRD_list[[2]]$CRGB_lowest_gene
  data_matrix[timepoints[3], "LCRD_gene"] <-
    LCRD_list[[3]]$CRGB_lowest_gene
  data_matrix[timepoints[4], "LCRD_gene"] <-
    LCRD_list[[4]]$CRGB_lowest_gene
  data_matrix[timepoints[5], "LCRD_gene"] <-
    LCRD_list[[5]]$CRGB_lowest_gene
  data_matrix[timepoints[6], "LCRD_gene"] <-
    LCRD_list[[6]]$CRGB_lowest_gene
  
  
  # BMD LCRD
  data_matrix[timepoints[1], "LCRD_BMD"] <-
    LCRD_list[[1]]$LCRD_BMD
  data_matrix[timepoints[2], "LCRD_BMD"] <-
    LCRD_list[[2]]$LCRD_BMD
  data_matrix[timepoints[3], "LCRD_BMD"] <-
    LCRD_list[[3]]$LCRD_BMD
  data_matrix[timepoints[4], "LCRD_BMD"] <-
    LCRD_list[[4]]$LCRD_BMD
  data_matrix[timepoints[5], "LCRD_BMD"] <-
    LCRD_list[[5]]$LCRD_BMD
  data_matrix[timepoints[6], "LCRD_BMD"] <-
    LCRD_list[[6]]$LCRD_BMD
  
  
    # genes LCRD
  data_matrix[timepoints[1], "first_mode_gene"] <-
    firstMode_list[[1]]$firstMode_gene
  data_matrix[timepoints[2], "first_mode_gene"] <-
    firstMode_list[[2]]$firstMode_gene
  data_matrix[timepoints[3], "first_mode_gene"] <-
    firstMode_list[[3]]$firstMode_gene
  data_matrix[timepoints[4], "first_mode_gene"] <-
    firstMode_list[[4]]$firstMode_gene
  data_matrix[timepoints[5], "first_mode_gene"] <-
    firstMode_list[[5]]$firstMode_gene
  data_matrix[timepoints[6], "first_mode_gene"] <-
    firstMode_list[[6]]$firstMode_gene
  
  
  # BMD firstMode
  data_matrix[timepoints[1], "first_mode_BMD"] <-
    firstMode_list[[1]]$firstMode_BMD
  data_matrix[timepoints[2], "first_mode_BMD"] <-
    firstMode_list[[2]]$firstMode_BMD
  data_matrix[timepoints[3], "first_mode_BMD"] <-
    firstMode_list[[3]]$firstMode_BMD
  data_matrix[timepoints[4], "first_mode_BMD"] <-
    firstMode_list[[4]]$firstMode_BMD
  data_matrix[timepoints[5], "first_mode_BMD"] <-
    firstMode_list[[5]]$firstMode_BMD
  data_matrix[timepoints[6], "first_mode_BMD"] <-
    firstMode_list[[6]]$firstMode_BMD
  
  
  
  
  
  # Convert the matrix to a data frame
  headers <- as.data.frame(data_matrix)
  
  # Use the kable function to create the table with grouped headers and row names
  library(kableExtra)
  
  kable(
    headers,
    col.names = c("gene", "BMD", "gene", "BMD", "gene", "BMD", "gene", "BMD"),
    row.names = TRUE
  ) %>%
    add_header_above(c(
      " " = 1,
      "5th percentile" = 2,
      "25th ranked gene" = 2,
      "LCDR" = 2,
      "first mode" = 2
    )) %>%
    kable_styling(full_width = TRUE)
  
}


```

#### Fill table {.tabset}

##### Gene level {.tabset}

###### log2fc
```{r}
if(runBMDonLog2fc) {
  createBMDtable(percentile_list = percentile_log2fc, ranked_list = ranked_log2fc, LCRD_list = LCRD_log2fc, firstMode_list = firstMode_log2fc)
}

```


##### Network level
```{r}
if(runBMDonNetworkScores) {
  createBMDtable(percentile_list = percentile_pathway, ranked_list = ranked_pathway, LCRD_list = LCRD_pathway, firstMode_list = firstMode_pathway)
  
}

```




## Analyse BMD of most sensitive networks {.tabset}
Select top 5 networks with the lowest median BMD per timepoint; indicating the most sensitive pathways to compound exposure


### Based on BMD run on pathway scores
#### Print top 5 pathways lowest median BMD per timepoint
Here for example if you run the BMD express on pathway scores (like eigengene scores)

```{r, results = 'asis'}

if(runBMDonNetworkScores) {
  timepoints = BMD_output_pathway_filter %>% pull(timepoint) %>% unique()
  pathways_oi_pertp = NULL
  
  for (i in timepoints) {
    pathways_oi_pertp[[i]] = BMD_output_pathway_filter %>%
      filter(timepoint == i) %>%
      arrange(Best.BMD) %>%
      dplyr::slice(1:5) %>%
      pull(Probe.ID)
    
    if ((pathways_oi_pertp[[i]] %>% length()) == 0) {
      cat(paste0("There are no modules showing a dose response after ", i))
    }
    
    # else{
    #   cat(paste0("The top 5 networks with the lowest BMD after ", i, " are:\n"))
    #   modules_oi_pertp[[i]] %>% print()
    #   cat("\n\n")
    #
    # }
    
  }
  
  DTlowestBMD = do.call("bind_cols", pathways_oi_pertp)
  
  kable(
    DTlowestBMD,
    caption = c("Top 5 most sensitive modules to compound exposure per timepoint")
  )
}


```

