---
title: "DisseminationFigures"
output: html_document
date: "2025-05-07"
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)
```



DISSEMINATION FIGURES
Here, we will generate figure 2 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, here, ggpubr, patchwork, grid, GGally, ggplotify, ComplexUpset)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR, "DisseminationFigures")

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



## Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))
```


# Plots

## BMCs from normalized counts

### Accumulation plots
```{r}
fig2a_AU = BMDaccumulationPlot(BMDoutput = AU_norm_BMD_select, partner = "Aristotle University") +
  theme(plot.title = element_text(color = "firebrick")) +  
  ylim(c(0, 6300))

fig2a_BPI = BMDaccumulationPlot(BMDoutput = BPI_norm_BMD_select, partner = "BPI") +
  theme(plot.title = element_text(color = "gray30")) +
  ylim(c(0, 6300))

fig2a_GU = BMDaccumulationPlot(BMDoutput = GU_norm_BMD_select, partner = "Ghent University") + 
  theme(plot.title = element_text(color = "gold1")) +
  ylim(c(0, 6300))

fig2a_LU = BMDaccumulationPlot(BMDoutput = LU_norm_BMD_select, partner = "Leiden University") +
  theme(plot.title = element_text(color = "blue3")) +
  ylim(c(0, 6300))

fig2a_SC = BMDaccumulationPlot(Sciensano_norm_BMD_select, partner = "Sciensano") + 
  theme(plot.title = element_text(color = "forestgreen")) +
  ylim(c(0, 6300))


# generate title
fig2a_row1_lab <- wrap_elements(panel = textGrob("BMC accumulation (normalized counts)", gp = gpar(fontface = "bold", fontsize = 18)))


fig2a_row1 = wrap_elements(fig2a_AU + fig2a_BPI + fig2a_GU + fig2a_LU + fig2a_SC + plot_layout(nrow = 1, guides = "collect", axis = "collect"))
fig2a_row1_total = wrap_elements(fig2a_row1_lab / fig2a_row1 + plot_layout(heights = c(0.25, 4)))
fig2a_row1_total

ggsave(plot = fig2a_row1_total, file.path(plotDir, "BMCaccumulation_all.pdf"), width = 15, height = 4)
```


### 24 hours
```{r}
fig2b_norm = norm_BMD_combined %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>%
  filter(timepoint == "24hr") %>%
  group_by(partner) %>%
  arrange(finalBMD) %>%
  mutate(ranked_gene = row_number()) %>%
  ggplot(aes(finalBMD, y = ranked_gene)) +
  geom_point(stat = "identity", aes(color = partner), size = 1) +
  labs(y = "Accumulation", 
       x = "BMC") +
  scale_color_manual(values = c("firebrick", "gray30", "gold1", "blue2", "forestgreen"),
                     breaks = c("AU", "BPI", "GU", "LU", "SC"),
                     labels = c("Aristotle University", "BPI", "Ghent University", "Leiden University", "Sciensano")) +
  theme(legend.title = element_blank(),
        legend.position = "bottom")


fig2b_norm


# generate title
fig2b_lab <- wrap_elements(panel = textGrob("BMC accumulation - 24 hours (normalized counts)", gp = gpar(fontface = "bold", fontsize = 22)))


fig2b_total = wrap_elements(fig2b_lab / fig2b_norm + plot_layout(heights = c(0.5, 6)))
fig2b_total

ggsave(plot = fig2b_total, file.path(plotDir, "BMCaccumulation_24hr.pdf"), width = 8.5, height = 10)

```


### BMC correlations
```{r}

fig2c_cor = GGally::ggpairs(norm_BMD_combined_wide, aes(color = timepoint),
                columns = c("AU_finalBMD", "BPI_finalBMD", "GU_finalBMD", "LU_finalBMD", "SC_finalBMD"),
                lower = list(continuous = GGally::wrap("points", size = 0.05)),
                diag = list(continuous = GGally::wrap("densityDiag"), mapping = ggplot2::aes(color=timepoint), alpha = 0.5)
                )


fig2c_cor_gg = wrap_elements(full = ggmatrix_gtable(fig2c_cor)) # convert to ggplot


# generate title
fig2c_lab <- wrap_elements(panel = textGrob("BMD correlations (normalized counts)", gp = gpar(fontface = "bold", fontsize = 22)))


# combine
fig2c = wrap_elements(fig2c_lab / fig2c_cor_gg + plot_layout(heights = c(0.5, 7))) 

ggsave(plot = fig2c, file.path(plotDir, "BMCcorrelations_total.pdf"), width = 13, height = 9)

```




## BMCs pathway of choice

### UpSet plot
Show overlap between p53 pathways

```{r}
KEGG_p53 = AU_pathway_overview %>% filter(ID == "hsa04115") %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id")) %>% pull(gene_symbol) 
WP_p53 = gsea_wikipathways_df %>% filter(pathway == "WP_P53_TRANSCRIPTIONAL_GENE_NETWORK") %>% pull(gene_symbol)
HALLMARK_p53 = GU_pathway_overview %>% filter(pathway == "HALLMARK_P53_PATHWAY") %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id")) %>% pull(gene_symbol) 
module_p53 = LU_pathway_overview_select %>% filter(module == "hRPTECTERT1_35") %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id")) %>% pull(gene_symbol)
GENOMARKTGxDDI = Sciensano_pathway_overview %>% filter(pathway %in% c("TGxDDI64", "GENOMARK84")) %>% pull(gene_symbol)

p53_geneList = list(
  KEGG = KEGG_p53,
  WikiPathways = WP_p53,
  HALLMARK = HALLMARK_p53,
  Modules = module_p53,
  Biomarkers = GENOMARKTGxDDI
)

membership_df = UpSetR::fromList(p53_geneList) %>%
  rownames_to_column("Gene") %>%
  mutate(across(-Gene, as.logical))


fig4a_upset = ComplexUpset::upset(
  membership_df,
  names(p53_geneList),
  min_size = 1,
  matrix = (
    intersection_matrix(geom = geom_point(shape = 'circle filled', size = 3)) +
      scale_color_manual(
        values = c('KEGG' = "firebrick", 'WikiPathways' = "gray30", 'HALLMARK' = "gold1", 'Modules' = "blue2", 'Biomarkers' = "forestgreen"),
      )
    ),
  queries = list(
    upset_query(set = 'KEGG', fill = "firebrick"),
    upset_query(set = "WikiPathways", fill = "gray30"),
    upset_query(set = "HALLMARK", fill = "gold1"),
    upset_query(set = "Modules", fill = "blue2"),
    upset_query(set = "Biomarkers", fill = "forestgreen")
  ))


fig4a_upset

fig4a_lab <- wrap_elements(panel = textGrob("Overlap genes in various p53 pathways", gp = gpar(fontface = "bold", fontsize = 18)))

fig4a = wrap_elements(fig4a_lab / 
                        fig4a_upset +
                        plot_layout(heights = c(0.5, 5))) 


fig4a

ggsave(plot = fig4a, file.path(plotDir, "UpSetPlot_p53Pathways.pdf"), width = 15, height = 5)
```



### Accumulation plot - highlight p53 pathways
```{r}
fig4b_AU = BMDaccumulationPlot_pathway_highlight(BMDoutput = AU_norm_KEGG_select,
                                                 partner = "Aristotle University",
                                                 pathway_oi = "p53 signaling pathway") +
  labs(subtitle = "highlight p53 signaling pathway (KEGG)") + 
  theme(plot.title = element_text(color = "firebrick"),
        plot.subtitle = element_text(color = "red")) + 
  xlim(0,50)



fig4b_BPI = BMDaccumulationPlot_pathway_highlight(BMDoutput = BPI_norm_WIKIPATHWAYS_select,
                                                 partner = "BPI",
                                                 pathway_oi = "p53 transcriptional gene network") + 
  labs(subtitle = "highlight p53 transcriptional gene network (WIKIPATHWAYS)") + 
  theme(plot.title = element_text(color = "gray30"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))



fig4b_GU = BMDaccumulationPlot_pathway_highlight(BMDoutput = GU_norm_HALLMARK_select,
                                                 partner = "Ghent University",
                                                 pathway_oi = "HALLMARK_P53_PATHWAY") +
  labs(subtitle = "highlight p53 pathway (HALLMARK)") +
  theme(plot.title = element_text(color = "gold1"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))



fig4b_LU = BMDaccumulationPlot_pathway_highlight(BMDoutput = LU_norm_modules_select,
                                                 partner = "Leiden University",
                                                 pathway_oi = "hRPTECTERT1_35") + 
  labs(subtitle = "highlight p53 WGCNA module (hRPTECTERT1_35)") +
  theme(plot.title = element_text(color = "blue2"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))


fig4b_SC = BMDaccumulationPlot_pathway_highlight(
  BMDoutput = Sciensano_norm_TXDDI_GENOMARK_select,
  partner = "Sciensano",
  pathway_oi = c("TGxDDI64", "GENOMARK84")) + 
  labs(subtitle = "highlight p53 biomarkers (TGxDDI64 & GENOMARK84)") +
  theme(plot.title = element_text(color = "forestgreen"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))


# generate title
fig4b_lab <- wrap_elements(panel = textGrob("BMC accumulation pathways from preferred collection", gp = gpar(fontface = "bold", fontsize = 22)))


fig4b = wrap_elements(fig4b_AU + fig4b_BPI + fig4b_GU + fig4b_LU + fig4b_SC + plot_layout(nrow = 1, guides = "collect", axis = "collect"))

fig4b_total = wrap_elements(fig4b_lab / fig4b + plot_layout(heights = c(0.4, 5)))
fig4b_total

ggsave(plot = fig4b_total, file.path(plotDir, "BMCaccumulation_pathwayPreferredCollection.pdf"), width = 24, height = 6)
```


### Boxplots gene BMC p53 pathways
```{r}
GENOMARK = Sciensano_pathway_overview %>% filter(pathway == "GENOMARK84") %>% pull(gene_symbol)

fig4c_AU = AU_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% KEGG_p53) %>%
  ggplot(aes(x=timepoint, y=finalBMD)) +
  geom_violin(color = "firebrick") + 
  geom_jitter(color = "firebrick") +
  labs(subtitle = "Genes in p53 signaling pathway (KEGG)",
       y = "BMC") + 
  ylim(c(0, 50)) +
  theme(axis.title.x = element_blank())


fig4c_BPI = BPI_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% WP_p53) %>%
  ggplot(aes(x=timepoint, y = finalBMD)) +
  geom_violin(color = "gray30") +
  geom_jitter(color = "gray30") +
  labs(subtitle = "Genes in p53 transcriptional network (WIKIPATHWAYS)",
       y = "BMC") +
  ylim(c(0, 50)) +
  theme(axis.title.x = element_blank())


fig4c_GU = GU_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% HALLMARK_p53) %>%
  ggplot(aes(x=timepoint, y = finalBMD)) +
  geom_violin(color = "gold1") +
  geom_jitter(color = "gold1") +
  labs(subtitle = "Genes in p53 pathway (HALLMARK)",
       y = "BMC") +
  ylim(c(0, 50)) +
  theme(axis.title.x = element_blank())


fig4c_LU = LU_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% module_p53) %>%
  ggplot(aes(x=timepoint, y = finalBMD)) +
  geom_violin(color = "blue2") +
  geom_jitter(color = "blue2") +
  labs(subtitle = "Genes in p53 WGCNA module (hRPTECTERT1_35)",
       y = "BMC") +
  ylim(c(0, 50)) +
  theme(axis.title.x = element_blank())
  


fig4c_SC = Sciensano_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% GENOMARK) %>%
  ggplot(aes(x=timepoint, y = finalBMD)) +
  geom_violin(color = "forestgreen") +
  geom_jitter(color = "forestgreen") +
  labs(subtitle = "p53 biomarkers (GENOMARK84)",
       y = "BMC") +
  ylim(c(0, 50)) +
  theme(axis.title.x = element_blank())



fig4c = wrap_elements(fig4c_AU + fig4c_BPI + fig4c_GU + fig4c_LU + fig4c_SC + plot_layout(nrow = 1, guides = "collect", axis = "collect"))
fig4c

# generate title
fig4c_lab <- wrap_elements(panel = textGrob("BMC distribution genes in p53 pathway", gp = gpar(fontface = "bold", fontsize = 18)))

fig4c_total = wrap_elements(fig4c_lab / fig4c + plot_layout(heights = c(0.5, 5)))


ggsave(plot = fig4c_total, file.path(plotDir, "BMCdistributionGenesP53Pathways.pdf"), width = 22, height = 6)
```

