---
title: "ManuscriptFigure2"
author: "IBBruns"
date: "2025-04-22"
output: html_document
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)
```



MANUSCRIPT FIGURE 2
Here, we will generate figure 2 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, here, ggpubr, patchwork, grid, GGally, ggplotify)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))
```


# Generate figure


## Figure 2 
Here we will have:   
1. The accumulation plot per partner (all timepoints combined in one plot) to determine the most sensitive timepoint
2. A zoom of the accumulation plot per partner (to see the slight differences in the beginning)
3. A combined accumulationplot of the most sensitive timepoint (24hr)

### Figure 2a

#### Plot first row
```{r}
fig2a_AU = BMDaccumulationPlot(BMDoutput = AU_norm_BMD_select, partner = "Aristotle University") +
  theme(plot.title = element_text(color = "firebrick")) +  
  ylim(c(0, 6300))

fig2a_BPI = BMDaccumulationPlot(BMDoutput = BPI_norm_BMD_select, partner = "BPI") +
  theme(plot.title = element_text(color = "gray30")) +
  ylim(c(0, 6300))

fig2a_GU = BMDaccumulationPlot(BMDoutput = GU_norm_BMD_select, partner = "Ghent University") + 
  theme(plot.title = element_text(color = "gold1")) +
  ylim(c(0, 6300))

fig2a_LU = BMDaccumulationPlot(BMDoutput = LU_norm_BMD_select, partner = "Leiden University") +
  theme(plot.title = element_text(color = "blue3")) +
  ylim(c(0, 6300))

fig2a_SC = BMDaccumulationPlot(Sciensano_norm_BMD_select, partner = "Sciensano") + 
  theme(plot.title = element_text(color = "forestgreen")) +
  ylim(c(0, 6300))


fig2a_row1 = fig2a_AU + fig2a_BPI + fig2a_GU + fig2a_LU + fig2a_SC + plot_layout(nrow = 1, guides = "collect", axis = "collect")
fig2a_row1

ggsave(plot = fig2a_row1, file.path(plotDir, "fig2a_firstrow.pdf"), width = 15, height = 5)

```



#### Zoom
Zoom in on first part (250 genes) plot

#### Plot zoom
```{r}
fig2a_AU_zoom = BMDaccumulationPlot_zoom(BMDoutput = AU_norm_BMD_select, partner = "Aristotle Univserity") + 
  theme(plot.title = element_text(color = "firebrick"))

fig2a_BPI_zoom = BMDaccumulationPlot_zoom(BMDoutput = BPI_norm_BMD_select, partner = "BPI") + 
  theme(plot.title = element_text(color = "gray30"))

fig2a_GU_zoom = BMDaccumulationPlot_zoom(BMDoutput = GU_norm_BMD_select, partner = "Ghent University") + 
  theme(plot.title = element_text(color = "gold1"))


fig2a_LU_zoom = BMDaccumulationPlot_zoom(BMDoutput = LU_norm_BMD_select, partner = "Leiden University") + 
  theme(plot.title = element_text(color = "blue3"))


fig2a_SC_zoom = BMDaccumulationPlot_zoom(Sciensano_norm_BMD_select, partner = "Sciensano") + 
  theme(plot.title = element_text(color = "forestgreen"))


fig2a_row2 = fig2a_AU_zoom + fig2a_BPI_zoom + fig2a_GU_zoom + fig2a_LU_zoom + fig2a_SC_zoom + plot_layout(nrow = 1, guides = "collect", axis = "collect")
fig2a_row2


```


# Generate total figure 2a
```{r}
# generate title
fig2a_row1_lab <- wrap_elements(panel = textGrob("BMC accumulation (normalized counts)", gp = gpar(fontface = "bold", fontsize = 18)))


# combine
fig2a = wrap_elements(fig2a_row1_lab /
  fig2a_row1 /
  fig2a_row2 + plot_layout(guides = "collect", axis = "collect", heights = c(0.5, 2.5, 2.5)))


fig2a
ggsave(plot = fig2a, file.path(plotDir, "fig2a_total.pdf"), width = 15, height = 5.5)


```





### Figure 2b
Select most sensitive timepoint (now 24hr) and combine accumulation plots of partners

```{r}

fig2b_norm = norm_BMD_combined %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>%
  filter(timepoint == "24hr") %>%
  group_by(partner) %>%
  arrange(finalBMD) %>%
  mutate(ranked_gene = row_number()) %>%
  ggplot(aes(finalBMD, y = ranked_gene)) +
  geom_point(stat = "identity", aes(color = partner), size = 1) +
  labs(y = "Accumulation", 
       x = "BMC") +
  scale_color_manual(values = c("firebrick", "gray30", "gold1", "blue2", "forestgreen"),
                     breaks = c("AU", "BPI", "GU", "LU", "SC"),
                     labels = c("Aristotle University", "BPI", "Ghent University", "Leiden University", "Sciensano")) +
  theme(legend.title = element_blank())

fig2b_zoom = fig2b_norm +
  ylim(c(0, 250)) +
  xlim(0, 3) 



# generate title
fig2b_lab <- wrap_elements(panel = textGrob("BMC accumulation - 24 hours (normalized counts)", gp = gpar(fontface = "bold", fontsize = 18)))



# combine
fig2b = wrap_elements(fig2b_lab / fig2b_norm / fig2b_zoom + plot_layout(guides = "collect", axis = "collect", heights = c(0.5, 5, 5)))
fig2b


# save figure 2b
ggsave(plot = fig2b, file.path(plotDir, "fig2b_total.pdf"), width = 5, height = 10.5)

ggsave(plot = fig2b_norm, file.path(plotDir, "fig2b_24hours.pdf"), width=5, height = 3)

```




### Figure 2c
BMD correlations

```{r}
fig2c_cor = GGally::ggpairs(norm_BMD_combined_wide, aes(color = timepoint),
                columns = c("AU_finalBMD", "BPI_finalBMD", "GU_finalBMD", "LU_finalBMD", "SC_finalBMD"),
                lower = list(continuous = GGally::wrap("points", size = 0.05)),
                diag = list(continuous = GGally::wrap("densityDiag"), mapping = ggplot2::aes(color=timepoint), alpha = 0.5)
                )


fig2c_cor_gg = wrap_elements(full = ggmatrix_gtable(fig2c_cor)) # convert to ggplot


# generate title
fig2c_lab <- wrap_elements(panel = textGrob("BMD correlations (normalized counts)", gp = gpar(fontface = "bold", fontsize = 18)))


# combine
fig2c = wrap_elements(fig2c_lab / fig2c_cor_gg + plot_layout(heights = c(0.5, 10))) 

ggsave(plot = fig2c, file.path(plotDir, "fig2c.pdf"), width = 10, height = 10.5)

```


## Create total figure 2
```{r}

figure2 = fig2a /
  (fig2b + fig2c) + 
  plot_layout(guides = "collect", axis = "collect", heights = c(5.5,5.5)) + plot_annotation(tag_levels = "A")

figure2

ggsave(plot = figure2, file.path(plotDir, "Figure2.pdf"), width = 15, height = 16)


```




