---
title: "ManuscriptFigure2"
author: "IBBruns"
date: "2025-04-22"
output: html_document
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)

# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue 
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)

```



MANUSCRIPT FIGURE 2
Here, we will generate figure 2 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, here, ggpubr, patchwork, grid, GGally, ggplotify)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))


tpod_original = data.table::fread(file.path(outputDir, "tpod_calculations_alltimepoints.txt")) %>%
  mutate(time_num = as.numeric(gsub("h", "", timepoint)),
         timepoint = factor(timepoint, levels = c("4h", "8h", "16h", "24h", "48h", "72h")),
         method = factor(method, levels = c("5th percentile", "25th ranked gene", "LCRD", "First mode")))


tpod_bootstrapping = data.table::fread(file.path(outputDir, "tpod_bootstrapping_ci_median.txt"))


DRGs = readxl::read_xlsx(file.path(inputDir, "DRG_PMFG_pertp.xlsx")) %>%
  mutate(timepoint = factor(timepoint, levels = c("4h", "8h", "16h", "24h", "48h", "72h")))


```


# Generate figure


## Figure 2 
Here we will have:   
1. The accumulation plot per partner (all timepoints combined in one plot) to determine the most sensitive timepoint
2. A zoom of the accumulation plot per partner (to see the slight differences in the beginning)
3. A combined accumulationplot of the most sensitive timepoint (24h)

### Figure 2a

#### Plot first row
```{r}
fig2a_AU = BMDaccumulationPlot(BMDoutput = AU_norm_BMD_select, partner = "AristotleU") +
  theme(plot.title = element_text(color = "firebrick")) +  
  ylim(c(0, 6300))

fig2a_BPI = BMDaccumulationPlot(BMDoutput = BPI_norm_BMD_select, partner = "BPI") +
  theme(plot.title = element_text(color = "gray30")) +
  ylim(c(0, 6300))

fig2a_GU = BMDaccumulationPlot(BMDoutput = GU_norm_BMD_select, partner = "GhentU") + 
  theme(plot.title = element_text(color = "gold1")) +
  ylim(c(0, 6300))

fig2a_LU = BMDaccumulationPlot(BMDoutput = LU_norm_BMD_select, partner = "LeidenU") +
  theme(plot.title = element_text(color = "blue3")) +
  ylim(c(0, 6300))

fig2a_SC = BMDaccumulationPlot(Sciensano_norm_BMD_select, partner = "Sciensano") + 
  theme(plot.title = element_text(color = "forestgreen")) +
  ylim(c(0, 6300))


fig2a_row1 = fig2a_AU + fig2a_BPI + fig2a_GU + fig2a_LU + fig2a_SC + plot_layout(nrow = 1, guides = "collect", axis = "collect")
fig2a_row1

ggsave(plot = fig2a_row1, file.path(plotDir, "fig2a_firstrow.pdf"), width = 15, height = 5)

```



#### Zoom
Zoom in on first part (250 genes) plot

#### Plot zoom
```{r}
fig2a_AU_zoom = BMDaccumulationPlot_zoom(BMDoutput = AU_norm_BMD_select, partner = "AristotleU") + 
  theme(plot.title = element_text(color = "firebrick"))

fig2a_BPI_zoom = BMDaccumulationPlot_zoom(BMDoutput = BPI_norm_BMD_select, partner = "BPI") + 
  theme(plot.title = element_text(color = "gray30"))

fig2a_GU_zoom = BMDaccumulationPlot_zoom(BMDoutput = GU_norm_BMD_select, partner = "GhentU") + 
  theme(plot.title = element_text(color = "gold1"))


fig2a_LU_zoom = BMDaccumulationPlot_zoom(BMDoutput = LU_norm_BMD_select, partner = "LeidenU") + 
  theme(plot.title = element_text(color = "blue3"))


fig2a_SC_zoom = BMDaccumulationPlot_zoom(Sciensano_norm_BMD_select, partner = "Sciensano") + 
  theme(plot.title = element_text(color = "forestgreen"))


fig2a_row2 = fig2a_AU_zoom + fig2a_BPI_zoom + fig2a_GU_zoom + fig2a_LU_zoom + fig2a_SC_zoom + plot_layout(nrow = 1, guides = "collect", axis = "collect")
fig2a_row2


```


# Generate total figure 2a
```{r}
# generate title
fig2a_row1_lab <- wrap_elements(panel = textGrob("BMC accumulation", gp = gpar(fontface = "bold", fontsize = 18)))


# combine
fig2a = wrap_elements(fig2a_row1_lab /
  fig2a_row1 /
  fig2a_row2 + plot_layout(guides = "collect", axis = "collect", heights = c(0.5, 2.5, 2.5)))


fig2a
ggsave(plot = fig2a, file.path(plotDir, "fig2a_total.pdf"), width = 15, height = 5.5)


```





### Figure 2b
Select most sensitive timepoint (now 24h) and combine accumulation plots of partners

```{r}

fig2b_norm = norm_BMD_combined %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>%
  filter(timepoint == "24h") %>%
  group_by(partner) %>%
  arrange(finalBMD) %>%
  mutate(ranked_gene = row_number()) %>%
  ggplot(aes(finalBMD, y = ranked_gene)) +
  geom_point(stat = "identity", aes(color = partner), size = 1) +
  labs(y = "Accumulation", 
       x = "BMC") +
  scale_color_manual(values = c("firebrick", "gray30", "gold1", "blue2", "forestgreen"),
                     breaks = c("AU", "BPI", "GU", "LU", "SC"),
                     labels = c("AristotleU", "BPI", "GhentU", "LeidenU", "Sciensano")) +
  theme(legend.title = element_blank())

fig2b_zoom = fig2b_norm +
  ylim(c(0, 250)) +
  xlim(0, 3) 



# generate title
fig2b_lab <- wrap_elements(panel = textGrob("BMC accumulation - 24h", gp = gpar(fontface = "bold", fontsize = 18)))



# combine
fig2b = wrap_elements(fig2b_lab / fig2b_norm / fig2b_zoom + plot_layout(guides = "collect", axis = "collect", heights = c(0.5, 5, 5)))
fig2b


# save figure 2b
ggsave(plot = fig2b, file.path(plotDir, "fig2b_total.pdf"), width = 5, height = 10.5)

ggsave(plot = fig2b_norm, file.path(plotDir, "fig2b_24hours.pdf"), width=5, height = 3)

```




### Figure 2c
BMD correlations

```{r}

fig2c_cor <- GGally::ggpairs(
  norm_BMD_combined_wide,
  aes(color = timepoint),
  columns = c("AU_finalBMD", "BPI_finalBMD", "GU_finalBMD", "LU_finalBMD", "SC_finalBMD"),
  columnLabels = c("AristotleU", "BPI", "GhentU", "LeidenU", "Sciensano"),
  lower = list(continuous = GGally::wrap("points", size = 0.05)),
  diag  = list(
    continuous = GGally::wrap("densityDiag", alpha = 0.5),
    mapping    = ggplot2::aes(color = timepoint)
    # continuous   = GGally::wrap("blankDiag")
  )
) +
  theme_bw() +
  scale_color_manual(values = timepoint_cols) +
  scale_fill_manual(values = timepoint_cols) +  # for density fill, if used
  # if densityDiag uses color only, this is harmless but keeps things consistent
  theme(strip.background = element_rect(fill = "white", colour = "black"),
        strip.text = element_text(face = "bold", size = 13))

# convert to ggplot
fig2c_cor_gg <- wrap_elements(full = ggmatrix_gtable(fig2c_cor))

# title
fig2c_lab <- wrap_elements(
  panel = textGrob("BMC correlations",
                   gp = gpar(fontface = "bold", fontsize = 18))
)

# combine
fig2c <- wrap_elements(fig2c_lab / fig2c_cor_gg + plot_layout(heights = c(0.5, 10)))

ggsave(plot = fig2c, file.path(plotDir, "fig2c.pdf"), width = 10, height = 10.5)


```


## tPOD calcuations

```{r}
fig2d = ggplot(tpod_original, aes(x = timepoint, y = tpod_orig, color = partner, group = partner)) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ method, nrow = 1) +
  labs(
    x = "",
    y = "tPOD",
    color = "",
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill = "white", colour = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11),
    legend.text = element_text(size = 12)
  ) +
  scale_color_manual(values = c("firebrick","gray30", "gold1", "blue2", "forestgreen"))




# title
fig2d_lab <- wrap_elements(
  panel = textGrob("Comparison of tPOD Estimates Across Partners over Time",
                   gp = gpar(fontface = "bold", fontsize = 18))
)

# combine
fig2d_total <- wrap_elements(fig2d_lab / fig2d + plot_layout(heights = c(0.5, 5)))
fig2d_total

ggsave(plot = fig2d_total, file.path(plotDir, "fig2d_total.pdf"), width = 10, height = 4)


```


## tPOD calculations + bootstrapping results
```{r}
timepoint_order = c("4h", "8h", "16h", "24h", "48h", "72h")
method_order = c("5th percentile", "25th ranked gene", "LCRD", "First mode")


### combine original tpods + bootstrap results
tpod_plot_df <- tpod_bootstrapping %>%
  left_join(tpod_original,
            by = c("partner","timepoint","method")) %>%
  mutate(timepoint = factor(timepoint, levels = timepoint_order),
         method = factor(method, levels = method_order))



### OPTION 1: FACET BY TPOD METHOD
fig2d_bs = ggplot(tpod_plot_df,
       aes(x = timepoint, y = tpod_median,
           ymin = tpod_lower, ymax = tpod_upper,
           color = partner)) +
  geom_pointrange(
    position = position_dodge(width = 0.6)
  ) +
  geom_point(
    aes(y = tpod_orig),
    shape = 21,
    fill = "white",
    size = 2.3,
    stroke = 0.6,
    position = position_dodge(width = 0.6)
  ) +
  geom_line(aes(x=timepoint, y=tpod_orig, group = partner)) +
  facet_wrap(~ method, scales = "free_y", nrow = 1) +
  labs(
    x = "",
    y = "tPOD",
    color = ""
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  ) +
  scale_y_log10() +
  scale_color_manual(values = c("firebrick", "gold1", "gray30", "blue2", "forestgreen"),
                     breaks = c("AristotleU", "GhentU", "BPI", "LeidenU", "Sciensano"))



# title
fig2d_lab <- wrap_elements(
  panel = textGrob("tPOD Estimates and Bootstrap Uncertainty",
                   gp = gpar(fontface = "bold", fontsize = 18))
)

# combine
fig2d_bs_total <- wrap_elements(fig2d_lab / fig2d_bs + plot_layout(heights = c(0.5, 5)))
fig2d_bs_total

ggsave(plot = fig2d_bs_total, file.path(plotDir, "fig2d_bs_total.pdf"), width = 10, height = 4)



```



# Add additional barplot
## Barplot

```{r}
label_steps = c(
  "DRG"  = "Dose Responsive Genes",
  "PMFG" = "Post Model Filter Genes"
)

  
barplot_DRGs = DRGs %>%
  ggplot(aes(x = partner, y = value, fill = value)) +
  geom_bar(stat = "identity") +
  facet_grid(step ~ timepoint,
             labeller = labeller(step = label_steps)) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "white", colour = "black"),
    strip.text.x = element_text(size = 14, face = "bold"),
    strip.text.y = element_text(size = 9, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_blank(),
    legend.position = "none"
  ) +
  scale_fill_gradient(low = "yellow", high = "red")

barplot_DRGs


# generate title
DRGs_lab <- wrap_elements(panel = textGrob("Number of Dose Responsive Genes", gp = gpar(fontface = "bold", fontsize = 18)))


# combine
DRGs_total = wrap_elements(DRGs_lab /
  barplot_DRGs + plot_layout(guides = "collect", axis = "collect", heights = c(0.5, 5)))


DRGs_total

ggsave(plot = DRGs_total, file.path(plotDir, "fig2_barplot_DRG.pdf"), width = 15, height = 8)

```




## Create total figure 2
```{r}

figure2 = DRGs_total /
  fig2a /
  (fig2b + fig2c + plot_layout(widths = c(0.6, 1))) /
  fig2d_bs_total +
  plot_layout(guides = "collect", axis = "collect", heights = c(4,4,6, 3)) + plot_annotation(tag_levels = "A")

figure2

ggsave(plot = figure2, file.path(plotDir, "Figure2_v4.pdf"), width = 17, height = 23)


```




