---
title: "Figure5"
output: html_document
date: "2025-04-25"
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)
```



MANUSCRIPT FIGURE 5
Here, we will generate figure 5 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, patchwork, here, grid)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))

```

# Wrangle data
```{r}
LU_pwScore_medianScore = tibble(
  module = LU_BMD_pathway_select$pathway,
  pwScore_BMD = LU_BMD_pathway_select$finalBMD,
  timepoint = LU_BMD_pathway_select$timepoint
) %>%
  left_join(LU_norm_modules_select %>%
              select("module" = Pathway.Name,
                     medianBMD,
                     timepoint), by = c("module", "timepoint"))

Sciensano_pwScore_medianScore = tibble(
  HALLMARKpw = Sciensano_BMD_pathway_select$pathway,
  pwScore_BMD = Sciensano_BMD_pathway_select$finalBMD,
  timepoint = Sciensano_BMD_pathway_select$timepoint
) %>%
  left_join(Sciensano_norm_HALLMARK_select %>%
              select("HALLMARKpw" = Pathway.Name,
                     medianBMD,
                     timepoint), by = c("HALLMARKpw", "timepoint")) 
```


# Generate figure

## Figure 5a
Show BMD accumulation module EGs and HALLMARK NES scores

```{r}
fig5a_LU = BMDaccumulationPlot_pwscores_highlight(BMDoutput = LU_BMD_pathway_select, partner = "Leiden University", HLpw = "hRPTECTERT1_35") +
  labs(subtitle = "Eigengene scores (co-expression modules)") +
  theme(plot.title = element_text(color = "blue2")) 


fig5a_SC = BMDaccumulationPlot_pwscores_highlight(BMDoutput = Sciensano_BMD_pathway_select, partner = "Sciensano", HLpw = "HALLMARK_P53_PATHWAY") +
  labs(subtitle = "HALLMARK NES scores") + 
  theme(plot.title = element_text(color = "forestgreen"))


# generate title
fig5a_lab <- wrap_elements(panel = textGrob("Obtain BMCs on pathway scores", gp = gpar(fontface = "bold", fontsize = 18)))


fig5a = fig5a_lab / fig5a_LU / fig5a_SC + plot_layout(axis = "collect", heights = c(0.5, 5, 5))
fig5a



ggsave(plot = fig5a, file.path(plotDir, "fig5a.pdf"), width = 7.5, height = 10.5)


```

## Figure 5b
Compare tPODs based on pathway scores vs Median BMCs 

```{r}
# correlate pw score BMDs and median BMDs (modules)
highlight_hRPTECTERT1_35 = LU_pwScore_medianScore %>% filter(module == "hRPTECTERT1_35")

fig5b_LU = LU_pwScore_medianScore %>%
  ggplot() +
  geom_point(aes(x=medianBMD, y=pwScore_BMD)) +
  geom_point(data = highlight_hRPTECTERT1_35, aes(x=medianBMD, y=pwScore_BMD), color = "red", size = 3) +
  geom_abline(lty = 3, color = "red") +
  facet_wrap(~timepoint) +
  labs(x="Median Pathway BMC",
       y="BMC Pathway Scores",
       title = "Leiden University",
       subtitle = "Eigengene scores (co-expression modules)") +
  theme(plot.title = element_text(color = "blue2"))



# correlate pw score BMDs and median BMDs (NES)
highlight_HALLMARKp53 = Sciensano_pwScore_medianScore %>% filter(HALLMARKpw == "HALLMARK_P53_PATHWAY")

fig5b_SC = Sciensano_pwScore_medianScore %>%
  ggplot() +
  geom_point(aes(x=medianBMD, y=pwScore_BMD)) +
  geom_point(data = highlight_HALLMARKp53, aes(x=medianBMD, y=pwScore_BMD), color = "red", size = 3) +
  geom_abline(lty = 3, color = "red") +
  facet_wrap(~timepoint) +
  labs(x="Median Pathway BMC",
       y="BMC Pathway Scores",
       title = "Sciensano",
       subtitle = "HALLMARK NES scores") +
  theme(plot.title = element_text(color = "forestgreen"))


# generate title
fig5b_lab <- wrap_elements(panel = textGrob("Compare BMCs from Pathway Scores with Median BMCs", gp = gpar(fontface = "bold", fontsize = 18)))


fig5b = fig5b_lab / fig5b_LU / fig5b_SC + plot_layout(axis = "collect", guides = "collect", heights = c(0.5, 5, 5))
fig5b

ggsave(plot = fig5b, file.path(plotDir, "fig5b.pdf"), width = 7.5, height = 10.5)




 highlight_data = plot_data %>% filter(Pathway.Name %in% pathway_oi)

    ggplot(plot_data, aes(x= medianBMD, y = ranked_pathway)) +
    geom_point(stat = "identity", aes(color = timepoint, shape = timepoint), size = 1.5) +
    geom_point(data = highlight_data, aes(x = medianBMD, y = ranked_pathway, shape = timepoint), color = "red", size = 3) +
    # geom_text(data = highlight_data, aes(x = medianBMD, y = ranked_pathway, label = Pathway.Name), vjust = -1, hjust = 1, color = "red") +
    labs(title = partner,
         y = "Accumulation",
         x = "BMC") +
    scale_color_manual(values = c("#E6E6E6", "#D1D1D1","#BBBBBB","#A0A0A0","#7F7F7F","#4D4D4D"))
    
    
    

```



## Figure 5c
Violin plots to show the distribution of the BMCs

```{r}


fig5c_LU = LU_pwScore_medianScore %>%
  pivot_longer(cols = -c(module, timepoint), names_to = "pwScore_median", values_to = "BMC") %>% 
  ggplot(aes(x=pwScore_median, y=BMC, color = pwScore_median)) +
  geom_violin() +
  geom_jitter(size = 0.8) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(title = "Leiden University",
       subtitle = "Eigengene scores (co-expression modules)") +
  theme(plot.title = element_text(color = "blue2"),
        axis.text.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_color_manual(values = c("red", "gray45"),
                     breaks = c("pwScore_BMD", "medianBMD"),
                     labels = c("Pathway Score BMC", "Median Pathway BMC"))



fig5c_SC = Sciensano_pwScore_medianScore %>%
  pivot_longer(cols = -c(HALLMARKpw, timepoint), names_to = "pwScore_median", values_to = "BMC") %>% 
  ggplot(aes(x=pwScore_median, y=BMC, color = pwScore_median)) +
  geom_violin() +
  geom_jitter(size = 0.8) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(title = "Sciensano",
       subtitle = "HALLMARK NES scores") +
  theme(plot.title = element_text(color = "forestgreen"),
       axis.text.x = element_blank(),
       axis.title.x = element_blank()) +
  scale_color_manual(values = c("red", "gray45"),
                     breaks = c("pwScore_BMD", "medianBMD"),
                     labels = c("Pathway Score BMC", "Median Pathway BMC"))


# generate title
fig5c_lab <- wrap_elements(panel = textGrob("Compare distribution BMCs from Pathway Scores & Median BMCs", gp = gpar(fontface = "bold", fontsize = 18)))


fig5c = fig5c_lab / fig5c_LU / fig5c_SC + plot_layout(axis = "collect", guides = "collect", heights = c(0.5, 3, 3))
fig5c

ggsave(plot = fig5c, file.path(plotDir, "fig5c.pdf"), width = 15, height = 6.5)

```



## Generate total figure 5
```{r}
fig5_top = wrap_elements(fig5a) + wrap_elements(fig5b) + plot_layout(widths = c(1.4, 2.1), nrow = 1, ncol = 2)

fig5 = wrap_elements(fig5_top) /
  wrap_elements(fig5c) +
  plot_layout(heights = c(10, 6), ncol = 1)


ggsave(plot = fig5, file.path(plotDir, "Figure5.pdf"), width = 15, height = 16)
```