---
title: "Figure4"
output: html_document
date: "2025-04-25"
---

# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)


# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue (replacing teal)
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)
```



MANUSCRIPT FIGURE 4
Here, we will generate figure 4 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
library(tidyverse)
library(patchwork)
library(here)
library(ComplexUpset)
library(UpSetR)
library(purrr)
library(pheatmap)
library(ggplotlify)
library(grid)

# biomanager
library(ComplexHeatmap)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))

```



# Generate figure

## Figure 4a
UpSet plot to show overlap between different p53 pathways

```{r}

KEGG_p53 = AU_pathway_overview %>% filter(ID == "hsa04115") %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id")) %>% pull(gene_symbol) %>% unique()
WP_p53 = gsea_wikipathways_df %>% filter(pathway == "WP_P53_TRANSCRIPTIONAL_GENE_NETWORK") %>% pull(gene_symbol) %>% unique()
HALLMARK_p53 = GU_pathway_overview %>% filter(pathway == "HALLMARK_P53_PATHWAY") %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id")) %>% pull(gene_symbol) %>% unique()
module_p53 = LU_pathway_overview_select %>% filter(module == "hRPTECTERT1_35") %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id")) %>% pull(gene_symbol) %>% unique()
# GENOMARKTGxDDI = Sciensano_pathway_overview %>% filter(pathway %in% c("TGxDDI64", "GENOMARK84")) %>% pull(gene_symbol)
GENOMARK = Sciensano_pathway_overview %>% filter(pathway %in% "GENOMARK84" & gene_symbol != "") %>% pull(gene_symbol) %>% unique()
TGxDDI = Sciensano_pathway_overview %>% filter(pathway %in% "TGxDDI64" & gene_symbol != "") %>% pull(gene_symbol) %>% unique()

p53_geneList = list(
  KEGG = KEGG_p53,
  WikiPathways = WP_p53,
  HALLMARK = HALLMARK_p53,
  Modules = module_p53,
  # Biomarkers = GENOMARKTGxDDI
  GENOMARK84 = GENOMARK,
  TGxDDI64 = TGxDDI
)


all_genes <- sort(p53_geneList %>% unlist(use.names = FALSE) %>% unique())

membership_df <- tibble::tibble(
  Gene = all_genes,
  !!!lapply(p53_geneList, function(v) all_genes %in% v)
) %>%
  mutate(across(-Gene, as.logical))

fig4a_upset = ComplexUpset::upset(
  membership_df,
  names(p53_geneList),
  min_size = 1,
  matrix = (
    intersection_matrix(geom = geom_point(shape = 'circle filled', size = 3)) +
      scale_color_manual(
        values = c('KEGG' = "firebrick", 'WikiPathways' = "gray30", 'HALLMARK' = "gold1", 'Modules' = "blue2", 'GENOMARK84' = "forestgreen", 'TGxDDI64' = "forestgreen"),
      )
    ),
  queries = list(
    upset_query(set = 'KEGG', fill = "firebrick"),
    upset_query(set = "WikiPathways", fill = "gray30"),
    upset_query(set = "HALLMARK", fill = "gold1"),
    upset_query(set = "Modules", fill = "blue2"),
    upset_query(set = "GENOMARK84", fill = "forestgreen"),
    upset_query(set = "TGxDDI64", fill = "forestgreen")
  ))


fig4a_total <- wrap_elements(
  fig4a_upset +
  plot_annotation(title = "Overlap Genes in P53 Pathways from Different Sources")) &
  theme(
    plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
    legend.position = "bottom"
  )

fig4a_total

ggsave(
  plot = fig4a_total,
  filename = file.path(plotDir, "fig4a.pdf"),
  width = 15,
  height = 5.5
)



```



## Figure 4b
Accumulation plot median BMD of choice, highlight p53 pathway

```{r}

fig4b_LU = BMDaccumulationPlot_pathway_highlight(BMDoutput = LU_norm_modules_select,
                                                 analysis_summary = "BMDE-noWTT-CPM-RF-S5",
                                                 pathway_oi = "hRPTECTERT1_35") + 
  labs(subtitle = "highlight p53 WGCNA module (hRPTECTERT1_35)") +
  theme(plot.title = element_text(color = "blue2"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))




fig4b_BPI = BMDaccumulationPlot_pathway_highlight(BMDoutput = BPI_norm_WIKIPATHWAYS_select,
                                                 analysis_summary = "DRO-Quad-VST-C10-S0",
                                                 pathway_oi = "p53 transcriptional gene network") + 
  labs(subtitle = "highlight p53 transcriptional gene network (WIKIPATHWAYS)") + 
  theme(plot.title = element_text(color = "gray30"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))

fig4b_AU = BMDaccumulationPlot_pathway_highlight(BMDoutput = AU_norm_KEGG_select,
                                                 analysis_summary = "DRO-Quad-UQ-RF-S0",
                                                 pathway_oi = "p53 signaling pathway") +
  labs(subtitle = "highlight p53 signaling pathway (KEGG)") + 
  theme(plot.title = element_text(color = "firebrick"),
        plot.subtitle = element_text(color = "red")) + 
  xlim(0,50)




fig4b_GU = BMDaccumulationPlot_pathway_highlight(BMDoutput = GU_norm_HALLMARK_select,
                                                 analysis_summary = "DRO-Quad-VST-RF-S0",
                                                 pathway_oi = "HALLMARK_P53_PATHWAY") +
  labs(subtitle = "highlight p53 pathway (HALLMARK)") +
  theme(plot.title = element_text(color = "gold1"),
        plot.subtitle = element_text(color = "red")) +
  xlim(c(0,50))





# fig4b_SC = BMDaccumulationPlot_pathway_highlight(
#   BMDoutput = Sciensano_norm_TXDDI_GENOMARK_select,
#   partner = "Sciensano",
#   pathway_oi = c("TGxDDI64", "GENOMARK84")) + 
#   labs(subtitle = "highlight p53 biomarkers (TGxDDI64 & GENOMARK84)") +
#   theme(plot.title = element_text(color = "forestgreen"),
#         plot.subtitle = element_text(color = "red")) +
#   xlim(c(0,50))




fig4b_total <- wrap_elements(
  ( fig4b_LU + fig4b_AU + fig4b_BPI  + fig4b_GU) +
  plot_layout(nrow = 1, guides = "collect", axis = "collect") +
  plot_annotation(title = "Median BMC Accumulation")) &
  theme(
    plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
    legend.position = "bottom"
  )

fig4b_total

ggsave(
  plot = fig4b_total,
  filename = file.path(plotDir, "fig4b.pdf"),
  width = 24,
  height = 5
)
```


## Figure 4c
Overview of BMC genes in p53 pathways of choice + add median BMC

```{r}
fig4c_LU = LU_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% module_p53) %>%
  ggplot(aes(x = timepoint, y = finalBMD)) +
  geom_violin(fill = "blue2", color = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5, color = "blue4") +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +  # median = thick line
  labs(subtitle = "WGCNA modules", y = "BMC") +
  coord_cartesian(ylim = c(0, 50)) +
  theme(axis.title.x = element_blank())
  


fig4c_SC_gm = Sciensano_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% GENOMARK) %>%
  ggplot(aes(x = timepoint, y = finalBMD)) +
  geom_violin(fill = "forestgreen", color = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5, color = "darkgreen") +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +  # median = thick line
  labs(subtitle = "GENOMARK84", y = "BMC") +
  coord_cartesian(ylim = c(0, 50)) +
  theme(axis.title.x = element_blank())


fig4c_SC_td = Sciensano_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% TGxDDI) %>%
  ggplot(aes(x = timepoint, y = finalBMD)) +
  geom_violin(fill = "forestgreen", color = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5, color = "darkgreen") +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +  # median = thick line
  labs(subtitle = "TGxDDI64", y = "BMC") +
  coord_cartesian(ylim = c(0, 50)) +
  theme(axis.title.x = element_blank())


fig4c_AU = AU_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>%
  mutate(ratio = finalBMDU/finalBMDL) %>% filter(ratio <= 40) %>% # add filter to include genes in pathway BMC calculation
  filter(gene_symbol %in% KEGG_p53) %>%
  ggplot(aes(x = timepoint, y = finalBMD)) +
  geom_violin(fill = "firebrick", color = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5, color = "red4") +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +  # median = thick line
  labs(subtitle = "KEGG", y = "BMC") +
  coord_cartesian(ylim = c(0, 50)) +
  theme(axis.title.x = element_blank())



fig4c_BPI = BPI_norm_BMD_select %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% filter(gene_symbol %in% WP_p53) %>%
  ggplot(aes(x = timepoint, y = finalBMD)) +
  geom_violin(fill = "gray30", color = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5, color = "gray1") +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +  # median = thick line
  labs(subtitle = "WikiPathways", y = "BMC") +
  coord_cartesian(ylim = c(0, 50)) +
  theme(axis.title.x = element_blank())




fig4c_GU <- GU_norm_BMD_select %>% 
  separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% 
  filter(gene_symbol %in% HALLMARK_p53) %>%
  ggplot(aes(x = timepoint, y = finalBMD)) +
  geom_violin(fill = "gold1", color = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5, color = "gold4") +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +  # median = thick line
  labs(subtitle = "HALLMARK", y = "BMC") +
  coord_cartesian(ylim = c(0, 50)) +
  theme(axis.title.x = element_blank())




fig4c_total <- wrap_elements(
  (fig4c_LU + fig4c_SC_gm + fig4c_SC_td + fig4c_AU + fig4c_BPI + fig4c_GU) +
  plot_layout(
    nrow   = 1,
    guides = "collect",
    axis   = "collect"
  ) +
  plot_annotation(
    title = "BMC Distribution of Genes in P53 Pathways"
  )) &
  theme(
    plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
    legend.position = "bottom"
  )

fig4c_total

ggsave(
  plot = fig4c_total,
  filename = file.path(plotDir, "fig4c.pdf"),
  width = 24,
  height = 5
)



```



## Figure 4d
Heatmap of genes that overlap in at 3/5 pathways

```{r}

pathway_cols <- colnames(membership_df %>% column_to_rownames("Gene"))

core_genes = membership_df %>%
  mutate(n_sets = rowSums(across(all_of(pathway_cols)))) %>%
  filter(n_sets >= 3) %>%
  pull(Gene)


# build matrix
plotMat <- norm_BMD_combined_wide %>%
  separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>%
  filter(timepoint == "24h") %>%
  filter(gene_symbol %in% core_genes) %>%
  select(-entrez_id, -timepoint) %>%
  # group_by(gene_symbol) %>%
  # summarize(across(where(is.numeric), ~ first(na.omit(.x))), .groups = "drop") %>%
  column_to_rownames("gene_symbol") %>%
  as.matrix()

# heatmap colors
max_val <- max(plotMat, na.rm = TRUE)
col_fun <- circlize::colorRamp2(c(0, 5, max_val), c("#B2182B", "#FEE08B", "#2166AC"))

# partner annotation
sets <- c("WikiPathways", "HALLMARK", "KEGG", "Modules", "GENOMARK84", "TGxDDI64")

# logical membership per set for each gene (row)
m_annot <- sapply(p53_geneList, function(v) rownames(plotMat) %in% v)
m_annot <- m_annot[, sets, drop = F]                      # dataframe for rowAnnotation
# (optional) keep this column order explicitly:
m_annot <- m_annot[, sets, drop = FALSE]

# colors for the annotation (TRUE colored; FALSE light grey)
set_cols  <- c(KEGG="firebrick", WikiPathways="gray30", HALLMARK="gold1",
               Modules="blue2", GENOMARK84="forestgreen", TGxDDI64="forestgreen")
anno_cols <- lapply(set_cols, function(clr) c("FALSE"="grey90","TRUE"=clr))

row_anno <- rowAnnotation(
  df  = m_annot,
  col = anno_cols,
  annotation_name_side = "top",
  width = unit(12, "mm")    # tweak to taste
)


# column labels
raw_cols <- colnames(plotMat)
friendly_labels <- raw_cols %>%
  sub("_finalBMD$", "", .) %>%
  dplyr::recode(AU = "DRO-Quad-UQ-RF-S0",
                GU = "DRO-Quad-VST-RF-S0",
                LU = "BMDE-noWTT-CPM-RF-S5",
                SC = "BMDE-WTT-CPM-RF-S0",
                BPI = "DRO-Quad-VST-C10-S0")

partner_cols = c("DRO-Quad-VST-C10-S0" = "gray30", 
                "DRO-Quad-VST-RF-S0" = "gold1",
                "BMDE-noWTT-CPM-RF-S5" = "blue2",
                "DRO-Quad-UQ-RF-S0" = "firebrick",
                "BMDE-WTT-CPM-RF-S0" = "forestgreen")

colnames(plotMat) <- friendly_labels

# --- compose heatmap + annotation (no top annotation) ---
ht <- row_anno + Heatmap(
  plotMat,
  name = "BMC",
  col = col_fun,
  na_col = "grey90",
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  # color column labels to match the partner mapping
  column_names_gp = gpar(col = partner_cols[friendly_labels]),
  heatmap_legend_param = list(
    at = c(0, 5, max_val),
    labels = c("0", "5", round(max_val, 1))
  ),
  cell_fun = function(j, i, x, y, w, h, fill) {
    val <- plotMat[i, j]
    if (!is.na(val)) grid::grid.text(sprintf("%.1f", val), x, y, gp = gpar(fontsize = 14))
  }
)



# Draw directly inside the PDF device (recommended)
pdf(file.path(plotDir, "fig4d_heatmap.pdf"), width = 12, height = 9)
draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()



# Grab ComplexHeatmap as grob
ht_grob <- grid::grid.grabExpr(
  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )
)

# Convert grob to patchwork-compatible object
fig4d_heatmap <- wrap_elements(ht_grob)

# Build final figure
fig4d_total <- wrap_elements(
  fig4d_heatmap +
  plot_annotation(
    title = "BMC Genes Overlapping in 3 or more P53 Pathways",
    subtitle = "BMC at 24h"
  )) &
  theme(
    plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 22, hjust = 0.5)
  )

fig4d_total

ggsave(
  plot = fig4d_total,
  filename = file.path(plotDir, "fig4d.pdf"),
  width = 24,
  height = 10
)



```



# Arrange total figure 4
```{r, fig.width=19, fig.height=24}
figure4 = fig4a_total /
  fig4b_total /
  fig4c_total /
  fig4d_total +
  plot_layout(heights = c(7.5, 5, 4, 12)) +
  plot_annotation(tag_levels = "A")

figure4

ggsave(plot = figure4, file.path(plotDir, "Figure4.pdf"), width = 19, height = 24)

```

