---
title: "Figure5"
output: html_document
date: "2025-04-25"
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)

# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue (replacing teal)
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)

```



MANUSCRIPT FIGURE 5
Here, we will generate figure 5 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, patchwork, here, grid)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))

rptec_modules = read_rds(file.path(dirname(inputDir), "RPTECmodules", "module_definition.rds"))
```

# Wrangle data
```{r}

# calcuate nr genes per pathway
LU_modules_genetotals <- rptec_modules %>%
  distinct(module, entrez_id) %>% 
  count(module, name = "n_genes_total")


SC_pathway_gene_totals <- Sciensano_pathway_overview %>%
  filter(str_starts(pathway, "HALLMARK_")) %>% 
  dplyr::rename("HALLMARKpw" = pathway) %>%
  distinct(HALLMARKpw, gene_symbol_entrez_id) %>% 
  count(HALLMARKpw, name = "n_genes_total")



# add percentage overlap to df
LU_pwScore_medianScore = tibble(
  module = LU_BMD_pathway_select$pathway,
  pwScore_BMD = LU_BMD_pathway_select$finalBMD,
  timepoint = LU_BMD_pathway_select$timepoint
) %>%
  left_join(LU_norm_modules_select %>%
              select("module" = Pathway.Name,
                     medianBMD,
                     timepoint,
                     n_genes_used), by = c("module", "timepoint")) %>% 
  left_join(LU_modules_genetotals, by = "module") %>%
  mutate(pct_overlap = 100*n_genes_used / n_genes_total)





Sciensano_pwScore_medianScore = tibble(
  HALLMARKpw = Sciensano_BMD_pathway_select$pathway,
  pwScore_BMD = Sciensano_BMD_pathway_select$finalBMD,
  timepoint = Sciensano_BMD_pathway_select$timepoint
) %>%
  left_join(Sciensano_norm_HALLMARK_select %>%
              select("HALLMARKpw" = Pathway.Name,
                     medianBMD,
                     timepoint,
                     n_genes_used), by = c("HALLMARKpw", "timepoint")) %>%
  left_join(SC_pathway_gene_totals, by = "HALLMARKpw") %>%
  mutate(pct_overlap = 100*n_genes_used / n_genes_total)


```


# Generate figure

## Figure 5a
Show BMD accumulation module EGs and HALLMARK NES scores

```{r}
fig5a_LU = BMDaccumulationPlot_pwscores_highlight(BMDoutput = LU_BMD_pathway_select, partner = "LeidenU", HLpw = "hRPTECTERT1_35") +
  labs(subtitle = "Eigengene scores (co-expression modules)") +
  theme(plot.title = element_text(color = "blue2")) 


fig5a_SC = BMDaccumulationPlot_pwscores_highlight(BMDoutput = Sciensano_BMD_pathway_select, partner = "Sciensano", HLpw = "HALLMARK_P53_PATHWAY") +
  labs(subtitle = "HALLMARK NES scores") + 
  theme(plot.title = element_text(color = "forestgreen"))


# generate title
fig5a_lab <- wrap_elements(panel = textGrob("BMC Accumulation Pathway Scores", gp = gpar(fontface = "bold", fontsize = 18)))


fig5a_total = wrap_elements(fig5a_lab / fig5a_LU / fig5a_SC + plot_layout(axis = "collect", heights = c(0.5, 5, 5)))
fig5a_total



ggsave(plot = fig5a_total, file.path(plotDir, "fig5a.pdf"), width = 7.5, height = 10.5)


```

## Figure 5b
Compare tPODs based on pathway scores vs Median BMCs 

```{r}
# # correlate pw score BMDs and median BMDs (modules)
# highlight_hRPTECTERT1_35 = LU_pwScore_medianScore %>% filter(module == "hRPTECTERT1_35")
# 
# fig5b_LU = LU_pwScore_medianScore %>%
#   ggplot() +
#   geom_point(aes(x=medianBMD, y=pwScore_BMD)) +
#   geom_point(data = highlight_hRPTECTERT1_35, aes(x=medianBMD, y=pwScore_BMD), color = "red", size = 3) +
#   geom_abline(lty = 3, color = "red") +
#   facet_wrap(~timepoint) +
#   labs(x="Median Pathway BMC",
#        y="BMC Pathway Scores",
#        title = "Leiden University",
#        subtitle = "Eigengene scores (co-expression modules)") +
#   theme(plot.title = element_text(color = "blue2"))
# 
# 
# 
# # correlate pw score BMDs and median BMDs (NES)
# highlight_HALLMARKp53 = Sciensano_pwScore_medianScore %>% filter(HALLMARKpw == "HALLMARK_P53_PATHWAY")
# 
# fig5b_SC = Sciensano_pwScore_medianScore %>%
#   ggplot() +
#   geom_point(aes(x=medianBMD, y=pwScore_BMD)) +
#   geom_point(data = highlight_HALLMARKp53, aes(x=medianBMD, y=pwScore_BMD), color = "red", size = 3) +
#   geom_abline(lty = 3, color = "red") +
#   facet_wrap(~timepoint) +
#   labs(x="Median Pathway BMC",
#        y="BMC Pathway Scores",
#        title = "Sciensano",
#        subtitle = "HALLMARK NES scores") +
#   theme(plot.title = element_text(color = "forestgreen"))
# 
# 
# # generate title
# fig5b_lab <- wrap_elements(panel = textGrob("Compare BMCs from Pathway Scores with Median BMCs", gp = gpar(fontface = "bold", fontsize = 18)))
# 
# 
# fig5b_total = fig5b_lab / fig5b_LU / fig5b_SC + plot_layout(axis = "collect", guides = "collect", heights = c(0.5, 5, 5))
# fig5b_total
# 
# ggsave(plot = fig5b_total, file.path(plotDir, "fig5b.pdf"), width = 7.5, height = 10.5)
#     

```



## Figure 5c
Violin plots to show the distribution of the BMCs > change

```{r}


# fig5c_LU = LU_pwScore_medianScore %>%
#   pivot_longer(cols = -c(module, timepoint), names_to = "pwScore_median", values_to = "BMC") %>% 
#   ggplot(aes(x=pwScore_median, y=BMC, color = pwScore_median)) +
#   geom_violin() +
#   geom_jitter(size = 0.8) +
#   facet_wrap(~timepoint, nrow = 1) +
#   labs(title = "Leiden University",
#        subtitle = "Eigengene scores (co-expression modules)") +
#   theme(plot.title = element_text(color = "blue2"),
#         axis.text.x = element_blank(),
#         axis.title.x = element_blank()) +
#   scale_color_manual(values = c("red", "gray45"),
#                      breaks = c("pwScore_BMD", "medianBMD"),
#                      labels = c("Pathway Score BMC", "Median Pathway BMC"))
# 
# 
# 
# fig5c_SC = Sciensano_pwScore_medianScore %>%
#   pivot_longer(cols = -c(HALLMARKpw, timepoint), names_to = "pwScore_median", values_to = "BMC") %>% 
#   ggplot(aes(x=pwScore_median, y=BMC, color = pwScore_median)) +
#   geom_violin() +
#   geom_jitter(size = 0.8) +
#   facet_wrap(~timepoint, nrow = 1) +
#   labs(title = "Sciensano",
#        subtitle = "HALLMARK NES scores") +
#   theme(plot.title = element_text(color = "forestgreen"),
#        axis.text.x = element_blank(),
#        axis.title.x = element_blank()) +
#   scale_color_manual(values = c("red", "gray45"),
#                      breaks = c("pwScore_BMD", "medianBMD"),
#                      labels = c("Pathway Score BMC", "Median Pathway BMC"))
# 
# 
# # generate title
# fig5c_lab <- wrap_elements(panel = textGrob("Compare distribution BMCs from Pathway Scores & Median BMCs", gp = gpar(fontface = "bold", fontsize = 18)))
# 
# 
# fig5c = fig5c_lab / fig5c_LU / fig5c_SC + plot_layout(axis = "collect", guides = "collect", heights = c(0.5, 3, 3))
# fig5c
# 
# ggsave(plot = fig5c, file.path(plotDir, "fig5c.pdf"), width = 15, height = 6.5)

```


## Figure 5b
Generate Bland-Altman plot
```{r}

plot_bland_altman <- function(df,
                              score_col        = "pwScore_BMD",
                              median_col       = "medianBMD",
                              time_col         = "timepoint",
                              partner,
                              pathwaySource,
                              pathway_col      = "pathway",
                              highlight_pathway = NULL) {
  
  # 1) Define all timepoints (to force empty facets)
  all_tps <- c("4h", "8h", "16h", "24h", "48h", "72h")
  df[[time_col]] <- factor(df[[time_col]], levels = all_tps)
  
  # 2) Prepare data
  ba_df <- df %>%
    dplyr::filter(!is.na(.data[[score_col]]),
                  !is.na(.data[[median_col]])) %>%
    dplyr::mutate(
      mean_bmd = (.data[[score_col]] + .data[[median_col]]) / 2,
      diff_bmd = .data[[score_col]] - .data[[median_col]],
      is_highlight = ifelse(.data[[pathway_col]] == highlight_pathway, TRUE, FALSE)
    )
  
  # 3) Calculate stats per timepoint
  stats_df <- ba_df %>%
    dplyr::group_by(.data[[time_col]]) %>%
    dplyr::summarise(
      mean_diff = mean(diff_bmd, na.rm = TRUE),
      sd_diff   = sd(diff_bmd, na.rm = TRUE),
      upper     = mean_diff + 1.96 * sd_diff,
      lower     = mean_diff - 1.96 * sd_diff,
      .groups   = "drop"
    )
  
  # 4) Join stats
  ba_df <- ba_df %>% dplyr::left_join(stats_df, by = time_col)
  
  # Split normal and highlighted data
  df_normal <- ba_df %>% dplyr::filter(!is_highlight)
  df_high   <- ba_df %>% dplyr::filter(is_highlight)
  
  # 5) Plot
  ggplot() +
    # regular points
    geom_point(
      data = df_normal,
      aes(x = mean_bmd, y = diff_bmd),
      size = 1.8,
      color = "black"
    ) +
    # highlight overlay point (red + bigger)
    geom_point(
      data = df_high,
      aes(x = mean_bmd, y = diff_bmd),
      size = 4,
      color = "red"
    ) +
    # BA lines
    geom_hline(aes(yintercept = mean_diff), data = stats_df,
               color = "blue", linewidth = 1, na.rm = TRUE) +
    geom_hline(aes(yintercept = upper), data = stats_df,
               color = "red", linetype = "dashed", na.rm = TRUE) +
    geom_hline(aes(yintercept = lower), data = stats_df,
               color = "red", linetype = "dashed", na.rm = TRUE) +
    
    # facets
    facet_wrap(
      as.formula(paste("~", time_col)),
      scales = "free",
      nrow = 2,
      drop = FALSE
    ) +
    
    labs(
      x = "Mean BMC (BMC Pathway Score & Median Pathway BMC)",
      y = "Difference (BMC Pathway Score − Median Pathway BMC)",
      title = partner,
      subtitle = pathwaySource
    ) 
}



# plot LU
LU_BAplot = plot_bland_altman(LU_pwScore_medianScore %>% dplyr::rename("pathway" = module), partner = "LeidenU", pathwaySource = "WGCNA Modules", highlight_pathway = "hRPTECTERT1_35") +
  theme(plot.title = element_text(color = "blue2"))

SC_BAplot = plot_bland_altman(Sciensano_pwScore_medianScore %>% dplyr::rename("pathway" = HALLMARKpw), partner = "Sciensano", pathwaySource = "HALLMARK Pathways", highlight_pathway = "HALLMARK_P53_PATHWAY") +
  theme(plot.title = element_text(color = "forestgreen"))


fig5b = LU_BAplot + SC_BAplot + plot_layout(nrow = 2, axis_titles = "collect")
fig5b


# generate title
fig5b_lab <- wrap_elements(panel = textGrob("Bland–Altman Plot Across Timepoints", gp = gpar(fontface = "bold", fontsize = 18)))


fig5b_total = wrap_elements(fig5b_lab / fig5b + plot_layout(heights = c(0.5, 7)))
fig5b_total

ggsave(plot = fig5b_total, file.path(plotDir, "fig5b.pdf"), width = 7.5, height = 10.5)



```



## Figure 5c
BMD distribution plot

```{r}
LU_pwScore_medianScore %>%
  ggplot() +
  geom_density(aes(x=medianBMD), alpha = 0.8) +
  geom_density(aes(x=pwScore_BMD), alpha = 0.8) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(title = "LeidenU",
       subtitle = "WGCNA Modules",
       x = "BMC") +
  scale_fill_manual(values = c("lightgreen", "orchid1"),
                    breaks = c("medianBMD", "pwScore_BMD"),
                    labels = c("Median Pathway BMC", "BMC Pathway Score"))



LU_BMC_distr = LU_pwScore_medianScore %>%
  pivot_longer(cols = c(medianBMD, pwScore_BMD),
               names_to = "type",
               values_to = "BMD") %>%
  ggplot(aes(x = BMD, fill = type)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~timepoint, nrow = 1) +
  scale_fill_manual(values = c("lightgreen", "orchid1"),
                    breaks = c("medianBMD", "pwScore_BMD"),
                    labels = c("Median Pathway BMC", "BMC Pathway Score")) +
  labs(title = "LeidenU",
       subtitle = "WGCNA Modules",
       x = "BMC") +
  theme(legend.title = element_blank(),
        plot.title = element_text(color = "blue2"))


SC_BMC_distr = Sciensano_pwScore_medianScore %>%
  pivot_longer(cols = c(medianBMD, pwScore_BMD),
               names_to = "type",
               values_to = "BMD") %>%
  ggplot(aes(x = BMD, fill = type)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~timepoint, nrow = 1) +
  scale_fill_manual(values = c("lightgreen", "orchid1"),
                    breaks = c("medianBMD", "pwScore_BMD"),
                    labels = c("Median Pathway BMC", "BMC Pathway Score")) +
  labs(title = "Sciensano",
       subtitle = "HALLMARK Pathways",
       x = "BMC") +
  theme(legend.title = element_blank(),
        plot.title = element_text(color = "forestgreen"))



fig5c = wrap_elements(LU_BMC_distr / SC_BMC_distr + plot_layout(axis = "collect", guides = "collect"))


# generate title
fig5c_lab <- wrap_elements(panel = textGrob("BMC Distribution - Median Pathway BMC vs BMC Pathway Score", gp = gpar(fontface = "bold", fontsize = 18)))


fig5c_total = wrap_elements(fig5c_lab / fig5c + plot_layout(heights = c(0.5, 7)))
fig5c_total

ggsave(plot = fig5c_total, file.path(plotDir, "fig5c.pdf"), width = 15, height = 8)

``` 

## Figure 5d
Compare pathway size & variation

```{r}


plot_bmc_diff_vs_pct_overlap <- function(df,
                                         partner, 
                                         pathwaySource,
                                         score_col   = "pwScore_BMD",
                                         median_col  = "medianBMD",
                                         pct_col     = "pct_overlap",
                                         time_col    = "timepoint",
                                         log_x       = FALSE,
                                         log_y       = FALSE) {
  
  # 1) Explicitly define all timepoints
  all_tps <- c("4h", "8h", "16h", "24h", "48h", "72h")
  df[[time_col]] <- factor(df[[time_col]], levels = all_tps)
  
  # 2) Filter and compute differences
  df_plot <- df %>%
    filter(!is.na(.data[[score_col]]),
           !is.na(.data[[median_col]]),
           !is.na(.data[[pct_col]]),
           is.finite(.data[[pct_col]])) %>%
    mutate(
      diff_signed = .data[[score_col]] - .data[[median_col]],
      diff_abs    = abs(diff_signed),
      rel_diff    = diff_signed / .data[[median_col]]
    )
  
  # 3) Plot
  p <- ggplot(df_plot,
              aes(x = .data[[pct_col]],
                  y = diff_abs,
                  color = .data[[time_col]])) +
    geom_point(alpha = 0.6) +
    geom_smooth(se = FALSE, method = "loess") +
    facet_wrap(as.formula(paste("~", time_col)), drop = FALSE, scales = "free", nrow = 1) +
    labs(
      x = "Percentage of pathway used for BMC calculation",
      y = "Absolute difference in BMC\n(Pathway Score − Median Pathway)",
      title = partner,
      subtitle = pathwaySource
    ) +
    scale_color_manual(values = timepoint_cols)
  
  if (log_x) p <- p + scale_x_log10()
  if (log_y) p <- p + scale_y_log10()
  
  p
}




LU_pwSize_BMC = plot_bmc_diff_vs_pct_overlap(df = LU_pwScore_medianScore, log_x = F, partner = "LeidenU", pathwaySource = "WGCNA Modules") +
  theme(plot.title = element_text(color = "blue2"))

SC_pwSize_BMC = plot_bmc_diff_vs_pct_overlap(df = Sciensano_pwScore_medianScore, log_x = F, partner = "Sciensano", pathwaySource = "HALLMARK Pathways") + 
  theme(plot.title = element_text(color = "forestgreen"))


fig5d = wrap_elements(LU_pwSize_BMC / SC_pwSize_BMC + plot_layout(axis = "collect", guides = "collect"))

# generate title

fig5d_lab <- wrap_elements(panel = textGrob("Effect of pathway coverage on disagreement between BMC methods", gp = gpar(fontface = "bold", fontsize = 18)))


fig5d_total = wrap_elements(fig5d_lab / fig5d + plot_layout(heights = c(0.5, 7)))
fig5d_total

ggsave(plot = fig5d_total, file.path(plotDir, "fig5d.pdf"), width = 15, height = 8)


```


## Generate total figure 5
```{r}
fig5_top = wrap_elements(fig5a) + wrap_elements(fig5c_total) + plot_layout(widths = c(1.4, 2.1), nrow = 1, ncol = 2)

fig5 = wrap_elements(fig5_top) /
  fig5d_total /
  fig5e_total +
  plot_layout(heights = c(10, 6, 6), ncol = 1) + plot_annotation(tag_levels = "A")

fig5

ggsave(plot = fig5, file.path(plotDir, "Figure5_v3.pdf"), width = 18, height = 22)


fig5 = (fig5a_total + fig5b_total + plot_layout(widths = c(1.4, 2.1), ncol = 2)) /
  fig5c_total /
  fig5d_total +
  plot_layout(heights = c(10, 6, 6), ncol = 1) + plot_annotation(tag_levels = "A")

fig5

ggsave(plot = fig5, file.path(plotDir, "Figure5.pdf"), width = 18, height = 22)


```