---
title: "ManuscriptFigure2"
author: "IBBruns"
date: "2025-04-22"
output: html_document
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 13,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 11,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)

# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue 
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)

```



MANUSCRIPT FIGURE 2
Here, we will generate figure 2 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, here, ggpubr, patchwork, grid, GGally, ggplotify)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))


tpod_original = data.table::fread(file.path(outputDir, "tpod_calculations_alltimepoints.txt")) %>%
  mutate(time_num = as.numeric(gsub("h", "", timepoint)),
         timepoint = factor(timepoint, levels = c("4h", "8h", "16h", "24h", "48h", "72h")),
         method = factor(method, levels = c("5th percentile", "25th ranked gene", "LCRD", "First mode", "Kneedle")))


tpod_bootstrapping = data.table::fread(file.path(outputDir, "tpod_bootstrapping_ci_median.txt"))


DRGs = data.table::fread(file.path(outputDir, "DRG_PMFG_pertimepoint.txt")) %>%
  mutate(timepoint = factor(timepoint, levels = c("4h", "8h", "16h", "24h", "48h", "72h")))


```


# Generate figure


## Figure 2 
Here we will have:  
1. Barplot with Dose Responsive Genes per timepoint per analysis_summary + Post Model Filter retained genes.  
2. The accumulation plot per analysis_summary (all timepoints combined in one plot) to determine the most sensitive timepoint.  
3. A zoom of the accumulation plot per analysis_summary (to see the slight differences in the beginning).  
4. A combined accumulationplot of the most sensitive timepoint (24h).  
5. Overview of distribution-based tPODs including bootstrapping results.   


### Figure 2a

## Barplot

```{r}
label_steps = c(
  "DRG"  = "Dose Responsive Genes",
  "PMFG" = "Post Model Filter Genes"
)

  
fig2a = DRGs %>%
  ggplot(aes(x = analysis_summary, y = value, fill = value)) +
  geom_bar(stat = "identity") +
  facet_grid(step ~ timepoint,
             labeller = labeller(step = label_steps)) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "white", colour = "black"),
    strip.text.x = element_text(size = 14, face = "bold"),
    strip.text.y = element_text(size = 9, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_blank(),
    legend.position = "none"
  ) +
  scale_fill_gradient(low = "yellow", high = "red")

fig2a

# combine 
fig2a_total <- wrap_elements(
  fig2a + 
    plot_layout(guides = "collect", axis = "collect") +
    plot_annotation(title = "Number of Dose Responsive Genes")
)

fig2a_total

ggsave(plot = fig2a_total, file.path(plotDir, "fig2a.pdf"), width = 15, height = 8)

```


### Figure 2b

#### Plot first row
```{r, fig.width=10}
fig2b_LU = BMDaccumulationPlot(BMDoutput = LU_norm_BMD_select, analysis_summary = "BMDE-noWTT-CPM-RF-S5") +
  theme(plot.title = element_text(color = "blue3")) +
  ylim(c(0, 6300))

fig2b_SC = BMDaccumulationPlot(Sciensano_norm_BMD_select, analysis_summary = "BMDE-WTT-CPM-RF-S0") + 
  theme(plot.title = element_text(color = "forestgreen")) +
  ylim(c(0, 6300))

fig2b_AU = BMDaccumulationPlot(BMDoutput = AU_norm_BMD_select, analysis_summary = "DRO-Quad-UQ-RF-S0") +
  theme(plot.title = element_text(color = "firebrick")) +  
  ylim(c(0, 6300))

fig2b_BPI = BMDaccumulationPlot(BMDoutput = BPI_norm_BMD_select, analysis_summary = "DRO-Quad-VST-C10-S0") +
  theme(plot.title = element_text(color = "gray30")) +
  ylim(c(0, 6300))

fig2b_GU = BMDaccumulationPlot(BMDoutput = GU_norm_BMD_select, analysis_summary = "DRO-Quad-VST-RF-S0") + 
  theme(plot.title = element_text(color = "gold1")) +
  ylim(c(0, 6300))




fig2b_row1 = wrap_plots(
  fig2b_LU, fig2b_SC, fig2b_AU, fig2b_BPI, fig2b_GU,
  nrow = 1
)

# ggsave(plot = fig2b_row1, file.path(plotDir, "fig2b_firstrow.pdf"), width = 15, height = 5)

```



#### Zoom
Zoom in on first part (250 genes) plot

#### Plot zoom
```{r}

fig2b_LU_zoom = BMDaccumulationPlot_zoom(BMDoutput = LU_norm_BMD_select) + 
  theme(plot.title = element_text(color = "blue3"))

fig2b_SC_zoom = BMDaccumulationPlot_zoom(Sciensano_norm_BMD_select) + 
  theme(plot.title = element_text(color = "forestgreen"))

fig2b_AU_zoom = BMDaccumulationPlot_zoom(BMDoutput = AU_norm_BMD_select) + 
  theme(plot.title = element_text(color = "firebrick"))

fig2b_BPI_zoom = BMDaccumulationPlot_zoom(BMDoutput = BPI_norm_BMD_select) + 
  theme(plot.title = element_text(color = "gray30"))

fig2b_GU_zoom = BMDaccumulationPlot_zoom(BMDoutput = GU_norm_BMD_select) + 
  theme(plot.title = element_text(color = "gold1"))



fig2b_row2 = wrap_plots(
  fig2b_LU_zoom, fig2b_SC_zoom, fig2b_AU_zoom, fig2b_BPI_zoom, fig2b_GU_zoom,
  nrow = 1
)
fig2b_row2

# ggsave(plot = fig2b_row2, file.path(plotDir, "fig2b_secondrow.pdf"), width = 15, height = 5)

```


# Generate total figure 2b
```{r}
# combine
fig2b_total <- wrap_elements(
  (fig2b_row1 / fig2b_row2) +
  plot_layout(guides = "collect", axis = "collect") +
  plot_annotation(title = "BMC Accumulation")) &
  theme(
    legend.position = "bottom"
  )

fig2b_total

ggsave(
  plot = fig2b_total,
  filename = file.path(plotDir, "fig2b.pdf"),
  width = 15,
  height = 5.5
)
```





### Figure 2c
Select most sensitive timepoint (now 24h) and combine accumulation plots of partners

```{r}

fig2c_norm = norm_BMD_combined %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>%
  filter(timepoint == "24h") %>%
  group_by(analysis_summary) %>%
  arrange(finalBMD) %>%
  mutate(ranked_gene = row_number()) %>%
  ggplot(aes(finalBMD, y = ranked_gene)) +
  geom_point(stat = "identity", aes(color = analysis_summary), size = 1) +
  labs(y = "Accumulation", 
       x = "BMC") +
  scale_color_manual(values = c("firebrick", "gray30", "gold1", "blue2", "forestgreen"),
                     breaks = c("DRO-Quad-UQ-RF-S0", "DRO-Quad-VST-C10-S0", "DRO-Quad-VST-RF-S0", "BMDE-noWTT-CPM-RF-S5", "BMDE-WTT-CPM-RF-S0")) +
  theme(
    # legend.title = element_blank()
    legend.position = "none"
    )

fig2c_zoom = fig2c_norm +
  ylim(c(0, 250)) +
  xlim(0, 3) 


# combine
fig2c_total <- wrap_elements(
  (fig2c_norm / fig2c_zoom) +
  plot_layout(
    guides  = "collect",
    axis    = "collect",
    heights = c(5, 5)
  ) +
  plot_annotation(title = "BMC Accumulation",
                  subtitle = "24h")) &
  theme(
    legend.position = "bottom"
  )

fig2c_total

ggsave(
  plot = fig2c_total,
  filename = file.path(plotDir, "fig2c.pdf"),
  width = 5,
  height = 10.5
)

```




### Figure 2d
BMD correlations

```{r}

fig2d_cor <- GGally::ggpairs(
  norm_BMD_combined_wide,
  aes(color = timepoint),
  columns = c( "LU_finalBMD", "SC_finalBMD",  "AU_finalBMD","BPI_finalBMD", "GU_finalBMD"),
  columnLabels = c("BMDE-noWTT-CPM-RF-S5", "BMDE-WTT-CPM-RF-S0",  "DRO-Quad-UQ-RF-S0", "DRO-Quad-VST-C10-S0", "DRO-Quad-VST-RF-S0"),
  lower = list(continuous = GGally::wrap("points", size = 0.05)),
  diag  = list(
    continuous = GGally::wrap("densityDiag", alpha = 0.5),
    mapping    = ggplot2::aes(color = timepoint)
    # continuous   = GGally::wrap("blankDiag")
  )
) +
  theme_bw() +
  scale_color_manual(values = timepoint_cols) +
  scale_fill_manual(values = timepoint_cols) +  # for density fill, if used
  # if densityDiag uses color only, this is harmless but keeps things consistent
  theme(strip.background = element_rect(fill = "white", colour = "black"),
        strip.text = element_text(face = "bold", size = 9))


# ggmatrix -> gtable (grob) -> patchwork element
fig2d_cor_gg <- wrap_elements(full = ggmatrix_gtable(fig2d_cor))

# combine with a patchwork-native title
fig2d_total <- wrap_elements(
  fig2d_cor_gg +
  plot_annotation(title = "BMC Correlations"))

fig2d_total

ggsave(
  plot = fig2d_total,
  filename = file.path(plotDir, "fig2d.pdf"),
  width = 10,
  height = 10.5
)

```


### Figure 2e


## tPOD calculations + bootstrapping results
```{r}
timepoint_order = c("4h", "8h", "16h", "24h", "48h", "72h")
method_order = c("5th percentile", "25th ranked gene", "First mode", "Kneedle", "LCRD")


### combine original tpods + bootstrap results
tpod_plot_df <- tpod_bootstrapping %>%
  left_join(tpod_original,
            by = c("analysis_summary","timepoint","method")) %>%
  mutate(timepoint = factor(timepoint, levels = timepoint_order),
         method = factor(method, levels = method_order))



### OPTION 1: FACET BY TPOD METHOD
fig2e_bs = ggplot(tpod_plot_df,
       aes(x = timepoint, y = tpod_median,
           ymin = tpod_lower, ymax = tpod_upper,
           color = analysis_summary)) +
  geom_pointrange(
    position = position_dodge(width = 0.6)
  ) +
  geom_point(
    aes(y = tpod_orig),
    shape = 21,
    fill = "white",
    size = 2.3,
    stroke = 0.6,
    position = position_dodge(width = 0.6)
  ) +
  geom_line(aes(x=timepoint, y=tpod_orig, group = analysis_summary)) +
  facet_wrap(~ method, scales = "free_y", nrow = 1) +
  labs(
    x = "",
    y = "tPOD",
    color = ""
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.position = "none"
  ) +
  # scale_y_log10() +
  scale_color_manual(values = c("blue2", "forestgreen", "firebrick", "gold1", "gray30"),
                     breaks = c("BMDE-noWTT-CPM-RF-S5", "BMDE-WTT-CPM-RF-S0", "DRO-Quad-UQ-RF-S0", "DRO-Quad-VST-RF-S0", "DRO-Quad-VST-C10-S0"))


fig2e_bs

# combine
fig2e_bs_total <- wrap_elements(
  fig2e_bs +
  plot_annotation(title = "tPOD Estimates and Bootstrap Uncertainty")) &
  theme(
    legend.position = "bottom"
  )

fig2e_bs_total

ggsave(
  filename = file.path(plotDir, "fig2e.pdf"),
  width = 10,
  height = 4
)


```





## Create total figure 2
```{r, fig.width=17, fig.height=23}

figure2 = fig2a_total /
  fig2b_total /
  (fig2c_total + fig2d_total + plot_layout(widths = c(0.6, 1))) /
  fig2e_bs_total +
  plot_layout(guides = "collect", axis = "collect", heights = c(4,4,6, 3)) + plot_annotation(tag_levels = "A") 

figure2

ggsave(plot = figure2, file.path(plotDir, "Figure2.pdf"), width = 17, height = 23)


```




