---
title: "Supplemental Figure 1"
output: html_document
date: "2025-12-08"
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)


# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue (replacing teal)
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)
```

SUPPLEMENTAL FIGURE 1
Check overlap of genes passing pre-filter


## Manual settings
```{r}
studyNR = "EUT046"

```


## Load packages
```{r}
pacman::p_load(tidyverse, data.table, ComplexUpset, patchwork, purrr)
```


## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```


# Load data
```{r}

# BMD normalized counts
AU_norm_BMD = fread(file.path(inputDir, "Aristotle", paste0(studyNR, "_log2cpm_All_BMD_output.txt")), check.names = TRUE) %>%
  mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"),
  levels = c("4h", "8h", "16h", "24h", "48h", "72h")
))

GU_norm_BMD = fread(file.path(inputDir, "Ghent", "BMD_output_normalized_counts.txt"), sep = " ", check.names = TRUE) %>% mutate(timepoint = factor(
  paste0(timepoint, "h"),
  levels = c("4h", "8h", "16h", "24h", "48h", "72h")
))


LU_WTT = fread(file.path(inputDir, "Leiden", paste0(studyNR, "_log2cpm_WTT_output.txt")), check.names = TRUE) %>%
  mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"), 
  levels = c("4h", "8h", "16h", "24h", "48h", "72h")
  )) 


Sciensano_norm_BMD = fread(file.path(inputDir, "Sciensano", "BMD_output_normalized_counts.txt"), check.names = TRUE) %>% mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"), 
  levels = c("4h", "8h", "16h", "24h", "48h", "72h")
  ))

BPI_norm_BMD = fread(file.path(inputDir, "BPI" , "All_FilteredDRomicV3.txt"), check.names = TRUE) %>% mutate(timepoint = factor(
  timepoint, levels = c("4h", "8h", "16h", "24h", "48h", "72h")))


```


# Generate figure
Generate UpSet plot to show overlap between dose responsive genes

## Dose responsive genes (pre-filter)

### Get dose responsive genes per timepoint
```{r}
AU_genes = AU_norm_BMD %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
BPI_genes = BPI_norm_BMD %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
GU_genes = GU_norm_BMD %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
LU_genes = LU_WTT %>% filter(Adjusted.P.Value < 0.05) %>% dplyr::select(timepoint, "gene_symbol_entrez_id" = Probe.ID)
SC_genes = Sciensano_norm_BMD %>% filter(Prefilter.P.Value < 0.01 & Max.Fold.Change.Absolute.Value >= 1.5) %>% dplyr::select(timepoint, gene_symbol_entrez_id)

```


### Generate plot per timepoint
```{r}

UpSetOutput = NULL

all_timepoints = LU_genes %>% pull(timepoint) %>% unique()

for(time in all_timepoints) {
  # get genelist per timepoint
  geneList = list(
    AU = AU_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    BPI = BPI_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    GU = GU_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    LU = LU_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    SC = SC_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id)
  )
  
  
  all_genes <- sort(geneList %>% unlist(use.names = FALSE) %>% unique())
  
  membership_df <- tibble::tibble(Gene = all_genes, !!!lapply(geneList, function(v)
    all_genes %in% v)) %>%
    mutate(across(-Gene, as.logical))
  
  supfig1 = ComplexUpset::upset(
    membership_df,
    names(geneList),
    min_size = 1,
    matrix = (
      intersection_matrix(geom = geom_point(shape = 'circle filled', size = 3)) +
        scale_color_manual(
          values = c(
            'AU' = "firebrick",
            'BPI' = "gray30",
            'GU' = "gold1",
            'LU' = "blue2",
            'SC' = "forestgreen"
          ),
        )
    ),
    queries = list(
      upset_query(set = 'AU', fill = "firebrick"),
      upset_query(set = "BPI", fill = "gray30"),
      upset_query(set = "GU", fill = "gold1"),
      upset_query(set = "LU", fill = "blue2"),
      upset_query(set = "SC", fill = "forestgreen")
    )
  )
  
  print(supfig1)
  
  # ggsave(plot = supfig1, file.path(
  #   plotDir,
  #   paste0("supFig1_upset_", time, ".pdf")),
  #   width = 15,
  #   height = 6
  # )
  
  UpSetOutput[[time]] = wrap_elements(supfig1)
}


```


## Generate final figure
```{r}

supfig1_combined <- reduce(UpSetOutput, `+`) + plot_layout(nrow =6)
supfig1_combined

ggsave(file.path(plotDir, "SupplementalFigure1.pdf"), width=13, height = 20)

```





## Retained genes after post model filter

### Get dose responsive genes per timepoint
```{r}
AU_genes = AU_norm_BMD %>% filter(adjpvalue < 0.01) %>% 
  filter() %>%
  dplyr::select(timepoint, gene_symbol_entrez_id)




BPI_genes = BPI_norm_BMD %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
GU_genes = GU_norm_BMD %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
LU_genes = LU_WTT %>% filter(Unadjusted.P.Value < 0.05) %>% dplyr::select(timepoint, "gene_symbol_entrez_id" = Probe.ID)
SC_genes = Sciensano_norm_BMD %>% filter(Prefilter.P.Value < 0.01 & Max.Fold.Change.Absolute.Value >= 1.5) %>% dplyr::select(timepoint, gene_symbol_entrez_id)

```

### Generate plot per timepoint
```{r}
UpSetOutput = NULL

all_timepoints = LU_genes %>% pull(timepoint) %>% unique()

for(time in all_timepoints) {
  # get genelist per timepoint
  geneList = list(
    AU = AU_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    BPI = BPI_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    GU = GU_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    LU = LU_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    SC = SC_genes %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id)
  )
  
  
  all_genes <- sort(geneList %>% unlist(use.names = FALSE) %>% unique())
  
  membership_df <- tibble::tibble(Gene = all_genes, !!!lapply(geneList, function(v)
    all_genes %in% v)) %>%
    mutate(across(-Gene, as.logical))
  
  supfig1 = ComplexUpset::upset(
    membership_df,
    names(geneList),
    min_size = 1,
    matrix = (
      intersection_matrix(geom = geom_point(shape = 'circle filled', size = 3)) +
        scale_color_manual(
          values = c(
            'AU' = "firebrick",
            'BPI' = "gray30",
            'GU' = "gold1",
            'LU' = "blue2",
            'SC' = "forestgreen"
          ),
        )
    ),
    queries = list(
      upset_query(set = 'AU', fill = "firebrick"),
      upset_query(set = "BPI", fill = "gray30"),
      upset_query(set = "GU", fill = "gold1"),
      upset_query(set = "LU", fill = "blue2"),
      upset_query(set = "SC", fill = "forestgreen")
    )
  )
  
  print(supfig1)
  
  # ggsave(plot = supfig1, file.path(
  #   plotDir,
  #   paste0("supFig1_upset_", time, ".pdf")),
  #   width = 15,
  #   height = 6
  # )
  
  UpSetOutput[[time]] = wrap_elements(supfig1)
}




```

