---
title: "Supplemental Figure 2"
author: "Imke Bruns"
date: "2025-12-16"
output: html_document
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)

# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue 
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)

```



MANUSCRIPT FIGURE 2
Here, we will generate figure 2 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, patchwork, purrr)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}
LU_WTT = data.table::fread(file.path(inputDir, "Leiden", "EUT046_log2cpm_WTT_output.txt"), check.names = TRUE) %>% mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"), 
  levels = c("4h", "8h", "16h", "24h", "48h", "72h")
  ))

LU_BMD = data.table::fread(file.path(inputDir, "Leiden", "EUT046_log2cpm_BMD_output.txt"), check.names = TRUE) %>% mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"), 
  levels = c("4h", "8h", "16h", "24h", "48h", "72h")
  ))

LU_total_input = data.table::fread(file.path(inputDir, "Leiden", "BMD_input_72hr_log2cpmcounts.txt"))
LU_total_input = LU_total_input[-1,]

```


# Generate figure
We generate a Venn diagram to show the overlap of genes removed by the prefilter & by the post model filters. This also shows which genes are removed by the prefilter, but retained after model filters. 

```{r}
library(ggVennDiagram)
library(VennDiagram)

all_timepoints = LU_BMD %>% pull(timepoint) %>% unique()
venn_output = NULL


for(time in all_timepoints){
  x <- list(prefilter_removed = LU_total_input %>% filter(!gene_symbol_entrez_id %in% (LU_WTT %>% filter(timepoint == time) %>% filter(Adjusted.P.Value < 0.05))) %>% pull(gene_symbol_entrez_id),
            postmodel_removed = LU_BMD %>% filter(timepoint == time) %>% filter(Best.BMD < 0.01 | Best.BMD > 50 | Best.BMD.BMDL > 20 | Best.BMDU.BMD > 20 | Best.RSquared < 0.8) %>% pull(Probe.ID))
  
  p1 = ggVennDiagram(x, category.names = c("Removed by pre-filter\n (WTT padj < 0.05)", "Removed by post-model filters")) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
  labs(
    # caption = "Top: Removed by post-model filters; Bottom: Removed by pre-filter (padj < 0.05)",
    subtitle = time
  ) +
    theme(plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ))
  
  print(p1)
  
  venn_output[[time]] = p1

}


reduce(venn_output, `+`) + plot_layout(nrow =3) 

ggsave(file.path(plotDir, "SupplementalFigure2_new.pdf"), width = 15, height = 15)



```



```{r}
time = "4h"

prefilter_removed = LU_total_input %>% filter(!gene_symbol_entrez_id %in% (LU_WTT %>% filter(timepoint == time) %>% filter(Adjusted.P.Value < 0.05))) %>% pull(gene_symbol_entrez_id)

postmodel_removed = LU_BMD %>% filter(timepoint == time) %>% filter(Best.BMD < 0.01 | Best.BMD > 50 | Best.BMD.BMDL > 20 | Best.BMDU.BMD > 20 | Best.RSquared < 0.8) %>% pull(Probe.ID)


prefilter_removed[(!prefilter_removed %in% postmodel_removed) %>% which()]




```




