---
title: "Supplemental Figure 2"
output: html_document
date: "2025-12-08"
---


# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)


# define 6 custom colors
timepoint_cols <- c(
  "#1f77b4",  # blue (replacing teal)
  "#d95f02",  # orange
  "#7570b3",  # violet
  "#e7298a",  # magenta/pink
  "#66a61e",  # green
  "#e6ab02"   # mustard yellow
)
```


SUPPLEMENTAL FIGURE 2
Check overlap of genes passing pre-filter


## Manual settings
```{r}
studyNR = "EUT046"

```


## Load packages
```{r}
pacman::p_load(tidyverse, data.table, ComplexUpset, patchwork, purrr)
```


## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```


# Load data
```{r}
timepoint_levels = c("4h", "8h", "16h", "24h", "48h", "72h")

# load data after prefilter but before post model filters
## BMD normalized counts
AU_norm_BMD_pref = fread(file.path(inputDir, "Aristotle", paste0(studyNR, "_log2cpm_All_BMD_output.txt")), check.names = TRUE) %>%
  mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"),
  levels = timepoint_levels
))

BPI_norm_BMD_pref = fread(file.path(inputDir, "BPI" , "All_UnfilteredDRomicV3.txt"), check.names = TRUE) %>% mutate(timepoint = factor(
  timepoint, levels = timepoint_levels)) %>% dplyr::rename("gene_symbol_entrez_id" = id)


folder_path_GU <- "input/EUT046/Ghent/Ab_initio_UGent/Output_046/Bootstrap_unfiltered"
GU_norm_BMD_pref <- list.files(folder_path_GU, pattern = "\\.txt$", full.names = TRUE) %>%
  map_dfr(~ {
    fread(.x, check.names = TRUE) %>%
      mutate(
        timepoint = factor(paste0(
          stringr::str_extract(basename(.x), "(?<=Time_)\\d+"),
          "h"
        ), levels = timepoint_levels
      )
      )
  }) %>%
  filter(adjpvalue < 0.01) %>%
  dplyr::rename("gene_symbol_entrez_id" = id)

LU_WTT = fread(file.path(inputDir, "Leiden", paste0(studyNR, "_log2cpm_WTT_output.txt")), check.names = TRUE) %>%
  mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"), 
  levels = timepoint_levels
  )) 


Sciensano_norm_BMD_pref = fread(file.path(inputDir, "Sciensano", "BMD_output_normalized_counts.txt"), check.names = TRUE) %>% mutate(timepoint = factor(
  str_extract(Analysis, "\\d{1,2}h"), 
  levels = timepoint_levels
  ))



# load data after filering post model filters
load(file.path(outputDir, "WrangledInput", "WrangledInputData.RData"))



```


# Generate figure
Generate UpSet plot to show overlap between dose responsive genes

## Dose responsive genes (pre-filter)

### Get dose responsive genes per timepoint
```{r}
AU_genes_pref = AU_norm_BMD_pref %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
BPI_genes_pref = BPI_norm_BMD_pref %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
GU_genes_pref = GU_norm_BMD_pref %>% filter(adjpvalue < 0.01) %>% dplyr::select(timepoint, gene_symbol_entrez_id)
LU_genes_pref = LU_WTT %>% filter(Adjusted.P.Value < 0.05) %>% dplyr::select(timepoint, "gene_symbol_entrez_id" = Probe.ID)
SC_genes_pref = Sciensano_norm_BMD_pref %>% filter(Prefilter.P.Value < 0.01 & Max.Fold.Change.Absolute.Value >= 1.5) %>% dplyr::select(timepoint, gene_symbol_entrez_id)

```


### Generate plot per timepoint
```{r}

UpSetOutput = NULL

all_timepoints = timepoint_levels

for(time in all_timepoints) {
  # get genelist per timepoint
  geneList = list(
    AU = AU_genes_pref %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    BPI = BPI_genes_pref %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    GU = GU_genes_pref %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    LU = LU_genes_pref %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    SC = SC_genes_pref %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id)
  )
  
  
  all_genes <- sort(geneList %>% unlist(use.names = FALSE) %>% unique())
  
  membership_df <- tibble::tibble(Gene = all_genes, !!!lapply(geneList, function(v)
    all_genes %in% v)) %>%
    mutate(across(-Gene, as.logical))
  
  supfig1 = ComplexUpset::upset(
    membership_df,
    names(geneList),
    min_size = 1,
    matrix = (
      intersection_matrix(geom = geom_point(shape = 'circle filled', size = 3)) +
        scale_color_manual(
          values = c(
            'AU' = "firebrick",
            'BPI' = "gray30",
            'GU' = "gold1",
            'LU' = "blue2",
            'SC' = "forestgreen"
          ),
        )
    ),
    queries = list(
      upset_query(set = 'AU', fill = "firebrick"),
      upset_query(set = "BPI", fill = "gray30"),
      upset_query(set = "GU", fill = "gold1"),
      upset_query(set = "LU", fill = "blue2"),
      upset_query(set = "SC", fill = "forestgreen")
    )
  )
  
  print(supfig1)
  
  # ggsave(plot = supfig1, file.path(
  #   plotDir,
  #   paste0("supFig1_upset_", time, ".pdf")),
  #   width = 15,
  #   height = 6
  # )
  
  UpSetOutput[[time]] = wrap_elements(supfig1)
}


```


## Generate final supplemental figure 2a
```{r}

supfig2a_combined <- wrap_elements(reduce(UpSetOutput, `+`) + plot_layout(nrow = 6))
supfig2a_combined

# generate title
supfig2a_lab = wrap_elements(panel = textGrob("Overlap Genes after Prefilter", gp = gpar(fontface = "bold", fontsize = 24)))


# create total supfig2a
supfig2a_total = wrap_elements(supfig2a_lab / supfig2a_combined + plot_layout(heights = c(0.2, 10)))

ggsave(plot = supfig2a_total, file.path(plotDir, "supfig2a.pdf"), width=10, height = 20)

```





## Retained genes after post model filter

### Get dose responsive genes per timepoint
```{r}
AU_genes_pmf = AU_norm_BMD_select %>% dplyr::select(timepoint, gene_symbol_entrez_id)
BPI_genes_pmf = BPI_norm_BMD_select  %>% dplyr::select(timepoint, gene_symbol_entrez_id)
GU_genes_pmf = GU_norm_BMD_select %>% dplyr::select(timepoint, gene_symbol_entrez_id)
LU_genes_pmf = LU_norm_BMD_select %>% dplyr::select(timepoint, gene_symbol_entrez_id)
SC_genes_pmf = Sciensano_norm_BMD_select %>% dplyr::select(timepoint, gene_symbol_entrez_id)

```


### Generate plot per timepoint
```{r}

UpSetOutput = NULL

all_timepoints = timepoint_levels

for(time in all_timepoints) {
  # get genelist per timepoint
  geneList = list(
    AU = AU_genes_pmf %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    BPI = BPI_genes_pmf %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    GU = GU_genes_pmf %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    LU = LU_genes_pmf %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id),
    SC = SC_genes_pmf %>% filter(timepoint == time) %>% pull(gene_symbol_entrez_id)
  )
  
  
  all_genes <- sort(geneList %>% unlist(use.names = FALSE) %>% unique())
  
  membership_df <- tibble::tibble(Gene = all_genes, !!!lapply(geneList, function(v)
    all_genes %in% v)) %>%
    mutate(across(-Gene, as.logical))
  
  supfig1 = ComplexUpset::upset(
    membership_df,
    names(geneList),
    min_size = 1,
    matrix = (
      intersection_matrix(geom = geom_point(shape = 'circle filled', size = 3)) +
        scale_color_manual(
          values = c(
            'AU' = "firebrick",
            'BPI' = "gray30",
            'GU' = "gold1",
            'LU' = "blue2",
            'SC' = "forestgreen"
          ),
        )
    ),
    queries = list(
      upset_query(set = 'AU', fill = "firebrick"),
      upset_query(set = "BPI", fill = "gray30"),
      upset_query(set = "GU", fill = "gold1"),
      upset_query(set = "LU", fill = "blue2"),
      upset_query(set = "SC", fill = "forestgreen")
    )
  )
  
  print(supfig1)
  
  # ggsave(plot = supfig1, file.path(
  #   plotDir,
  #   paste0("supFig1_upset_", time, ".pdf")),
  #   width = 15,
  #   height = 6
  # )
  
  UpSetOutput[[time]] = wrap_elements(supfig1)
}


```


## Generate final supplemental figure 2b
```{r}

supfig2b_combined <- wrap_elements(reduce(UpSetOutput, `+`) + plot_layout(nrow =6))

supfig2b_combined

# generate title
supfig2b_lab = wrap_elements(panel = textGrob("Overlap Genes after Post Model Filters", gp = gpar(fontface = "bold", fontsize = 24)))


# create total supfig2a
supfig2b_total = wrap_elements(supfig2b_lab / supfig2b_combined + plot_layout(heights = c(0.2, 10)))

ggsave(plot = supfig2b_total, file.path(plotDir, "supfig2b.pdf"), width=10, height = 20)

```



## Generate final supplemental figure 2
```{r}
supfig2_total = supfig2a_total + supfig2b_total + plot_layout(ncol = 2, widths = c(0.8, 1))

supfig2_total

ggsave(plot = supfig2_total, file.path(plotDir, "SupplementalFigure2.pdf"), width = 20, height = 20)

```

