---
title: "SupplementalFigure1"
output: html_document
date: "2025-04-24"
---

# Set up script
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
standard_plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold",
      vjust = 2,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      face = "italic",
      vjust = 2,
      hjust = 0.5
    ),
    axis.title.x = element_text(size = 12, vjust = 0.25),
    axis.title.y = element_text(size = 12, vjust = 1),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )


theme_set(standard_plotTheme)
```



MANUSCRIPT SUPPLEMENTAL FIGURE 1
Here, we will generate supplemental figure 1 from the PARC ab initio manuscript

# Settings

## Set studyNR
```{r}
studyNR = "EUT046"
```


## Load packages
```{r}
pacman::p_load(tidyverse, here, patchwork, grid)
```


## Load functions
```{r}
source(here("functions.R"), local = knitr::knit_global())
```



## Directories
```{r}
inputDir = file.path(here(), "input", studyNR)
outputDir = file.path(here(), "output", studyNR)

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

plotDir = file.path(here(), "figures", studyNR)

if(!dir.exists(plotDir)){
  dir.create(plotDir)
}
```



# Load data
```{r}

# BMD log2fc
# Leiden University
LU_fc_BMD = read.delim(file.path(inputDir, "Leiden", paste0(studyNR, "_log2fc_BMD_output.txt"))) %>%
  dplyr::rename("finalBMD" = Best.BMD, 
                "gene_symbol_entrez_id" = Probe.ID,
                "finalBMDL" = Best.BMDL,
                "finalBMDU" = Best.BMDU) %>%
  mutate(timepoint = factor(str_extract(Analysis, "\\d{1,2}hr"), levels = c("4hr", "8hr", "16hr", "24hr", "48hr", "72hr"))) %>% 
  filter(finalBMD >= 0.01*0.01 & finalBMD <= 50,
         Best.BMD.BMDL < 20,
         Best.BMDU.BMD < 20,
         Best.RSquared >= 0.8) # for Leiden filter needs te be applied; this is the direct output from BMDExpress


# R-ODAF
ODAF_fc_BMD = read.delim(file.path(inputDir, "ODAF", paste0(studyNR, "_RODAF_log2fc_BMD_output.txt"))) %>% 
  dplyr::rename("finalBMD" = "Best.BMD", 
                "gene_symbol_entrez_id" = "Probe.ID") %>%
  mutate(timepoint = factor(str_extract(Analysis, "\\d{1,2}hr"), levels = c("4hr", "8hr", "16hr", "24hr", "48hr", "72hr"))) %>% 
  filter(finalBMD >= 0.01*0.01 & finalBMD <= 50,
         Best.BMD.BMDL < 20,
         Best.BMDU.BMD < 20,
         Best.RSquared >= 0.8) # for Leiden filter needs te be applied; this is the direct output from BMDExpress


```



# Wrangle data
```{r}
LU_fc_select = LU_fc_BMD %>% dplyr::select(gene_symbol_entrez_id, timepoint, finalBMD) %>% mutate(partner = "LU")
ODAF_fc_select = ODAF_fc_BMD %>% dplyr::select(gene_symbol_entrez_id, timepoint, finalBMD) %>% mutate(partner = "R-ODAF")

df_list = list(LU_fc_select, ODAF_fc_select)

fc_BMD_combined = bind_rows(df_list) %>% as_tibble()
```



# Generate figure

## Accumulation plots
```{r}
sup1a_LU = BMDaccumulationPlot(BMDoutput = LU_fc_BMD, partner = "Leiden University") +
  theme(plot.title = element_text(color = "blue2")) + 
  ylim(c(0, 5300))

sup1a_ODAF = BMDaccumulationPlot(BMDoutput = ODAF_fc_BMD, partner = "R-ODAF") +
  theme(plot.title = element_text(color = "orange")) +
  ylim(c(0, 5300))


sup1a_row1 = sup1a_LU + sup1a_ODAF + plot_layout(guides = "collect", axis = "collect")
sup1a_row1


```


## Accumulation plots zoom
```{r}
sup1a_LU_zoom = BMDaccumulationPlot_zoom(BMDoutput = LU_fc_BMD)
sup1a_ODAF_zoom = BMDaccumulationPlot_zoom(BMDoutput = ODAF_fc_BMD)

sup1a_row2 = sup1a_LU_zoom + sup1a_ODAF_zoom + plot_layout(guides = "collect", axis = "collect")
sup1a_row2

```


## Generate total supplemental figure 1a
```{r}
# generate title
sup1a_lab <- wrap_elements(panel = textGrob("BMC accumulation (log2fc)", gp = gpar(fontface = "bold", fontsize = 18)))


# combine
sup1a = wrap_elements(sup1a_lab /
  sup1a_row1 /
  sup1a_row2 + plot_layout(guides = "collect", axis = "collect", heights = c(0.5, 2.5, 2.5)))


sup1a
ggsave(plot = sup1a, file.path(plotDir, "sup1a_total.pdf"), width = 7.5, height = 5.5)


```


# Supplemental figure 1b
```{r}
sup1b = fc_BMD_combined %>% separate(gene_symbol_entrez_id, into = c("gene_symbol", "entrez_id"), sep = "_") %>% 
  filter(timepoint == "24hr") %>%
  group_by(partner) %>%
  arrange(finalBMD) %>%
  mutate(ranked_gene = row_number()) %>%
  ggplot(aes(finalBMD, y = ranked_gene)) + 
  geom_point(stat = "identity", aes(color = partner), size = 1) + 
  labs(y = "Accumulation",
       x = "BMC",
       title = "BMC accumulation - 24 hours (log2fc)") +
  scale_color_manual(values = c("blue2", "orange"),
                     breaks = c("LU", "R-ODAF"),
                     labels = c("Leiden University", "R-ODAF")) +
  theme(legend.title = element_blank()) 



ggsave(plot = sup1b, file.path(plotDir, "sup1b.pdf"), width = 7.5, height = 5.5)

```


# Combine to make total suplemental figure 1
```{r}

sup1 = sup1a + sup1b

ggsave(plot = sup1, file.path(plotDir, "SupplementalFigure1.pdf"), width = 15, height = 5.5)

```



