---
title: "Finalsforpaper_EUT046"
author: "Friedel Dewulf"
date: "2024-06-28"
output: html_document
---

# Loading of packages
```{r, results = FALSE, message = FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(gtools)
library(DESeq2)
library(pheatmap)
library(apeglm)
library("edgeR")
library("org.Hs.eg.db")
library("DRomics")
library(clusterProfiler)

```

# Set parameters according to R-ODAF: platform is TempOSeq
```{r}
#minCoverage <- 1000000
#MinCount <- 0.5
#pAdjValue <- 0.05

Platform <- "TempO-seq"

if (Platform=="RNA-Seq"){
  minCoverage <- 500000
  MinCount<- 1
  pAdjValue<- 0.01 
} else if (Platform=="TempO-seq") {
  minCoverage <- 1000000
  MinCount<- 0.5
  pAdjValue<- 0.05 
} else { print("Platform/technology not recognized") }
```

# Load input files

```{r}
EUT046 <- read.csv("Data/RPTEC-TERT1_CIS/EUT046_counts_per_gene_per_sample_raw_final.csv",
                           header = TRUE, row.names = 1)
EUT046_meta <- read.csv("Data/RPTEC-TERT1_CIS/EUT046_metadata_final.csv",
                                header = TRUE, row.names = 1)


# For the EUT datasets, I need to specify the concentrations better
EUT046_meta$CONC <- unlist(lapply(strsplit(EUT046_meta$CONCENTRATION, "u"), function(x) x[1]))

# Remove timepoint zero
EUT046_meta <- EUT046_meta[EUT046_meta$TIME != 0,]
EUT046 <- EUT046[,rownames(EUT046_meta)]

```


Adjustments in the metadata
```{r}
EUT046_meta$CONC <- factor(EUT046_meta$CONC, levels = c(0,0.1,0.5,1,2.5,5,10,20,30,50), labels = c(0,0.1,0.5,1,2.5,5,10,20,30,50))
EUT046_meta$TIME <- factor(EUT046_meta$TIME, levels = c(4,8,16,24,48,72), labels = c(4,8,16,24,48,72))
EUT046_meta$CONC_TIME <- factor(paste(EUT046_meta$CONC, EUT046_meta$TIME, sep = "_"))

all(rownames(EUT046_meta) %in% colnames(EUT046))
EUT046 <- EUT046[, rownames(EUT046_meta)]

```


# Step 1: Quality control of samples based on sequening depth
```{r}
depth_046 <- data.frame(x = colSums(EUT046))
ggplot(depth_046, aes(x = x)) +
  geom_histogram(binwidth = 50000) +
  labs(title = "046") +
  geom_vline(xintercept = 500000, color = "red", size = 2) +
  geom_vline(xintercept = 1000000, color = "orange", size = 2)
min <- min(depth_046$x)
print(paste0("046: ",min))
```

# Step 2: Create the dds object + function DESeq
```{r}
dds <- DESeqDataSetFromMatrix(countData = EUT046, colData = EUT046_meta, design = ~ CONC_TIME)
dds_deseq <- DESeq(dds, quiet = TRUE)
```


# Step 3: Filter low readcounts: relevance filter

```{r}
SampPerGroup <- table(EUT046_meta[,"CONC_TIME"])
Counts <- counts(dds_deseq, normalized = TRUE)
rawcounts <- counts(dds_deseq, normalized = FALSE)
CPMdds <- cpm(counts(dds_deseq, normalized = TRUE))

Filter <- matrix(data=NA, ncol=3, nrow= nrow(Counts))
rownames(Filter) <- rownames(Counts)
colnames(Filter) <- c("Low","quantile","spike")
  
#apply the relevance filter
for (gene in 1:nrow(dds_deseq)){
  CountsPass <- NULL
  for (group in 1:length(SampPerGroup)){
    sampleCols <- grep(dimnames(SampPerGroup)[[1]][group], EUT046_meta[,"CONC_TIME"])
    Check <- sum(CPMdds[gene,sampleCols] >= MinCount) >= 0.75*SampPerGroup[group]
    CountsPass <- c(CountsPass,Check)
  }
  if ( sum(CountsPass) > 0) {Filter[gene,1] <- 1 }	else { Filter[gene,1] <- 0 }
}

dds_filtered <- dds_deseq[Filter[,1] == 1,]

passed_relevance <- Filter[,1]
compte <- Counts[Filter[,1] == 1,]
rawcounts <- rawcounts[Filter[,1] == 1,]
Filter2 <- Filter[rownames(Filter) %in% rownames(compte),]
print(paste0("Relevance filtering removed ",nrow(dds_deseq)- nrow(Filter2)," genes from the ",nrow(dds_deseq)," assessed. ", nrow(Filter2)," genes remaining",sep=""))



```


# Step 4: Extracting transformed values
```{r}
vst <- vst(dds_filtered)

vst_mat <- assay(vst)
vst_cor <- cor(vst_mat)
```

  


# Step 5: Quality control samples: Hierarchical clustering: remover when replicates correlation is < 0.80
```{r}
pheatmap(vst_cor, annotation = EUT046_meta[,c("TIME","CONC")], fontsize_row = 3, fontsize_col = 3)

```
There is in general some samples under 0.8, but maybe they are not replicates.
The goal now is too see whether these low correlations might be between replicates.

```{r}
# vector of all the unique treatments
conc_time <- unique(EUT046_meta$CONC_TIME)

for (i in 1:length(conc_time)){
  conc_time_temp <- conc_time[i]
  samples_temp <- rownames(EUT046_meta)[which(EUT046_meta$CONC_TIME == conc_time_temp)]
  cor_temp <- vst_cor[samples_temp, samples_temp]
  cor_temp_decision <- cor_temp < 0.8
  if (sum(cor_temp_decision) > 0) {
    print(conc_time_temp)
  }
  
}

```

Correlations between replicates are always > 0.8


# Step 6: Add other gene id's

```{r}
manifest <- read.table("Data/Manifests/Temposeq_manifest_Human_WT_1.2_realignment_96percent_2023-05-15.txt", header = TRUE)
```

```{r}
probes <- data.frame(Probe_ID = rownames(dds_filtered))

probes_annot <- left_join(probes, manifest[, c("Probe_ID", "entrez_id", "gene_symbol")], by = "Probe_ID")

length(unique(probes_annot$entrez_id)) # --> So there are quite some probes with the same entrez id

symbol_entrez <- paste(probes_annot$gene_symbol, probes_annot$entrez_id, sep = "_")

symbol_entrez <- data.frame(symbol_entrez = symbol_entrez)

bio_annot <- cbind(symbol_entrez, probes_annot)

```




# Prepare data for DRomics

I need the raw counts per time point as an input for DRomics
I will sum over entrez id's!

```{r}
times <- c(4,8,16,24,48,72)

for (i in 1:length(times)){
  t <- times[i]
  temp <- counts(dds_filtered[, dds_filtered$TIME == t], normalized = FALSE)
  
  temp2 <- cbind(bio_annot, temp)
  
  # Sum the values for genes with the same Entrez ID
  gene_expression_summed <- temp2 %>%
  group_by(symbol_entrez) %>%
  summarise(across(starts_with("S_"), sum))
  
  write.table(gene_expression_summed, file = paste0("Output_046/Rawcounts_",t, ".txt"), sep = "\t", col.names = TRUE, row.names = FALSE)
  write.table(dds_filtered[, dds_filtered$TIME == t]$CONC, file = paste0("Output_046/CONC",t, ".txt"), row.names = FALSE, col.names = FALSE, sep = "\t")
}
```

# DRomics
```{r}
perc_5_list <- vector("list", length = length(times))
value_25th_list <- vector("list", length = length(times))


for (i in 1:length(times)){
  t <- times[i]
  signalmatrix_temp <- read.table(paste0("Output_046/Rawcounts_", t, ".txt"), header = TRUE)
  rownames(signalmatrix_temp) <- signalmatrix_temp$symbol_entrez
  signalmatrix_temp$symbol_entrez <- NULL
  dose_temp <- read.table(paste0("Output_046/CONC", t, ".txt"))
  dose_temp <- dose_temp[,1]
  data4DRomics_temp <- formatdata4DRomics(signalmatrix = signalmatrix_temp,
                                   dose = dose_temp)
  o.RNAseq_temp <- RNAseqdata(data4DRomics_temp, transfo.method = "vst", backgrounddose = 0)
  
  # Selection of significantly responding items
  s_quad_temp <- itemselect(o.RNAseq_temp, select.method = "quadratic", FDR = 0.01)
  print(paste0("Items after filter: ", length(s_quad_temp$adjpvalue)))
  write.table(length(s_quad_temp$adjpvalue), file = paste0("Output_046/Itemselect_Time_", t, ".txt"))
  
  # Fit dose-response model
  f_temp <- drcfit(s_quad_temp, progressbar = FALSE)
  # Save the results of the fits
  write.table(f_temp$fitres,file = paste0("Output_046/Fits/Fits_Time_", t, ".txt"))
  
  #calculation of bmd
  r_temp <- bmdcalc(f_temp, z = 1, x = 10)
  # Save the bmd results
  write.table(r_temp$res,file = paste0("Output_046/BMDs/BMD_Time_", t, ".txt"), row.names = FALSE)
  
  # Calculation of confidence intervals on the BMDs by bootstrap
  b_temp <- bmdboot(r_temp, niter = 1000, progressbar = TRUE, parallel = "snow", ncpus = 12)
  # Save the unfiltered bootstrap results
  write.table(b_temp$res,file = paste0("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_", t, ".txt"), row.names = FALSE)
  # Filtering BMD according to estimation quality:
  # I will only keep the items with finite BMD point estimate and CI bounds
  subres <- bmdfilter(b_temp$res, BMDtype = "zSD", BMDfilter = "finiteCI")
  #Save the filtered bootstrap results
  write.table(subres,file = paste0("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_", t, ".txt"), row.names = FALSE)
  
  # Calculate the 5th percentile and the 25th lowest ranked gene
  perc_5 <- quantile(subres$BMD.zSD, 0.05)
  value_25th <- sort(subres$BMD.zSD)[25]
  perc_5_list[[i]] <- perc_5
  value_25th_list[[i]] <- value_25th
  
  # Biological interpretation
  #1) Import the dataframe with DRomics results: the output $res of bmdcalc or bmdboot functions
  #I will use the subres
  subres$entrez <- as.numeric(lapply(strsplit(subres$id, split = "_"), function(x) x[2]))
  #2) Import the dataframe with biological annotation
  # I will load the hallmark gene sets from the molecular signature database
  library(msigdbr)
  h_gene_sets_human <- msigdbr(species = "Homo sapiens", category = "H")
  #manifest <- read.delim("Data/Manifests/Temposeq_manifest_Human_WT_1.2_realignment_96percent_2023-05-15.txt", header = TRUE, stringsAsFactors = FALSE)
  
  #3) Merging of both previous dataframes
  extendedres_temp <- merge(x = subres, y = h_gene_sets_human, by.x = "entrez", by.y = "entrez_gene")
  
  # Overrepresentation analysis of the hallmark gene sets: determine p values for each gene sets
  ora_results <- enricher(subres$entrez, TERM2GENE = h_gene_sets_human[,c("gs_name","entrez_gene")], pvalueCutoff = 0.10, pAdjustMethod = "fdr")
  ora <- ora_results@result

  
  tr <- sensitivityplot(extendedres_temp, BMDtype = "zSD", group = "gs_name", 
                  BMDsummary = "median.and.IQR", BMD_log_transfo = FALSE)
  
  # Now I will summarise the information together with the overrepresentation analysis
  tr_data <- tr$data
  colnames(tr_data)[1] <- "ID"
  ora_tot <- left_join(tr_data, ora, by = "ID")
  # Save these results
  write.table(ora_tot,file = paste0("Output_046/ORA_HALLMARK_Time_", t, ".txt"), row.names = FALSE)
  #Now also only save the results with p < 0.10
  ora_tot_p <- ora_tot[ora_tot$pvalue <= 0.10,]
  write.table(ora_tot_p,file = paste0("Output_046/ORA_HALLMARK_p0.10_Time_", t, ".txt"), row.names = FALSE)
}

```


# Prepare the output files to be appropriate for the reporting framework
```{r}
# sample metadata
samplemetadata <- read.csv("Data/RPTEC-TERT1_CIS/EUT046_metadata_final.csv", header = TRUE)
write.table(samplemetadata, file = "input/sample_metadata.txt", row.names = FALSE)

# normalized counts: BMD_output_normalized_counts
# Question: should this be the filtered counts (after relevance filter)? --> I will create both
normcounts <- counts(dds_filtered, normalized = TRUE)
normcounts2 <- cbind(bio_annot, normcounts)
# Sum the values for genes with the same Entrez ID
normcounts_summed <- normcounts2 %>%
  group_by(symbol_entrez) %>%
  summarise(across(starts_with("S_"), sum))
normcounts_summed <- data.frame(normcounts_summed)
colnames(normcounts_summed)[1] <- "gene_symbol_entrez_id"

write.table(normcounts_summed, file = "input/normalized_counts.txt")


# pathway overview
library(msigdbr)
h_gene_sets_human <- msigdbr(species = "Homo sapiens", category = "H")
pathoverview <- data.frame(gene_symbol_entrez_id = paste(h_gene_sets_human$gene_symbol, h_gene_sets_human$entrez_gene, sep = "_"), pathway = h_gene_sets_human$gs_name)
write.table(pathoverview, file = "input/pathway_overview.txt")


# BMD output normalized counts
BMD4 <- read.table("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_4.txt", header = TRUE)
BMD8 <- read.table("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_8.txt", header = TRUE)
BMD16 <- read.table("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_16.txt", header = TRUE)
BMD24 <- read.table("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_24.txt", header = TRUE)
BMD48 <- read.table("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_48.txt", header = TRUE)
BMD72 <- read.table("Output_046/Bootstrap_filtered/Bootstrap_filtered_Time_72.txt", header = TRUE)
BMD4$timepoint <- "4hr"
BMD8$timepoint <- "8hr"
BMD16$timepoint <- "16hr"
BMD24$timepoint <- "24hr"
BMD48$timepoint <- "48hr"
BMD72$timepoint <- "72hr"
BMD_tot <- rbind(BMD4, BMD8, BMD16, BMD24, BMD48, BMD72)

colnames(BMD_tot)[1] <- "gene_symbol_entrez_id"
write.table(BMD_tot, file = "input/BMD_output_normalized_counts.txt")

# Change some of the colnames of BMD_output_normalized_counts
BMD_tot <- read.table("input/BMD_output_normalized_counts.txt", header = TRUE)
BMD_tot_renames <- BMD_tot
colnames(BMD_tot_renames)[which(colnames(BMD_tot_renames) == "BMD.zSD")] <- "Best.BMD"
colnames(BMD_tot_renames)[which(colnames(BMD_tot_renames) == "BMD.zSD.lower")] <- "Best.BMDL"
colnames(BMD_tot_renames)[which(colnames(BMD_tot_renames) == "BMD.zSD.upper")] <- "Best.BMDU"
write.table(BMD_tot_renames, file = "input/BMD_output_normalized_counts.txt")

# pathways
# I have the values for all the pathways, but also files with pathways wherefore the p values is below 0.10
#SAVE UNFILTERED
HALLMARK4 <- read.table("Output_046/ORA_HALLMARK_Time_4.txt", header = TRUE)
HALLMARK8 <- read.table("Output_046/ORA_HALLMARK_Time_8.txt", header = TRUE)
HALLMARK16 <- read.table("Output_046/ORA_HALLMARK_Time_16.txt", header = TRUE)
HALLMARK24 <- read.table("Output_046/ORA_HALLMARK_Time_24.txt", header = TRUE)
HALLMARK48 <- read.table("Output_046/ORA_HALLMARK_Time_48.txt", header = TRUE)
HALLMARK72 <- read.table("Output_046/ORA_HALLMARK_Time_72.txt", header = TRUE)
HALLMARK4$timepoint <- "4hr"
HALLMARK8$timepoint <- "8hr"
HALLMARK16$timepoint <- "16hr"
HALLMARK24$timepoint <- "24hr"
HALLMARK48$timepoint <- "48hr"
HALLMARK72$timepoint <- "72hr"
HALLMARK_tot <- rbind(HALLMARK4, HALLMARK8, HALLMARK16, HALLMARK24, HALLMARK48, HALLMARK72)

pathwaysBMD <- data.frame(Gene.That.Passed.All.Filters = HALLMARK_tot$nb_of_items, Pathway.Name = HALLMARK_tot$ID, BMD.firstquartile = HALLMARK_tot$firstquartile, BMD.Median = HALLMARK_tot$secondquartile, BMD.thirdquartile = HALLMARK_tot$thirdquartile, GeneRatio = HALLMARK_tot$GeneRatio, BgRatio = HALLMARK_tot$BgRatio, pvalue = HALLMARK_tot$pvalue, p.adjust = HALLMARK_tot$p.adjust, qvalue = HALLMARK_tot$qvalue, nb_of_items = HALLMARK_tot$nb_of_items, timepoint = HALLMARK_tot$timepoint)

write.table(pathwaysBMD, file = "input/medianBMD_pathways_normalized_counts_ALL.txt")

# Now also create file with filtering on p.adjust < 0.10 and nb_of_items >=3
pathwaysBMD_filtered <- read.table(file = "input/medianBMD_pathways_normalized_counts_ALL.txt") %>% filter(nb_of_items >= 3) %>% filter(p.adjust <= 0.10)

write.table(pathwaysBMD_filtered, file = "input/medianBMD_pathways_normalized_counts_filteredonp0.10andn.txt")

# Now also create file with filtering on nb_of_items >=3
pathwaysBMD_filtered <- read.table(file = "input/medianBMD_pathways_normalized_counts_ALL.txt") %>% filter(nb_of_items >= 3)

write.table(pathwaysBMD_filtered, file = "input/medianBMD_pathways_normalized_counts_filteredonn.txt")



```
I also want to compare how many genes were removed before and after filtering with finiteCI

```{r}
BMD4_un <- read.table("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_4.txt", header = TRUE)
BMD8_un <- read.table("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_8.txt", header = TRUE)
BMD16_un <- read.table("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_16.txt", header = TRUE)
BMD24_un <- read.table("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_24.txt", header = TRUE)
BMD48_un <- read.table("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_48.txt", header = TRUE)
BMD72_un <- read.table("Output_046/Bootstrap_unfiltered/Bootstrap_unfiltered_Time_72.txt", header = TRUE)

BMD4_n <- nrow(BMD4)
BMD4_un_n <- nrow(BMD4_un)
BMD8_n <- nrow(BMD8)
BMD8_un_n <- nrow(BMD8_un)
BMD16_n <- nrow(BMD16)
BMD16_un_n <- nrow(BMD16_un)
BMD24_n <- nrow(BMD24)
BMD24_un_n <- nrow(BMD24_un)
BMD48_n <- nrow(BMD48)
BMD48_un_n <- nrow(BMD48_un)
BMD72_n <- nrow(BMD72)
BMD72_un_n <- nrow(BMD72_un)


```


