---
output:
  pdf_document: default
  html_document: default
---
# Pipeline for *limma-voom* & *DRomics* analysis of EUT046

---
title: "PARC WP 5.3.1 Ab Initio"
author: "D.R. Schultz"
date: "03 July 2024"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Settings {.tabset}

## Study information
* Cell culture: RPTEC/TERT1
* Compounds: Cisplatin 
Formulation preparation: 
* Cisplatin: cisplatin reconstituted in DMEM
Dose levels: 
* Cisplatin: 0, 0.1, 0.5, 1, 2.5, 5, 10, 20, 30, 50 uM
* Time points: 0, 4, 8, 16, 24, 48, 72 hours
Transcriptome analysis: at the specified timepoints, monolayers were lysed using lysis buffer and collected, samples were then snap frozen and send to BioClavis for TempO-seq analysis


## Perform differential expression analysis using *limma-voom*
### Load required libraries
pacman::p_load(
  BiocManager, limma, edgeR, biobroom, janitor, ggplot2, dplyr, tidyverse, clusterProfiler, broom, 
  ggrepel, ggforce, RSQLite, Biobase, GO.db, pvclust, BiocParallel, 
  pheatmap, RColorBrewer, reshape2, DOSE, purrr, tibble, skimr, doRNG, stringr
)

```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
suppressPackageStartupMessages({
  if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman", quiet = TRUE)
  pacman::p_load(
    BiocManager, limma, Glimma, edgeR, biobroom, pca3d, magick, 
    EnhancedVolcano, janitor, ggplot2, dplyr, tidyverse, clusterProfiler, broom, 
    ggrepel, ggforce, RSQLite, Biobase, ggvenn, GO.db, pvclust, BiocParallel, 
    pheatmap, RColorBrewer, reshape2, DOSE, purrr, tibble, missForest, skimr, doRNG, 
    stringr, msigdbr
  )
})
```

### Read in files: 
```{r, echo=FALSE}
inputDir = "C:/Users/HERACLES-admin/Desktop/WindowsNewShare/Bioinformatics/Transcriptomics/TempOSeq/Data/Human_EUT046 RPTEC-TERT1_CIS/Limma"
```

1.    Raw count data from TempoSeq

2.    Manifest file with sample information

3.    Sample data relationship file

  
```{r, class.source='fold-show', results = FALSE, message=FALSE}
Countsx <- read.csv("EUT046_counts_per_gene_per_sample_raw_final.csv")
Manifest <- read.delim("Temposeq_manifest_Human_WT_1.2_realignment_96percent_2023-05-15.txt")
SDRF <- read.delim("Targets.txt", check.names=FALSE, stringsAsFactors=FALSE)

```


```{r, echo=FALSE}
#reformat
entrez_id <-Manifest[,c(9,1)]
Countsy <- left_join(Countsx, entrez_id, by = c("Column1" = "Probe_ID"))
Counts <- Countsy[,c(206, 2:205)]

#Format SDRF
SDRF[,c("Sample", "Timepoint", "Treatment", "Replicate", "TrtGroup")]

Treatment <-SDRF[,"Treatment"]
Sample <- SDRF[, "Sample"]
Timepoint <- SDRF[, "Timepoint"]
Replicate <- SDRF[, "Replicate"]
TrtGroup <- SDRF[, "TrtGroup"] 


levels_trt <- unique(SDRF[,"Treatment"])
levels_time <- unique(SDRF[, "Timepoint"])
replicate <- unique(SDRF[, "Replicate"])
trtgrp <- unique(SDRF[, "TrtGroup"])

Treatment2 <- factor(Treatment, levels=levels_trt)
Replicate2 <- factor(Replicate, levels = replicate)
Timepoint2 <- factor(Timepoint, levels = levels_time)
TrtGroup <- factor(TrtGroup, levels=trtgrp)
```


```{r, echo=FALSE, results = FALSE}
countData <- as.matrix(Counts) # Each column represents a sample, and each row represents a gene

librarySizes <- numeric(ncol(countData))

for (i in 1:ncol(countData)) {
  # Convert counts to numeric
  counts <- as.numeric(countData[, i])
  
  # Calculate the library size for the current sample
  librarySize <- sum(counts)
  
  # Store the library size value in the vector or data frame
  librarySizes[i] <- librarySize
  
  # Print the library size for the current sample
  cat("Sample", i, "library size:", librarySize, "\n")
}

librarySizeObj <- librarySizes[c(2:205)]

librarySizeData <- data.frame(Sample = Sample, LibrarySize = librarySizes[c(2:205)],
                              Timepoint = Timepoint, Treatment = Treatment)

threshold1 <- as.numeric(quantile(librarySizeObj, probs = 0.1))
threshold25 <- as.numeric(quantile(librarySizeObj, probs = 0.25))

```

### Calculate and visualise library sizes to check for poor quality samples

```{r, class.source='fold-show'}

ggplot(librarySizeData, aes(x = Sample, y = LibrarySize)) +
  geom_jitter(aes (colour = Treatment, shape = Timepoint)) +
  xlab("Sample") +
  ylab("Library Size") +
  ggtitle("Library Size Distribution")+
  geom_hline(yintercept = threshold25) +
  scale_shape_manual(values = c(16, 17, 18, 19, 20, 21, 22))  # Specify shapes for each timepoint 

```

### Preprocess data (normalise (upper quartile), relevance filter, combine duplicate genes (based on EntrezID)

```{r, echo=FALSE, results = FALSE}
x <- as.data.frame(Countsy)
dim(x) 
geneid <- x$entrez_id

rownames(x) <- x$Column1
x <- x[,c(2:205)]
group <- paste0(SDRF$Treatment, "_", SDRF$Timepoint)
dge <- DGEList(counts = x, group = group, genes = x$Column1)

dim(dge)

#Upper quartile was found to perform better by Li et al 2020

```

```{r, class.source='fold-show'}
dge.norm <- calcNormFactors(dge,
                       method = "upperquartile",
                       p = 0.75)


cpm_values <- edgeR::cpm(dge.norm, normalized.lib.sizes = TRUE, log = FALSE)
min_replicates_threshold <- ceiling(0.75 * 3)
relevance_filtered_data <- cpm_values[rowSums(cpm_values >= 1) >= min_replicates_threshold, ]
relevance_filtered_data <- rownames_to_column(as.data.frame(relevance_filtered_data))

dim(relevance_filtered_data)
```

```{r, echo = FALSE}
entrez_data <- left_join(relevance_filtered_data, entrez_id, by = c("rowname" = "Probe_ID"))
entrez_data2 <- entrez_data[,c(206, 2:205)]
colnames(entrez_data2)[1] <- "geneid"
```

```{r, class.source='fold-show'}
dge.norm.sum <- aggregate(entrez_data2[, -1], by = list(entrez_data2$geneid), sum)
dim(dge.norm.sum)
```


```{r, echo = FALSE}

```

### Check and apply pseudocount for DRomics analysis

```{r, class.source='fold-show'}
df <- dge.norm.sum[,-1]
smallest_non_zero <- df %>%
  select_if(is.numeric) %>%  # Select only numeric columns
  unlist() %>%  # Convert to a vector
  .[. > 0] %>%  
  min()  # 
```

```{r,echo=FALSE}
largest_non_zero <- df %>%
  select_if(is.numeric) %>%  # Select only numeric columns
  unlist() %>% 
  .[. > 0] %>%  
  max()  # 

# range of rows with lowest count
rows_with_0_155 <- apply(df, 1, function(x) any(x == smallest_non_zero))
range_values <- apply(df[rows_with_0_155, ], 1, function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
names(range_values) <- which(rows_with_0_155)
range_values <- as.data.frame(range_values)
```

```{r,class.source='fold-show'}
pseudocount <- smallest_non_zero * 0.1 # 10% of lowest
dge.norm.sum2 <- column_to_rownames(dge.norm.sum, "Group.1")
pseudo_count_data <- dge.norm.sum2 + pseudocount
dge.norm.sum.log <- log2(pseudo_count_data)

```

### Visualise to check efficacy using MDS, PCA, correlations, densities, and boxplots

```{r, echo = FALSE, results = FALSE, warning=FALSE}
#MDS
cpm_values <- edgeR::cpm(dge.norm.sum[,-1], normalized.lib.sizes = TRUE)
plot(log(cpm_values))

num_groups <- length(levels_trt)
color_palette <- colorRampPalette(c("blue", "red"))(num_groups)
group_colors <- setNames(color_palette, levels_trt)
plotMDS(dge.norm.sum[,-1], labels = dge.norm.sum$samples$group, col = group_colors, main = "MDS")

#Correlations
DMEM <- cpm_values[, grepl("DMEM", colnames(cpm_values))]
corr_mat <- round(cor(DMEM),5) 
melted_corr_mat <- melt(corr_mat)

ggplot(data = melted_corr_mat, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  labs(x = "Variable 1", y = "Variable 2", fill = "Correlation") +
  ggtitle("Correlation Matrix Plot") +
  theme_minimal()

corr_mat <- round(cor(cpm_values),5) 
melted_corr_mat <- melt(corr_mat)
ggplot(data = melted_corr_mat, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  labs(x = "Variable 1", y = "Variable 2", fill = "Correlation") +
  ggtitle("Correlation Matrix Plot") +
  theme_minimal(base_size = 10) +  # Adjusts the base text size for theme_minimal
  theme(
    plot.title = element_text(size = 12),  # Adjust title size if needed
    axis.title = element_text(size = 11),  # Adjust axis titles size if needed
    axis.text = element_text(size = 8)     # Adjust axis text size if needed
  )

#PCA

data <- dge.norm.sum %>% 
  as.data.frame() %>% 
  dplyr::select(!Group.1)

#data <- column_to_rownames(data, "genes")

pca_matrix <- data %>% 
  as.matrix() %>% 
  t()

sample_pca <- prcomp(pca_matrix)
#pca_matrix[1:24, 1:5]
as_tibble(pca_matrix, rownames = "Sample")
pc_eigenvalues <- sample_pca$sdev^2
pc_eigenvalues <- tibble(PC = factor(1:length(pc_eigenvalues)),
                         variance = pc_eigenvalues) %>% 
  mutate(pct = variance/sum(variance)*100) %>% 
  mutate(pct_cum = cumsum(pct))

#pc_eigenvalues

pc_eigenvalues %>% 
  ggplot(aes(x = PC)) +
  geom_col(aes(y = pct)) + 
  geom_line(aes(y = pct_cum, group = 1)) +
  geom_point(aes(y = pct_cum)) +
  theme(panel.background = element_rect(fill = "lightgray", colour = "white")) +
  labs(x = "Principal Component", y = "Fraction of Variance Explained")

pc_scores <- sample_pca$x

pc_scores <- pc_scores %>% 
  as_tibble(rownames = "Sample")

pc_scores

suppressWarnings({
  # Assuming this is where the warning is generated
  my_colors <- brewer.pal(15, "Paired")
})
```

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
pc_scores %>% 
  ggplot(aes(x = PC1, y = PC2, 
             colour = Sample, shape = Timepoint2, size = 1)) +
  scale_shape_manual(values=1:nlevels(Timepoint2)) +
  ggtitle("PCA Plot of Preprocessed Count Data") +
  theme(plot.title = element_text(hjust = 0.5), 
        panel.background = element_rect(fill = "lightgray", colour = "white")) +
  geom_point(aes(colour = Sample), size = 3) +
  geom_label_repel(aes(label = Sample),
                   size = 2.5,
                   box.padding = 0.35,
                   point.padding = 0.5,
                   segment.color = 'grey50',
                   max.overlaps = 10) +
  theme_light() +
  guides(size = "none") +
  theme(legend.position="none")

dge.norm.sum3 <- DGEList(counts = dge.norm.sum2, group = group, genes = rownames(dge.norm.sum2))

L <- mean(dge.norm.sum3$samples$lib.size) * 1e-6
M <- median(dge.norm.sum3$samples$lib.size) * 1e-6

lcpm_raw <- cpm(dge, log = TRUE)
lcpm.cutoff <- log2(10/M + 3/L)
nsamples <- ncol(dge)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm_raw[,1]), col=col[1], lwd=2, ylim=c(0,0.6), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
  den <- density(lcpm_raw[,i])
  lines(den$x, den$y, col=col[i], lwd=2)
}

#legend("topright", samplenames, text.col=col, bty="n")

lcpm <- cpm(dge.norm.sum3, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.6), las=2, main="", xlab="")
title(main="B. Preprocessed Data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
  den <- density(lcpm[,i])
  lines(den$x, den$y, col=col[i], lwd=2)
}
#legend("topright", samplenames, text.col=col, bty="n")
par(mfrow=c(1,1))

boxplot(lcpm_raw, las=2, col=col, main="")
title(main="A. Raw data",ylab="Log-cpm")

boxplot(lcpm, las=2, col=col, main="")
title(main="A. Preprocessed data",ylab="Log-cpm")
```

### Continue with limma-voom with weights on separate time points

## Perform Benchmark Dose Analysis using *DRomics*
### Load libraries 
pacman::p_load(DRomics, dplyr, ggplot2, tidydr, tidyverse)

```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
suppressPackageStartupMessages({
  if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman", quiet = TRUE)
  pacman::p_load(DRomics, dplyr, ggplot2, tidydr, tidyverse, enrichR, knitr, clusterProfiler, org.Hs.eg.db,
                 enrichplot, DOSE, ggplot2, tinytex)
})

```
### Import data for benchmark dose analysis 
1.    Preprocessed universe data from limma (separated by timepoint, pseudocount, log base 2)

### Read in file and set initial conditions
```{r, echo=FALSE}
inputDir = "C:/Users/HERACLES-admin/Desktop/WindowsNewShare/Bioinformatics/Transcriptomics/TempOSeq/Data/Human_EUT046 RPTEC-TERT1_CIS/BMD/DRomics/8h"
```


```{r, class.source='fold-show', warning=FALSE, error=FALSE, message=FALSE}
o <- continuousomicdata("C:/Users/HERACLES-admin/Desktop/WindowsNewShare/Bioinformatics/Transcriptomics/TempOSeq/Data/Human_EUT046 RPTEC-TERT1_CIS/BMD/DRomics/8h/Preprocessed_Pseudo_Count_Log2_8h.txt", backgrounddose = 0, check = TRUE)
print(o)
```

```{r, echo=FALSE}
plot(o, col = "darkgreen")
```

### Select responsive items 
```{r, class.source = 'fold-show'}
s <- itemselect(o, FDR = 0.01) #select.method = 'quadratic', 
print(s)
```

### Fit dose-response models to responsive items
```{r, class.source = 'fold-show', warning=FALSE, error=FALSE, message=FALSE}
f <- drcfit(s, progressbar = TRUE, parallel = 'snow', ncpus = 6)
print(f)
```

```{r,echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
plot(f, plot.type = 'dose_fitted')

plot(f, plot.type = "dose_residuals")

plot(f, items = 12) #plot top 12 most responsive items

```

### Calculate BMD

```{r, class.source = 'fold-show'}
r <- bmdcalc(f, z = 1, x = 10)
head(r$res, 10)

```


```{r, echo=FALSE}
bmdplotwithgradient(r$res, BMDtype = 'zSD', BMD_log_transfo = FALSE, add.label = FALSE, facetby = 'typology', xmax = 50)

plot(f, plot.type = 'dose_fitted', BMDoutput = r, BMDtype = 'zSD', dose_log_transfo = FALSE)
plot(r, BMDtype = "zSD", plottype = "ecdf") + theme_bw() # plottype = "hist", "density"

```

### Bootstrap confidence intervals
```{r, class.source = 'fold-show'}
b <- bmdboot(r, niter = 1000, progressbar = TRUE, parallel = 'snow', ncpus = 6)
plot(b)

```


```{r,echo=FALSE}
subres <- bmdfilter(b$res, BMDfilter = "definedBMD")
bmdplot(subres, BMDtype = "zSD", 
        add.CI = TRUE,
        colorby = "trend", 
        point.size = 2,
        point.alpha = 0.5,
        line.size = 0.8,
        line.alpha = 0.5)

tested.doses <- unique(f$omicdata$dose)
g <- curvesplot(r$res, xmax = max(tested.doses), colorby = "trend",
                line.size = 0.8, line.alpha = 0.3) +
  geom_vline(xintercept = tested.doses, linetype = 2) + theme_bw()
print(g)
```

### Calculate Gene Set BMD

```{r, echo = FALSE}
bres <- bmdfilter(b$res, BMDfilter = "definedCI")
bres$id <- as.numeric(bres$id)

```
Include only genes with BMD upper:BMD lower ratios less than 40
```{r, class.source = 'fold-show'}
BMDresults <- bres %>% 
  mutate(BMDRatio = BMD.zSD.upper/BMD.zSD.lower) %>% 
  filter(BMDRatio <= 40)
```

Perform KEGG enrichment analysis. Then, extract genesets that are at least 5% populated and have at least 3 genes.
```{r, class.source = 'fold-show'}
universe <- as.character(o$item)

kegg <- enrichKEGG(gene = BMDresults$id, 
                   organism = 'hsa', 
                   keyType="kegg", 
                   universe = universe,
                   qvalueCutoff = 1,
                   pvalueCutoff = 1)
```

```{r,echo=FALSE}
kegg_results <- kegg@result

#Extract the gene set information
genesets <- kegg@geneSets
get_size <- function(x) {
  if (is.data.frame(x)) {
    return(nrow(x))  # Number of rows for data frames
  } else if (is.matrix(x)) {
    return(nrow(x))  # Number of rows for matrices
  } else {
    return(length(x))  # Length for other types
  }
}

geneset_size <- data.frame(
  Name = names(genesets),
  Size = sapply(genesets, get_size)
)

kegg_all <- left_join(kegg_results, geneset_size, by = c("ID" = "Name"))

kegg_all <- kegg_all %>% 
  mutate(Ratio = Count / Size) 

kegg_all2 <- kegg_all %>% 
  filter(Ratio >= 0.05) %>% 
  filter(Count >=3)

split_cols <- strsplit(kegg_all2$geneID, "/")
max_splits <- max(lengths(split_cols))

new_cols <- paste0("col", 1:max_splits)

kegg_all2[new_cols] <- t(sapply(split_cols, `length<-`, max_splits))


key <- bres[,c(1,20)]

for (col in 14:ncol(kegg_all2)) {
  matched_values <- key$BMD.zSD[match(kegg_all2[[col]], key$id)]
  kegg_all2[[col]] <- ifelse(is.na(matched_values), kegg_all2[[col]], matched_values)
}
for (i in 14:81) {
  # Convert each column to numeric, coercing problematic values to NA
  kegg_all2[[i]] <- as.numeric(as.character(kegg_all2[[i]]))
}

kegg_geneset_BMD <- kegg_all2 %>% 
  rowwise() %>%  
  mutate(
    Mean = mean(c_across(14:81), na.rm = TRUE),
    Min = min(c_across(14:81), na.rm = TRUE),
    Max = max(c_across(14:81), na.rm = TRUE),
    Median = median(c_across(14:81), na.rm = TRUE),
    SD = sd(c_across(14:81), na.rm = TRUE),
    Range_Min = min(c_across(14:81), na.rm = TRUE),
    Range_Max = max(c_across(14:81), na.rm = TRUE)
  ) %>%
  ungroup()

kegg_geneset_BMD_trim <- kegg_geneset_BMD[,c(1:13, 82:88)]
```

Calculate Mean, Min, Max, Median, Standard Deviation, Minimum Range, and Maximum Range
```{r, class.source = 'fold-show'}
print(kegg_geneset_BMD_trim)
```
Perform MSig BROAD enrichment analysis. Then, extract genesets that are at least 5% populated and have at least 3 genes.
```{r, echo = FALSE}
msig_h <- msigdbr(species = "Homo sapiens", category = "H")%>%
  dplyr::select(gs_name, entrez_gene) %>%
  dplyr::rename(ont = gs_name, gene = entrez_gene)
msig_h

BMDresults$id <- as.character(BMDresults$id)

msig_ora <- enricher(gene = BMDresults$id,
                     pvalueCutoff = 1,
                     qvalueCutoff = 1,
                     universe = universe,
                     TERM2GENE = msig_h) %>%
  as_tibble
msig_ora

msig_all <- msig_ora %>% 
  mutate(Size = as.numeric(str_extract(BgRatio, "^[^/]+"))) %>% 
  mutate(Ratio = Count / Size) 

msig_all2 <- msig_all %>% 
  filter(Ratio >= 0.05) %>% 
  filter(Count >=3)

split_cols <- strsplit(msig_all2$geneID, "/")
max_splits <- max(lengths(split_cols))

new_cols <- paste0("col", 1:max_splits)

msig_all2[new_cols] <- t(sapply(split_cols, `length<-`, max_splits))


key <- bres[,c(1,20)]

for (col in 12:ncol(msig_all2)) {
  matched_values <- key$BMD.zSD[match(msig_all2[[col]], key$id)]
  msig_all2[[col]] <- ifelse(is.na(matched_values), msig_all2[[col]], matched_values)
}

for (i in 12:53) {
  # Convert each column to numeric, coercing problematic values to NA
  msig_all2[[i]] <- as.numeric(as.character(msig_all2[[i]]))
}

msig_geneset_BMD <- msig_all2 %>% 
  rowwise() %>%  
  mutate(
    Mean = mean(c_across(12:53), na.rm = TRUE),
    Min = min(c_across(12:53), na.rm = TRUE),
    Max = max(c_across(12:53), na.rm = TRUE),
    Median = median(c_across(12:53), na.rm = TRUE),
    SD = sd(c_across(12:53), na.rm = TRUE),
    Range_Min = min(c_across(12:53), na.rm = TRUE),
    Range_Max = max(c_across(12:53), na.rm = TRUE)
  ) %>%
  ungroup()

msig_geneset_BMD_trim <- msig_geneset_BMD[,c(1:11, 54:60)]
```

Calculate Mean, Min, Max, Median, Standard Deviation, Minimum Range, and Maximum Range
```{r, class.source = 'fold-show'}
print(msig_geneset_BMD_trim)
```

Calculate the first mode for genes and pathways
```{r, echo = FALSE}
dens_gene <- density(bres$BMD.zSD)

first_derivative <- diff(dens_gene$y) / diff(dens_gene$x)
second_derivative <- diff(first_derivative) / diff(dens_gene$x[-length(dens_gene$x)])

first_derivative_full <- c(first_derivative, NA)
second_derivative_full <- c(NA, second_derivative, NA)

first_mode_index <- which(first_derivative_full[-1] < 0 & first_derivative_full[-length(first_derivative_full)] > 0 & second_derivative_full[-c(1, length(second_derivative_full))] < 0)[1]

first_mode <- dens_gene$x[first_mode_index + 1]

# Plot the density and first mode
plot(dens_gene, main = "Density Estimation with First Mode")
abline(v = first_mode, col = "red", lwd = 2, lty = 2)
text(first_mode, max(dens_gene$y), labels = paste("First Mode:", round(first_mode, 2)), pos = 4, col = "red")
```